---
title: "6. Building a data object in R"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{6. Building a data object in R}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("RCEATTLE_EVAL_VIGNETTES"), "true"),
  fig.width = 7,
  fig.height = 5
)
```

Rceattle normally reads input data from an Excel workbook via `read_data()`. The same data object can be built entirely in R as a named list, useful when:

- Simulating data for testing or MSE operating models
- Generating inputs programmatically (e.g., multi-scenario loops or sensitivity analyses)
- Integrating Rceattle into pipelines that don't use Excel

Everything passed to `fit_mod()` lives in this list, so understanding its structure also helps you modify existing data objects read from Excel. This vignette walks through each component. The result is validated by `build_data(.check = TRUE)` (the default) and fitted with `fit_mod()`, or exported back to Excel with `write_data()`.

If you would rather start from a workbook than a list, `write_template()` writes a minimal, structurally complete single-species starter workbook on the current column names that you can open, edit, and read back with `read_data()`:

```{r write-template}
write_template("my_data_template.xlsx")
dat <- read_data("my_data_template.xlsx")
```

## The quick path: `build_data()`

Most of the time you do not need to assemble the whole list by hand. `build_data()` is a constructor that takes only the blocks a model uses and default-fills the rest (via `clean_data()`), so a single-species model does not need the predation, growth, or empirical-selectivity blocks. It has three combinable entry points:

```{r}
library(Rceattle)

# 1. Copy-and-edit an existing data object (a bundled dataset, a read_data()
#    result, or a combine_data() result):
dat <- build_data(base = BS2017SS, projyr = 2060)

# 2. Read a workbook, then override a few blocks:
dat <- build_data(file = "GOA_pollock.xlsx",
                  fleet_control = fc)          # fc overrides the sheet

# 3. Assemble from named blocks (the rest are default-filled):
dat <- build_data(nspp = 1, styr = 1977, endyr = 2023,
                  fleet_control = fc, weight = wt, maturity = mat,
                  sex_ratio = sr, catch_data = catch, index_data = survey,
                  comp_data = ages)
```

`build_data()` checks each block name against the recognized schema, so a typo (`maturty`) or a case-slip (`Maturity`) is caught at construction with a suggestion (not much later inside a fit) and legacy names (`fsh_biom`, `srv_biom`, `wt`, `pmature`, `Pyrs`) are mapped to their canonical equivalents. Printing the result shows an indented specification tree rather than the raw list:

```{r}
dat            # <Rceattle data>: dimensions -> fleets -> processes -> ...
```

### Which inputs does my model need?

Rather than reading the table below by hand, ask `data_requirements()`: it reports, for a given configuration, which blocks are **Required**, **Optional** (used if supplied, otherwise default-filled), or **Ignored** (not consulted because the feature is off), the exact conditions the fit-time validation enforces.

```{r}
data_requirements(msmMode = 0)            # single-species: predation Ignored
data_requirements(msmMode = 1)            # multispecies: diet + bioenergetics Required
data_requirements(BS2017SS, msmMode = 0)  # preview an existing object as single-species
```

The conditional requirements it encodes:

| Field                                    | Required when                                  |
|------------------------------------------|------------------------------------------------|
| `comp_data`                              | (always optional)                              |
| `caal_data`                              | `any(growth_model > 0)` (estimating growth)    |
| `emp_sel`                                | any fleet has `Selectivity = "Fixed"`          |
| `NByageFixed`                            | `any(estDynamics > 0)`                         |
| `env_data`                               |  env linkages / `Ceq > 1`                      |
| `diet_data`                              | `msmMode > 0`                                  |
| `ration_data`                            | `msmMode > 0`                                  |
| Bioenergetics scalars (`Ceq`, `CA`, `CB`, `Tco`, ...) | `msmMode > 0`                     |

In default single-species mode (`msmMode = 0`) you can omit any of these by leaving the field unset.

## Building the list by hand

The rest of this vignette assembles the list field by field. This is the machinery `build_data()` wraps, useful when you are simulating data, generating inputs programmatically, or want to understand exactly what each block contains. Feel free to skip sections you don't need.

## Setup

```{r}
library(Rceattle)

# Define core dimensions up front, these drive the size of all downstream arrays.
nspp     <- 2   # number of species (or stocks)
nages    <- 10  # maximum age classes (same for all species here)
nlengths <- 20  # number of length bins (for age-length keys / CAAL data)
nyrs     <- 30  # number of hindcast years
years    <- 1:nyrs

simData <- list()
```

## 1. Species / model controls

These scalars and vectors describe the biological structure of each species and control which population model is used.

```{r}
simData$nspp    <- nspp
simData$styr    <- 1       # first year of hindcast
simData$endyr   <- nyrs    # last year of hindcast
simData$projyr  <- nyrs + 10  # last year of the projection period

# Character labels shown in plots and output tables
simData$spnames <- paste0("Species", 1:nspp)

# nsex: 1 = combined-sex model (male and female treated as one group),
#        2 = sex-structured model (separate selectivity and mortality by sex)
simData$nsex    <- rep(1, nspp)

# Calendar month of spawning, used to compute the fraction-of-year weight
# applied to spawning stock biomass (e.g., 6 = mid-year spawning)
simData$spawn_month <- rep(6, nspp)

simData$nages   <- rep(nages, nspp)   # number of age classes per species
simData$minage  <- rep(1, nspp)       # minimum age (usually 1)
simData$nlengths <- rep(nlengths, nspp)

# estDynamics controls how population abundance is estimated:
#   0 = estimate population dynamics (standard statistical catch-at-age)
#   1 = fix N-at-age exactly to values in NByageFixed (e.g., external estimates)
#   2 = scale NByageFixed by a single estimated scalar per species
simData$estDynamics <- rep(0, nspp)

# Indices linking each species to its weight and growth datasets.
# Multiple weight schedules can exist (e.g., fleet-specific); these point to
# which one is used for total biomass (pop) and spawning biomass (ssb).
simData$pop_wt_index <- 1:nspp
simData$ssb_wt_index <- 1:nspp
simData$pop_age_transition_index <- rep(1, nspp)

# Length-weight relationship coefficients: W = alpha * L ^ beta.
# Used internally to convert length-based selectivity to weight when estimating growth.
simData$alpha_wt_len <- rep(0.00001, nspp)
simData$beta_wt_len  <- rep(3, nspp)

# Prior or initial standard deviation for log-recruitment deviations. Controls the
# strength of the recruitment penalty in penalized-likelihood mode.
simData$sigma_rec <- rep(1, nspp)

# Non-modeled ("other") prey biomass available to predators. Used in
# multi-species mode to prevent unrealistically high predation mortality when
# modeled prey are scarce. Setting this too low forces predators to over-eat
# modeled prey; too high dilutes the predation signal.
simData$other_food <- rep(1e5, nspp)
```

## 2. Fleet control table

Each row defines one fleet (fishery or survey) and controls how it is modeled. Multiple fleets can target the same species.

```{r}
# Fleet_type:
#   "Fishery", contributes catch and catch-at-age/length composition data
#   "Survey", contributes an abundance index and composition data
#   "Off", fleet is present in the data but excluded from the likelihood

# Selectivity options (Selectivity column), use the string name or integer code:
#   "Fixed"              (0), fixed at values in emp_sel; not estimated
#   "Logistic"           (1), two-parameter ascending logistic (a50, slope)
#   "NonParametric"      (2), freely estimated selectivity-at-age/length with smoothness penalty
#   "DoubleLogistic"     (3), ascending + descending dome-shaped logistic
#   "DescendingLogistic" (4), descending logistic only (dome-shaped for oldest ages)
#   "Hake"               (5), non-parametric selectivity à la Taylor et al. (Pacific hake)
#   "2DAR1"              (6), 2D autoregressive random effects (bin × year)
#   "3DAR1"              (7), 3D autoregressive random effects (bin × year × cohort)
#
# Selectivity_dimension: "Age" or "Length"
#   Length-based selectivity requires an age-to-length transition matrix (see below).

# Catchability options (Catchability column), use the string name or integer code:
#   "Fixed"                (0), Q fixed at Catchability_init; not estimated
#   "Estimated"            (1), freely estimated on log scale
#   "Estimated-with-prior" (2), estimated with a lognormal prior (Catchability_init, Catchability_prior_sd)
#   "Analytical"           (3), Ludwig & Walters / Martell (1994) analytical Q
#   "PowerEquation"        (4), power-equation Q (habitat area scaling)
#   "Environmental"        (5), mu_q + X * beta, where X is a covariate in env_data
#   "AR1"                  (6), REMOVED in 5.12.0; data_check() errors. Use a q
#                                  linkage with ar1(1 | Year) and observe = <column>
#   NA, not applicable (use for fisheries without CPUE)

# Comp_distribution / CAAL_distribution options:
#   "Multinomial", standard multinomial
#   "DirichletMultinomial", overdispersed; estimates a concentration parameter
#   "MultinomialAFSC", AFSC variant of multinomial (marginal age + length)

simData$fleet_control <- data.frame(
  Fleet_name  = paste0(c("Survey", "Fishery"),
                       rep(paste(" Species", 1:nspp), each = 2)),
  Fleet_code  = 1:(nspp * 2),       # unique integer ID per fleet
  Fleet_type  = rep(c("Survey", "Fishery"), nspp),
  Species     = rep(1:nspp, each = 2),
  Month       = 0,                   # 0 = mid-year; calendar month otherwise

  # Selectivity_index links fleets that share the same selectivity curve.
  # Fleets with the same index share all selectivity parameters.
  Selectivity_index         = 1:(nspp * 2),
  Selectivity               = "Logistic",
  Selectivity_dimension     = "Age",
  N_sel_bins                = NA,    # used for NonParametric only
  Sel_curve_pen1            = NA,    # smoothness penalty (NonParametric)
  Sel_curve_pen2            = NA,
  # Time_varying_sel:
  #   0 / "Off" = time-invariant (default)
  #   1 / "IID" = penalized log-normal deviates (SD fixed at Time_varying_sel_sd
  #       unless random_sel = TRUE in fit_mod)
  #   2 / "AR1" = REMOVED in 5.16.0; data_check() errors on it
  #   3 / "Block" = time blocks (specify via R-side mapping)
  #   4 / "RandomWalk" = random walk
  #   5 / "RandomWalkAscending" = random walk on ascending portion of double logistic
  Time_varying_sel          = 0,
  Time_varying_sel_sd = 1,
  Bin_first_selected        = 1,     # youngest fully-selected age or length bin
  Sel_norm_bin             = "Off", # "Max", "Off", or a bin to normalize at
  Sel_norm_bin_upper       = "Off", # "Off", or the top of a bin range
  # Two-sex models only: whether the normalization reference is pooled over the
  # sexes ("AcrossSexes", default, keeps relative sex selectivity) or taken per
  # sex ("WithinSex", both sexes reach 1, only the shape differs).
  Sel_norm_scope           = "AcrossSexes",

  # Composition likelihoods
  Comp_distribution = "Multinomial",
  Comp_weights = 1,                  # input effective sample size multiplier
  CAAL_distribution = "Multinomial",
  CAAL_weights = 1,

  # Index units and weight/growth linkage
  Observation_units     = 1,          # 1 = weight (mt), 2 = numbers
  Weight_index         = rep(1:nspp, each = 2),
  Age_transition_index = 1,

  # Catchability_index links fleets that share the same catchability parameter.
  Catchability_index              = 1:(nspp * 2),
  Catchability         = rep(c("Estimated", NA), nspp),
  Catchability_init              = rep(c(1,   NA), nspp),
  Catchability_prior_sd           = rep(c(0.2, NA), nspp),
  # Time_varying_q:
  #   0 / "Off" = time-invariant
  #   1 / "IID" = penalized log-normal deviates (SD fixed at Time_varying_q_sd
  #       unless random_q = TRUE in fit_mod)
  #   2 / "AR1" = REMOVED in 5.16.0; data_check() errors on it
  #   3 / "Block" = time blocks
  #   4 / "RandomWalk" = random walk
  Time_varying_q          = rep(c(0, NA), nspp),
  Time_varying_q_sd = rep(c(1, NA), nspp),

  # Estimate_index_sd:
  #   0 = fix at Log_sd from index_data
  #   1 = estimate a single SD
  #   2 = analytical sigma (Ludwig & Walters 1994)
  Estimate_index_sd  = rep(c(0,  NA), nspp),
  Index_sd     = rep(c(1,  NA), nspp),

  # Estimate_catch_sd:
  #   0 = fix at Log_sd from catch_data (treats catch as known)
  #   1 = estimate
  #   2 = analytical sigma (Ludwig & Walters 1994); concentrates the sd out of
  #       the catch likelihood, so catch stops pinning F as tightly
  Estimate_catch_sd  = rep(c(NA, 0),  nspp),
  Catch_sd     = rep(c(NA, 1),  nspp),

  # Proj_F_proportion: how future F is distributed across fishery fleets per species.
  # Must sum to 1 across fishery fleets for each species. NA for surveys.
  Proj_F_proportion = rep(c(NA, 1), nspp)
)
```

## 3. Survey index data

One row per fleet-year combination for survey fleets. `Observation` is the log-scale biomass index (or numbers if `Observation_units = 2`). A negative `Year` value tells Rceattle to predict the index without including it in the likelihood, which is useful for model checking.

```{r}
simData$index_data <- data.frame(
  Fleet_name        = rep(paste("Survey Species", 1:nspp), each = nyrs),
  Fleet_code        = rep(seq(1, nspp * 2, by = 2), each = nyrs),
  Species           = rep(1:nspp, each = nyrs),
  Year              = rep(years, nspp),
  Month             = 0,
  # Selectivity_block: integer index into the selectivity/Q parameter block.
  # Used when Time_varying_sel or Time_varying_q = 3 (time blocks).
  Selectivity_block = 1,
  Observation       = rnorm(nyrs * nspp),   # replace with real log-index values
  Log_sd            = 0.1                   # observation SE on the log scale
)
```

## 4. Catch data

One row per fishery fleet-year combination. `Log_sd` is typically set small (0.01–0.05) when catch is treated as known; increase it if catch is highly uncertain.

```{r}
simData$catch_data <- data.frame(
  Fleet_name        = rep(paste("Fishery Species", 1:nspp), each = nyrs),
  Fleet_code        = rep(seq(2, nspp * 2, by = 2), each = nyrs),
  Species           = rep(1:nspp, each = nyrs),
  Year              = rep(years, nspp),
  Month             = 0,
  Selectivity_block = 1,
  Catch             = abs(rnorm(nyrs * nspp)),  # replace with real catch (mt)
  Log_sd            = 0.05
)
```

## 5. Composition data (optional)

Age or length composition data. `Age0_Length1 = 0` for age compositions, `1` for length compositions. `Sex = 0` for combined sex, `1` for females, `2` for males, `3` for joint female+male. Composition columns (`Comp_1`, `Comp_2`, ...) can be raw counts or proportions; Rceattle normalizes internally.

Skip this section entirely (don't set `simData$comp_data`) if no composition data are available, `clean_data()` will fill it with an empty data.frame.

```{r}
comp_matrix <- matrix(abs(rnorm(nyrs * nspp * nages * 2)),
                      nrow = nyrs * nspp * 2, ncol = nages)
colnames(comp_matrix) <- paste0("Comp_", 1:nages)

simData$comp_data <- cbind(
  data.frame(
    Fleet_name   = c(rep(paste("Survey Species",  1:nspp), each = nyrs),
                     rep(paste("Fishery Species", 1:nspp), each = nyrs)),
    Fleet_code   = rep(c(seq(1, nspp * 2, 2), seq(2, nspp * 2, 2)), each = nyrs),
    Species      = c(rep(1:nspp, each = nyrs), rep(1:nspp, each = nyrs)),
    Sex          = 0,
    Age0_Length1 = 0,
    Year         = rep(years, 2 * nspp),
    Month        = 0,
    Sample_size  = 200
  ),
  comp_matrix
)
```

## 6. Conditional age-at-length (CAAL) data (optional)

CAAL data provide observed age frequencies within each length bin and allow the model to jointly fit length and age compositions, and optionally estimate the growth curve internally (see `vignette("model-parameterizations")` and `?build_growth`). One row per fleet-sex-year-length-bin combination.

`caal_data` becomes **required** when growth is being estimated internally (`any(growth_model > 0)`); otherwise it can be omitted.

```{r}
caal_matrix <- matrix(abs(rnorm(nyrs * nspp * nages * 2 * nlengths)),
                      nrow = nyrs * nspp * 2 * nlengths, ncol = nages)
colnames(caal_matrix) <- paste0("CAAL_", 1:nages)

simData$caal_data <- cbind(
  data.frame(
    Fleet_name  = c(rep(paste("Survey Species",  1:nspp), each = nyrs * nlengths),
                    rep(paste("Fishery Species", 1:nspp), each = nyrs * nlengths)),
    Fleet_code  = rep(c(seq(1, nspp * 2, 2), seq(2, nspp * 2, 2)),
                      each = nyrs * nlengths),
    Species     = c(rep(1:nspp, each = nyrs * nlengths),
                    rep(1:nspp, each = nyrs * nlengths)),
    Sex         = 0,
    Year        = rep(years, 2 * nspp * nlengths),
    Length      = rep(1:nlengths, 2 * nspp * nyrs),
    Sample_size = 200
  ),
  caal_matrix
)
```

## 7. Biological inputs

### Age-to-length transition matrix

The probability that a fish of age *a* is observed in length bin *l*. Required when fitting length compositions or estimating growth internally. Each row must sum to 1. The placeholder below uses an identity matrix (age *a* maps exactly to length bin *a*); replace with empirical or model-based growth matrices.

```{r}
age_trans_list <- lapply(1:nspp, function(sp) {
  tmp <- matrix(0, nrow = nages, ncol = nlengths)
  diag(tmp[1:min(nages, nlengths), 1:min(nages, nlengths)]) <- 1
  colnames(tmp) <- paste0("Length_", 1:nlengths)
  cbind(data.frame(Age_transition_name  = paste("Species", sp),
                   Age_transition_index = sp,
                   Species              = sp,
                   Sex                  = 0,
                   Age                  = 1:nages), tmp)
})
simData$age_trans_matrix <- do.call(rbind, age_trans_list)
```

### Ageing error matrix

The probability that a fish of true age *t* is read as observed age *o*. An identity matrix (no error) is a conservative starting point. Ageing error matrices can be generated with standard tools (e.g., the AgeingError package) and substituted here.

```{r}
age_error_list <- lapply(1:nspp, function(sp) {
  tmp <- as.data.frame(diag(1, nages))
  colnames(tmp) <- paste0("Obs_age", 1:nages)
  cbind(data.frame(Species = sp, True_age = 1:nages), tmp)
})
simData$age_error <- do.call(rbind, age_error_list)
```

### Weight-at-age

Mean weight (kg) at each age. Setting `Year = 0` applies the same schedule to all years (time-invariant). Providing multiple rows with different years enables time-varying weight-at-age, which is used in empirical weight-at-age models.

```{r}
weight_matrix <- matrix((1:nages / nages)^3 * 0.01,
                        nrow = nspp, ncol = nages, byrow = TRUE)
colnames(weight_matrix) <- paste0("Age", 1:nages)

simData$weight <- cbind(
  data.frame(Wt_name  = paste("Species", 1:nspp),
             Wt_index = 1:nspp,
             Species  = 1:nspp,
             Sex      = 0,
             Year     = 0),   # Year = 0 → time-invariant
  weight_matrix
)
```

### Maturity-at-age

Proportion mature at each age, used to compute spawning stock biomass. Values should range from 0 to 1. For sex-structured models (`nsex = 2`) this represents the proportion of females that are mature.

```{r}
mat_matrix <- matrix(1, nrow = nspp, ncol = nages)
colnames(mat_matrix) <- paste0("Age", 1:nages)
simData$maturity <- cbind(data.frame(Species = 1:nspp), mat_matrix)
```

### Sex ratio at age

Proportion female at each age. Used in combined-sex models (`nsex = 1`) to weight spawning biomass contributions. In two-sex models (`nsex = 2`) this is the proportion female at recruitment.

```{r}
sex_matrix <- matrix(0.5, nrow = nspp, ncol = nages)
colnames(sex_matrix) <- paste0("Age", 1:nages)
simData$sex_ratio <- cbind(data.frame(Species = 1:nspp), sex_matrix)
```

Where species have different numbers of age bins, `maturity` and `sex_ratio` are as wide as the longest-lived species, and **each row must be filled across its own species' `nages`**. Columns past that are padding and may be `NA`.

Getting this wrong is quiet. Both tables are summed over age: spawning biomass uses `maturity` (times `sex_ratio` where a species is modelled one-sex), and spawner-per-recruit sums the same schedule. So a `maturity` gap leaves SSB `NA` for any species, and a `sex_ratio` gap does the same for a one-sex species. On a **two-sex** species `sex_ratio` reaches only spawner-per-recruit, so the model fits normally and the failure waits until a harvest control rule asks for reference points. `data_check()` names the table, the species and the missing ages.

Rows are read by position: the `Species` column is dropped before the model sees these tables, so row *i* must be species *i*.

### Fixed selectivity (optional)

Pre-specified selectivity-at-age values for fleets whose selectivity is fixed (not estimated). **Required** when any fleet has `Selectivity = "Fixed"`. Skip if all fleets estimate their own selectivity (as in this example).

When you do need it, the table looks like:

```{r}
# Example for one fleet whose selectivity is fixed at the supplied row of values
emp_sel <- data.frame(
  Fleet_name = "Survey Species 1",
  Fleet_code = 1,
  Species    = 1,
  Sex        = 0,
  Year       = 0,                       # 0 = applied to every year
  t(matrix(c(0.1, 0.5, 1, rep(1, nages - 3)), ncol = 1,
           dimnames = list(paste0("Comp_", 1:nages), NULL)))
)
simData$emp_sel <- NULL
```

### Fixed numbers-at-age (optional)

Used when `estDynamics > 0`:

- `1` = numbers-at-age fixed exactly to these values
- `2` = numbers-at-age scaled by a single estimated scalar per species

**Required** when `any(estDynamics > 0)`. Skip if you're estimating population dynamics from scratch (the default `estDynamics = 0`).

When you do need it, the table has columns `Species_name`, `Species`, `Sex`, `Year`, then `Age1 ... Age<nages>` for the fixed numbers-at-age.

### Natural mortality (M1 base)

Residual natural mortality-at-age before any predation component. In single-species mode, total mortality M = M1. In multi-species mode, M = M1 + M2 where M2 is predation mortality estimated from the bioenergetics model. When switching to multi-species mode it is important to reduce M1 accordingly (see `vignette("single-vs-multispecies")`).

```{r}
m_matrix <- matrix(0.2, nrow = nspp, ncol = nages)
colnames(m_matrix) <- paste0("Age", 1:nages)
simData$M1_base <- cbind(data.frame(Species = 1:nspp, Sex = 0), m_matrix)
```

M1 can also be estimated rather than fixed. See `?build_M1` for options including estimating a single scalar, an age-specific vector, or a temperature-dependent function.

### Environmental covariate data (optional)

An optional time series used in environment-recruitment, -M1, -catchability, or -growth relationships. Missing years are filled with the column mean. Column names after `Year` can be anything and are referenced by column order.

Skip this if no environmental covariates are used. `clean_data()` will fall back to a Year-only data.frame with no indices, and validation only complains when a feature actually needs one (env-q catchability, a consumption equation that reads a temperature, `Ceq` 1, 2 or 3, or any env linkage). With no index to read, consumption uses the constant `fT = 1` of `Ceq = 4`. That is an error under `msmMode > 0`, where consumption drives predation mortality, and a warning under single-species, where it reaches only the reported `ration`.

```{r}
simData$env_data <- data.frame(
  Year    = 1:nyrs,
  EnvData = 1       # constant = no environmental effect (placeholder)
)
```

## 8. Multi-species bioenergetics parameters (optional in single-species mode)

These parameters govern the Wisconsin bioenergetics model used to compute predation mortality (M2) in multi-species mode. In single-species mode they reach only the reported `ration` and `consumption_at_age`, never the objective, and can be omitted entirely; Rceattle fills any missing scalars with safe defaults (`Ceq = 4`, the constant form that reads no environmental index).

In multi-species mode (`msmMode > 0`) **all** of `Ceq`, `Cindex`, `Pvalue`, `fday`, `CA`, `CB`, `Qc`, `Tco`, `Tcm`, `Tcl`, `CK1`, `CK4` must be supplied as length-`nspp` vectors; `data_check()` will list any that are missing or wrong-length. Likewise `diet_data` and `ration_data` are required.

See `?build_M1` and the CEATTLE technical documentation for full details on each parameter.

```{r}
# Ceq: consumption (bioenergetics) equation form for each species:
#   1 = Exponential (Stewart et al. 1983)
#   2 = Warm-water temperature dependence (Kitchell et al. 1977)
#   3 = Cool/cold-water temperature dependence (Thornton & Lessem 1979)
#   4 = bypass bioenergetics; use ration_data directly (set CA = 1, fday = 1, CB = 0)
simData$Ceq    <- rep(4,  nspp)

simData$Cindex <- rep(1,  nspp)   # which env_data covariate column drives it, counting from 1
simData$Pvalue <- rep(1,  nspp)   # proportion of maximum consumption realised
simData$fday   <- rep(1,  nspp)   # foraging days per year

# Wisconsin model coefficients (CA, CB) and temperature dependence parameters
simData$CA  <- rep(1,  nspp)
simData$CB  <- rep(-1, nspp)
simData$Qc  <- rep(1,  nspp)
simData$Tco <- rep(1,  nspp)
simData$Tcm <- rep(1,  nspp)
simData$Tcl <- rep(1,  nspp)
simData$CK1 <- rep(1,  nspp)
simData$CK4 <- rep(1,  nspp)

# Diet composition likelihood for each predator species
# 0 = Multinomial, 1 = Dirichlet-multinomial
simData$Diet_distribution      <- rep(0, nspp)
simData$Diet_comp_weights <- rep(1, nspp)

# Ration data (optional in single-species mode): observed mean ration-at-age from
# bioenergetics studies. Required when msmMode > 0. Skip in single-species mode.
#
# When provided, columns are: Species, Sex, Year, Age1, ..., Age<nages>.

# Diet composition data: observed stomach content proportions (predator-prey pairs).
# Required when msmMode > 0; skip in single-species mode.
#
# When provided, columns are: Pred, Prey, Pred_sex, Prey_sex, Pred_age,
# Prey_age, Year, Sample_size, Stomach_proportion_by_weight.
```

## 9. Validate and fit

`plot_data()` prints a summary of the data list.

```{r}
plot_data(simData)
```

```{r}
# Key fit_mod() arguments:
#
# estimateMode: 0 = full fit (hindcast + projection)
#               1 = hindcast only
#               2 = projection only (fix parameters, run HCR)
#               3 = evaluate only (MakeADFun, no optimisation), useful for debugging
#
# msmMode:      0 = single-species, 1 = Type II MSVPA, 2 = Type III MSVPA
#
# initMode:     0 = freely estimated initial N-at-age
#               1 = equilibrium, F_init = 0, no init deviations
#               2 = equilibrium, F_init = 0, with init deviations (default)
#               3 = non-equilibrium: F_init estimated, init devs included
#               4 = non-equilibrium: F_init scales R0
#               5 = equilibrium, F_init = 0, seeded by R_init displaced by the
#                   year-1 recruitment deviation, no init deviations
#
# phase:        TRUE = use parameter phasing for improved convergence (recommended)

sim_run <- fit_mod(
  data_list    = simData,
  estimateMode = 3,    # change to 0 for a real fit
  msmMode      = 0,
  initMode     = 2,
  random_rec   = FALSE,
  fit_control = fit_control(
    phase        = TRUE,
    verbose      = 1))
```

## 10. Export back to Excel (optional)

`write_data()` exports the data list to an Excel workbook in the standard Rceattle template format, which can then be edited and re-read with `read_data()`, so you can move between R-based and spreadsheet-based workflows.

```{r}
write_data(simData, file = tempfile(fileext = ".xlsx"))
```

## 11. Persist and share a run configuration (optional)

`write_data()` stores the *data*; the model *structure* and estimation controls
live in the `fit_mod()` call (or in a `model_config()` slot). `save_config()`
writes that full run configuration, the `model_config()` structure plus the
estimation controls and `fit_control()` bundle, to a documented, git-diffable
YAML file, so two assessment runs diff cleanly and a configuration can be archived
alongside the data. `load_config()` reads it back, and passing it as
`fit_mod(config = ...)` reproduces the fit (filling only the arguments you did not
pass explicitly). The model structure is overlaid field by field: a field the
config leaves at its default keeps what the data object holds, so a config
built from `model_config()` does not drop linkages attached with
`build_data(model_config = )`, and a field both set differently is taken from
the config with a warning:

```{r}
cfg_file <- tempfile(fileext = ".yaml")
save_config(model_config(msmMode = 1, initMode = "OffsetEquilibrium"), cfg_file)
cfg <- load_config(cfg_file)
# fit_mod(simData, config = cfg)   # applies the stored structure + controls
```

Every fit also records the configuration it used in `fit$run_config`, so
`save_config(fit, "run.yaml")` archives exactly what produced a result.
