---
title: "Environmental linkages and priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Environmental linkages and priors}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval = identical(Sys.getenv("RCEATTLE_EVAL_VIGNETTES"), "true")
)
```

## Overview

Environmental drivers (temperature, prey density, climate indices)
enter stock assessments through recruitment, natural mortality,
growth, catchability, and selectivity. Rceattle exposes a
**long-format linkage table**: one row per estimated coefficient
tying an environmental term (or any user-specified design column) to
a process parameter. Describe linkages with `linkage_spec()`,
a formula-driven helper, and pass the result to the relevant
`build_*()` function. Each process then applies its environmental
offset to the linked parameter, year by year.

The linkage path is wired for **growth** (von Bertalanffy and Richards),
**natural mortality** (`M1`), **recruitment** (`R0`, `alpha`, `beta`),
**catchability** (`q`), **selectivity** (each form's own parameters), and,
prior-only, the **Dirichlet-multinomial composition weights**.

## A first example: temperature on `K`

Suppose we want sex- and species-specific von Bertalanffy growth, with
the von Bertalanffy `K` parameter responding to a temperature index by
species:

```{r}
library(Rceattle)
data(whamGrowthData)
plot_data(whamGrowthData)

# Use logistic selectivity for all fleets in this example.
whamGrowthData$fleet_control$Selectivity <- "Logistic"
whamGrowthData$fleet_control$Selectivity_dimension <- "Length"

# 1. Attach an env_data table to the data list. Must include `Year`
#    plus whatever covariates show up in your formulas. Here `temp` is
#    a placeholder; in real use it would be a standardized SST series,
#    PDO index, etc.
yrs <- whamGrowthData$styr:whamGrowthData$endyr
whamGrowthData$env_data <- data.frame(
  Year = yrs,
  temp = rnorm(length(yrs), 0, 0.1)
)

# 2. Build a growth specification with one linkage per parameter that
#    you want to vary. linkage_spec() captures (formula, by, link,
#    init, bounds, priors) for one process parameter.
growth_spec <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    K = linkage_spec(
      formula = ~ temp,             # 2 columns: (Intercept) and `temp`
      by      = ~ species + sex,    # one beta per (species, sex, column)
      priors  = list(temp = normal(0, 0.1)) # Tight prior
    )
  )
)

# 3. Fit as usual.
fit <- fit_mod(
  data_list   = whamGrowthData,
  growthFun   = growth_spec,
  estimateMode = 0, 
  msmMode     = 0,
  fit_control = fit_control(verbose = 0)
)
summary(fit)
```


## The `linkage_spec()` API

```r
linkage_spec(
  formula,
  param     = NULL,
  by        = ~ species,
  species   = NULL,                 # ids or spnames
  sex       = NULL,
  fleet     = NULL,                 # Fleet_codes or Fleet_names
  link      = "log",                # or "identity"
  init      = NULL,
  bounds    = NULL,
  priors    = NULL,
  re_group  = NA_character_,
  est_phase = 1L
)
```

Key arguments (see `?linkage_spec` for the full list):

* **`formula`**: one-sided R formula describing the linear
  predictor (anything `model.matrix()` understands: `~ 1`, `~ temp`,
  `~ temp + PDO`, `~ poly(temp, 2)`, factor predictors, etc.).
* **`param`**: target parameter name (e.g. `"K"`). Filled in
  automatically when the spec is registered under a `linkages = list(K = ...)`
  list key.
* **`by`**: stratifying factors (default `~species`); use
  `~species + sex` for per-(species, sex) coefficients, or `NULL`
  to share a single coefficient set.
* **`species`** / **`sex`** / **`fleet`**: restrict a spec to specific
  strata. Use with the multi-spec form (*Per-species* / *Per-sex formulas*
  below) to give different species, sexes, or fleets different formulas.
  `species` and `fleet` accept either 1-based ids or the model's own names
  (`data_list$spnames` / `fleet_control$Fleet_name`); `sex` accepts
  integers (`1L`, `2L`) or strings (`"Females"`/`"Males"`,
  case-insensitive). Prefer names for fleets: an id that is wrong but in
  range attaches the linkage to a different fleet and the model still fits,
  while a misspelled name errors and lists the fleets you do have.
* **`link`**: relates the linear predictor to the natural-scale
  target. `"log"` (default) gives a multiplicative effect on natural
  `param`; `"identity"` gives an additive effect on natural `param`.
* **`init`** / **`bounds`**: named lists keyed by design-matrix
  column (`init = list(temp = 0.05)`, `bounds = list(temp = c(-2, 2))`).
* **`priors`**: named list keyed by design-matrix column. Each
  entry is either an `Rceattle_prior` or a list keyed by species id;
  see *Priors* below.
* **`re_group`** / **`est_phase`**: random-effects grouping (placeholder)
  and estimation phase ordinal (`0` fixes, `1+` estimates).

## Priors

Four families are available, plus the implicit "none":

```r
prior_normal(mean, sd)
prior_lognormal(meanlog, sdlog)
prior_gamma(shape, rate)
prior_beta(shape1, shape2)
```

Inside `linkage_spec(priors = ...)` the unprefixed shorthand `normal()`,
`lognormal()`, `gamma()`, `beta()` resolves to the corresponding
`prior_*` constructors via a private data mask, this is scoped to the
`priors` argument and does not mask `base::gamma()` / `base::beta()`
elsewhere:

```{r}
linkage_spec(
  formula = ~ temp + PDO,
  by      = ~ species,
  priors  = list(
    temp = normal(0, 1),     # informative, ~95% in (-2, 2)
    PDO  = normal(0, 5)      # weak
  )
)
```

**Prior semantics.** On `(Intercept)` rows the prior is evaluated on the
*natural scale* (e.g. `normal(0.3, 0.1)` on `K` means
`dnorm(K_natural, 0.3, 0.1)`). Use `lognormal(meanlog, sdlog)` to put a normal
directly on the log scale; under the default `fit_control(bias_adjust_proc =
TRUE)` it is centred at `meanlog - sdlog^2/2`, so `exp(meanlog)` is the prior
mean (the median with `FALSE`). On covariate rows the prior applies to the
coefficient as-is, regardless of link function.

For programmatic prior assembly (e.g. species-specific priors built in
a loop) call the `prior_*` constructors directly:

```{r}
priors <- lapply(species_specific_sds, function(s) prior_normal(0, s))
names(priors) <- c("temp", "PDO", "salinity")
linkage_spec(formula = ~ temp + PDO + salinity, priors = priors)
```

Each prior's negative log-density contributes to the `Linkage-table priors`
row of the joint NLL:

```{r}
fit$quantities$jnll_comp["Linkage-table priors", ]
```

## Putting a prior on the parameter itself

Bayesian-style informative priors can be applied directly to the base growth
parameters (`K`, `L1`, `Linf`, `m`), natural mortality (`M1`), or recruitment parameters (`R0`, `alpha`, `beta`).

A prior attached to an `(Intercept)` row is **re-targeted to the base
parameter**. The same syntax is used for any other column.

```{r}
# Example: prior on K for von Bertalanffy growth
growth_spec <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    K = linkage_spec(
      formula = ~ 1,                              # intercept only, no env effects
      by      = ~ species + sex,
      init    = list(`(Intercept)` = 0.3),   # start at K = 0.3
      # Use `lognormal(log(0.3), 0.1)` to get a normal prior on log(K).
      priors  = list(`(Intercept)` = lognormal(log(0.3), 0.1))
    )
  )
)
```

The same pattern works for any linkage target, `Linf`, `L1`,
`m` (Richards), `M1`, `R0`, `alpha`, `beta`.

An `M1` intercept prior replaces `build_M1(M1_use_prior = TRUE)` for that
species, and `fit_mod()` refuses the pair unless `M2_use_prior = TRUE` (a prior
on total M).

### Growth SD endpoints

The standard deviations of length-at-age at the `L1` and `Linf` anchors
(`sd_L1`, `sd_Linf`) are exposed as linkage targets too, so the
same `init` / `bounds` / `priors` contract applies. They act on the
growth SD-at-age rather than on the mean growth parameters, and, because
the SD-at-age has no year dimension, only
intercept-bearing formulas are honored. Slope rows on these targets
raise a warning and have no effect; slope-only formulas (`~ 0 + temp`)
error.

```{r}
# Prior on the SD at L1 anchored at exp(log(5)) = 5 length units,
# with a tight normal prior on the log scale.
growth_spec_sd <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    sd_L1   = linkage_spec(
      formula = ~ 1,
      by      = ~ species + sex,
      init    = list(`(Intercept)` = 5),
      # Prior on log-scale SD; switch to `normal(5, 0.3)`
      # for a Normal on sd_L1.
      priors  = list(`(Intercept)` = lognormal(log(5), 0.3))
    ),
    sd_Linf = linkage_spec(
      formula = ~ 1,
      by      = ~ species + sex,
      init    = list(`(Intercept)` = 15)
    )
  )
)

# After fit_mod(), the init values set the SD at L1 and the SD at Linf
# (on the log scale), and the priors evaluate against those SD
# parameters via the intercept re-targeting path described above.
```

You can also key the prior by species id for species-
specific priors:

```{r}
rec_spec <- build_srr(
  srr_fun = 0,
  linkages = list(
    R0 = linkage_spec(
      formula = ~ 1,
      by      = ~ species,
      init    = list(`(Intercept)` = 1000),
      # Species-keyed priors on log-scale R0.
      priors  = list(`(Intercept)` = list(
        `1` = lognormal(log(1000), 0.2),
        `2` = lognormal(log(1500), 0.3),
        `3` = lognormal(log(800),  0.25)
      ))
    )
  )
)
```

The prior contribution appears in the `Linkage-table priors` row of the
joint negative log-likelihood:

```{r}
fit$quantities$jnll_comp["Linkage-table priors", ]
```

You can inspect the linked parameter values via the linkage table,
the estimated slope coefficients, and the base-parameter array:

```{r}
fit$data_list$linkage_table[fit$data_list$linkage_table$param == "K", ]
fit$estimated_params$beta_linkage          # (Intercept) rows are 0; slopes are estimated
fit$estimated_params$log_growth_pars       # base K / L1 / Linf / m (on log scale internally)
```

The same approach works for **`M1`** when you want a prior on
natural mortality itself:

```{r}
data("GOApollock") # Using GOA pollock as an example

# One-species model, lognormal(log(0.34), 0.01) prior on M1
# applied via the linkage intercept (equivalent to a Normal on log_M1).
# The structural M1_model is kept at "sex_age_invariant" (= 1).
M1_spec <- build_M1(
  M1_model = "sex_age_invariant",
  linkages = list(
    M1 = linkage_spec(
      formula = ~ 1,
      by      = ~ species,
      init    = list(`(Intercept)` = 0.34),
      # Prior is on the natural-scale for M1.
      # `lognormal(log(0.34), 0.01)` evaluates Normal on log(M1).
      # Use `normal(0.34, 0.01)` for a Normal directly on natural M1.
      priors  = list(`(Intercept)` = lognormal(log(0.34), 0.01))
    )
  )
)

# Fit the model - no manual mapping needed
fit <- fit_mod(
  data_list = GOApollock,
  M1Fun     = M1_spec,
  estimateMode = 0,
  fit_control = fit_control(phase = TRUE)
)
```

The linkage prior is applied verbatim, with no lognormal bias-correction term.

## Species- and (species, sex)-specific priors

Each entry under `priors = list(...)` may be:

* a single prior, applies to every species/sex row that uses
  that design column,
* a **named list of priors keyed by species id**: one prior per
  specified species, or
* a **named list of priors keyed by species id and sex**: one prior per
 specified species and sex, or
* a **two-level nested list keyed first by specified species, then by specified sex
  id**.

```{r}
# Species-only keying
linkage_spec(
  formula = ~ temp,
  by      = ~ species,
  priors  = list(
    temp = list(
      `1` = normal(0, 0.1),    # tight prior for species 1
      #`2` = normal(0, 0.5),   # No prior species 2
      `3` = normal(0, 1.0)     # weakest for species 3
      # any species not listed gets no prior
    )
  )
)
```

For sexually dimorphic stocks, the same `temp` column can take
different priors per (species, sex). At each level the value can be
either an `Rceattle_prior` (applies to all sub-strata) or a list
that drills down further:

```{r}
linkage_spec(
  formula = ~ temp,
  by      = ~ species + sex,
  priors  = list(
    temp = list(
      # Species 1: different tightness per sex.
      `1` = list(`1` = normal(0, 0.1),    # sp 1, sex 1 (e.g. females)
                 `2` = normal(0, 0.2)),   # sp 1, sex 2 (e.g. males)
      # Species 2: scalar applies to every sex.
      `2` = normal(0, 0.5)
      # Any (species, sex) combination not listed gets no prior.
    )
  )
)
```

The same syntax works for any prior family. This is the
recommended pattern when a covariate has different prior strength
across stocks (e.g. when one species has prior literature on its
temp sensitivity but another doesn't), or when within a species
the sexes are expected to respond differently. Built
programmatically, it's just:

```{r}
sd_by_species <- c(`1` = 0.1, `2` = 0.5, `3` = 1.0)
priors_per_sp <- lapply(sd_by_species, function(s) prior_normal(0, s))

linkage_spec(
  formula = ~ temp,
  by      = ~ species,
  priors  = list(temp = priors_per_sp)
)
```


## Per-species formulas

If species have qualitatively different env relationships, e.g.
species 1 responds only to temperature, species 2 also to PDO,
register *multiple specs against the same parameter* and use the
`species` argument on each to scope it to a subset of stocks:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    K = list(
      linkage_spec(
        formula = ~ temp,
        by      = ~ species,
        species = 1L,                                    # only sp 1
        priors  = list(temp = normal(0, 0.5))
      ),
      linkage_spec(
        formula = ~ temp + PDO,
        by      = ~ species,
        species = 2L,                                    # only sp 2
        priors  = list(temp = normal(0, 1),
                       PDO  = normal(0, 1))
      )
    )
  )
)
```


Each parameter key accepts both forms:

```r
linkages = list(K = single_spec)             # one formula for all sp
linkages = list(K = list(spec_a, spec_b))    # different formulas per sp
```

Inspecting the result, `fit$linkages$K` is either an
`Rceattle_linkage_spec` or a `list` of them, 
and `fit$data_list$linkage_table` shows the
materialized rows with `species` and `X_col`.

This pattern composes with species-specific *priors*, each spec
holds its own priors, and within a spec the priors can themselves
be species-keyed if you want even finer control:

```{r}
linkage_spec(
  formula = ~ temp,
  by      = ~ species + sex,
  species = c(1L, 2L),
  priors  = list(
    temp = list(
      `1` = normal(0, 0.3),
      `2` = normal(0, 0.7)
    )
  )
)
```

## Per-sex formulas

The `sex` argument on `linkage_spec()` mirrors `species`: it scopes a
spec to a subset of sex ids. Combined with the multi-spec list form,
this lets you register *separate specs per sex* against the same
parameter, typical use cases are sex-dimorphic M priors or distinct
env responses for females vs. males:

```{r}
# Female and male M1 each get their own intercept-only prior.
M1_spec <- build_M1(
  M1_model = "sex_specific",
  linkages = list(
    M1 = list(
      linkage_spec(
        formula = ~ 1,
        by      = ~ species + sex,
        sex     = "Females",                              # or sex = 1L
        init    = list(`(Intercept)` = 0.12),
        priors  = list(`(Intercept)` = lognormal(log(0.12), 0.0001))
      ),
      linkage_spec(
        formula = ~ 1,
        by      = ~ species + sex,
        sex     = "Males",                                # or sex = 2L
        init    = list(`(Intercept)` = 0.20),
        priors  = list(`(Intercept)` = lognormal(log(0.20), 0.0001))
      )
    )
  )
)
```

The `sex` filter applies only when `by` includes `sex`; otherwise `fit_mod()`
warns that it has no effect, and a filter that matches none of the model's
levels warns that the spec is dropped. If you
want different *formulas* per sex (rather than just different priors
or inits), use the multi-spec form, one spec with `sex = 1L` and
its own RHS, another with `sex = 2L` and a different RHS. 

If you only need different priors per (species, sex) for a single
shared formula, the nested-priors form (see *Species- and (species,
sex)-specific priors* above) is simpler, one spec, one set of
design columns, prior values keyed by `species` then `sex`. Reach for
the per-sex multi-spec form when you need different RHS terms,
different `init`/`bounds`, or different `est_phase` between sexes.

## Polynomial and basis-expansion formulas

The formula passed to `linkage_spec()` is fed straight to
[`stats::model.matrix()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/model.matrix.html),
so any expression `model.matrix()` understands works, polynomial
bases, splines, interactions, and arithmetic transformations.

**Polynomials.** Use `poly()` for orthogonal polynomial bases (the
default; numerically well-conditioned), or `poly(..., raw = TRUE)`
for raw monomials. Each basis order shows up as a separate design
column with its own coefficient:

```{r}
# 4th-order orthogonal polynomial in temp, by species.
linkage_spec(
  formula = ~ poly(temp, 4),
  by      = ~ species
)
# design columns produced:
#   (Intercept), poly(temp, 4)1, poly(temp, 4)2,
#   poly(temp, 4)3, poly(temp, 4)4
```

`poly()` orthogonalises the basis against the rows of `env_data`
(every row supplied, including any beyond the projection horizon)
so the columns are zero-mean over that span and the linear
slope, quadratic curvature, etc. are estimated independently,
generally the right choice if you don't have a specific scientific
reason to prefer raw monomials.

**Arbitrary arithmetic.** Use `I()` to insert a literal expression
(otherwise `temp^2` is interpreted as a formula crossing operator
and reduces to `temp`):

```{r}
# Linear + raw quadratic in temp.
linkage_spec(
  formula = ~ temp + I(temp^2),
  by      = ~ species,
  priors  = list(`I(temp^2)` = normal(0, 0.1))   # tight prior on curvature
)
```

The design column name is the deparsed call (`I(temp^2)`), so when
you key `init`, `bounds`, or `priors` against it you have to use
the same string, backticks let you write it as a name.

**Splines.** Natural cubic splines via `splines::ns()` work
out-of-the-box and are often a better choice than high-order
polynomials when you want flexibility without strong tail behavior:

```{r}
linkage_spec(
  formula = ~ splines::ns(temp, df = 4),
  by      = ~ species
)
```

**Interactions.** `temp:PDO` adds a single product column;
`temp * PDO` is shorthand for `temp + PDO + temp:PDO`:

```{r}
linkage_spec(
  formula = ~ temp * PDO,
  by      = ~ species,
  priors  = list(`temp:PDO` = normal(0, 0.5))
)
```

**Mixing forms across species.** Combined with the per-species
formula machinery, you can give one species a polynomial response
and another a linear one without duplicating columns:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    K = list(
      linkage_spec(formula = ~ poly(temp, 3),
                   by = ~ species, species = 1L),
      linkage_spec(formula = ~ temp,
                   by = ~ species, species = 2L)
    )
  )
)
```


A practical caution: same-named columns must be numerically
identical across specs that use them. With deterministic
formulas evaluated against the same `env_data` this is automatic;
it can only fail if you've built specs against different
`env_data` frames.

## How the pieces fit together

When you fit, `fit_mod()` combines the linkages from every `build_*()`
call into a single linkage table and a shared design matrix (both stored
on `data_list`), then applies per-process offsets on top of the base
parameters each year. With no linkages the offsets are zero, so a model
without linkages is unaffected. Any priors and bounds you set on
a linkage flow through the same table and add to the joint negative
log-likelihood.

## Natural mortality

`build_M1()` supports the same `linkages` argument. The valid
parameter key is `M1`; the offset enters additively on the log
scale, so a coefficient of `0.3` on
a normalized temperature index multiplies `M1` by `exp(0.3 *
temp[yr])` each year.

**Why `M1` and not `M`?** CEATTLE decomposes total natural mortality
as `M = M1 + M2`. Only `M1` (residual / non-predation) is a
parameter; `M2` (predation) is derived from predator abundance,
suitability, and ration each iteration. Environmental effects on
predation therefore propagate via upstream parameters (e.g. a temp
effect on predator `K`, on recruitment, or on bioenergetics ration
inputs).

```{r}
yrs <- GOApollock$styr:GOApollock$endyr
GOApollock$env_data <- data.frame(
  Year = yrs,
  temp = rnorm(length(yrs), 0, 0.1)
)

M1_spec <- build_M1(
  M1_model = "sex_age_invariant",            # or M1_model = 1 (back-compat)
  linkages = list(
    M1 = linkage_spec(
      formula = ~ temp,
      by      = ~ species,
      priors  = list(temp = normal(0, 0.1))
    )
  )
)

fit <- fit_mod(
  data_list   = GOApollock,
  M1Fun       = M1_spec,
  estimateMode = 0,
  msmMode     = 0
)
```

The structural M choices (`M1_model`) and the random-effects
choices (`M1_re`) accept either integer codes or readable strings
interchangeably:

```{r}
build_M1(M1_model = "sex_age_specific", M1_re = "ar1_age")   # = 3, 4
build_M1(M1_model = 3,                  M1_re = 4)            # equivalent
```

`linkages` on M is age-aware: by default a row's `age_bin` is `NA`
and the offset broadcasts across every age. To pin an env effect to
a specific age slice, set `by = ~ species + age_bin` and supply
`strata$age_bin` when calling materialize directly (`fit_mod()`
provides this automatically). For most operational uses the default
broadcast is what you want.

The `M1_indices = c(...)` column-index argument for `M1_model = 4` / `5`
also adds an environmental effect to `M1` on the log scale, and either
may be used. Prefer `linkages`; it composes naturally with priors,
bounds, and per-species formulas.

Inspecting the result:

```{r}
fit$data_list$linkage_table         # rows have process == "M"
fit$quantities$M_linkage_offset     # [nspp, max_sex, max_age, nyrs]
fit$quantities$jnll_comp["Linkage-table priors", ]
```

Growth and M linkages **share the same table**: one fit can hold
linkage rows for both processes, and the underlying coefficient
vector `beta_linkage` indexes both.

## Recruitment

`build_srr()` accepts the same `linkages` argument with three
allowed parameter keys:

* **`R0`**: acts under mean recruitment (`srr_fun = "mean"`, penalty
  form included); the right target if recruitment should respond to env
  without fitting an SRR. Under a curve fitted in the hindcast a
  single-species `R0` is derived from alpha and beta, so an `R0` linkage
  is refused. In a multispecies model `R0` sets only the initial level
  (`R_init`), so only an intercept-only linkage (`~ 1`, a prior or fixed
  value) is accepted there.
* **`alpha`** (productivity) and **`beta`** (density-dependence),
  only do work when the model has a curve: `srr_fun` (or, for the
  penalty form, `srr_pred_fun`) is `"BevertonHolt"` or `"Ricker"`.

```{r}
# Mean-recruitment with a temperature linkage on R0.
rec_spec <- build_srr(
  srr_fun  = "mean",                       # = 0
  linkages = list(
    R0 = linkage_spec(
      formula = ~ temp,
      priors  = list(temp = normal(0, 0.5))
    )
  )
)

# Beverton-Holt with env effects on alpha (and species-keyed
# priors so each stock has its own sensitivity).
rec_spec_bh <- build_srr(
  srr_fun  = "BevertonHolt",               # = 2
  linkages = list(
    alpha = linkage_spec(
      formula = ~ temp,
      priors  = list(
        temp = list(`1` = normal(0, 0.3),
                    `2` = normal(0, 0.7))
      )
    )
  )
)

fit <- fit_mod(
  data_list   = data_list,
  recFun      = rec_spec,
  estimateMode = 0,
  msmMode     = 0
)
```

### Priors and fixed values on R0, alpha and beta

An intercept-only formula (`~ 1`) acts on the parameter itself, so the
same grammar gives a prior or a fixed value. This is the recommended
alternative to `srr_est_mode` / `srr_prior`:

```{r}
# Lognormal prior on alpha (mean 5, log-scale SD 0.5), species 1 only.
rec_prior <- build_srr(
  srr_fun  = "BevertonHolt",
  linkages = list(
    alpha = linkage_spec(~ 1, species = 1,
                         priors = list(`(Intercept)` = prior_lognormal(log(5), 0.5)))
  )
)

# Alpha fixed at 20: an intercept held at its `init`.
rec_fixed <- build_srr(
  srr_fun  = "BevertonHolt",
  linkages = list(
    alpha = linkage_spec(~ 1, init = list(`(Intercept)` = 20), est_phase = 0)
  )
)
```

`prior_lognormal(p1, p2)` is a normal density on the log of the
parameter, centred so `p1` is the log of its mean under the default
`fit_control(bias_adjust_proc = TRUE)`, and of its median with `FALSE`. For
a Ricker curve it gives the same objective as
`srr_est_mode = "LognormalPrior"`, which is deprecated (5.33.0); using both is
refused. The one prior a linkage
cannot express is on Beverton-Holt steepness, which needs spawning
biomass per recruit: `srr_est_mode = "LognormalPrior"` or `"BetaPrior"`
remains for it, in single-species models only.

The structural choice (`srr_fun`) accepts either integer codes or
readable strings interchangeably:

```{r}
build_srr(srr_fun = "mean")           # = 0
build_srr(srr_fun = "BevertonHolt")   # = 2
build_srr(srr_fun = "Ricker")         # = 4
```

Inspecting the result:

```{r}
fit$data_list$linkage_table             # rows have process == "recruitment"
fit$quantities$recruitment_linkage_offset  # [nspp, n_rec_params, nyrs]
                                           # n_rec_params = 3 (R0, alpha, beta)
```

Practical note: a linkage row's `age_bin` is ignored for
recruitment (recruitment is at age 0 by definition). The
`sex_ratio` data input partitions recruits across sexes downstream.

## Catchability

Survey/index catchability is indexed by **fleet**, not by
species/sex/age, so its linkages use `by = ~ fleet`. `build_catchability()`
takes the same `linkages` argument with a single parameter key, `q`:

```{r}
fit <- fit_mod(
  data_list = data_list,
  qFun      = build_catchability(
    linkages = list(
      q = linkage_spec(~ temp, by = ~ fleet)   # one coefficient per fleet
    )
  ),
  estimateMode = 0,
  msmMode      = 0
)

# One coefficient set per fleet; the offset lands here:
fit$quantities$q_linkage_offset      # [n_flt, nyrs]
```

Restrict a spec to particular fleets with the `fleet` argument of
`linkage_spec()`, and attach priors the same way as any other process. Name
the fleets rather than numbering them: the names are checked against
`fleet_control$Fleet_name` when the model is assembled, whereas a
`Fleet_code` typo that stays in range quietly links the wrong survey:

```{r}
build_catchability(linkages = list(
  q = linkage_spec(~ temp, by = ~ fleet, fleet = c("Pollock", "EIT_Pollock"),
                   priors = list(temp = normal(0, 1)))))
```

The `Catchability = "Environmental"` / `Time_varying_q` column pair does
the same thing by naming covariates by their position in a comma-separated
list of `env_data` column indices; a model set up that way still fits, with
a deprecation warning pointing at `build_catchability()`. As a linkage the
effect is written as a formula, holds priors/bounds/phase, and names the
fleet explicitly.

An intercept-bearing formula (`~ temp`) leaves the base catchability
estimable to hold the level and fixes the intercept row at 0, exactly as
the growth/M/recruitment intercepts re-target their base parameters.

### A prior on the survey catchability

Often you know something about a survey's catchability from outside the model:
an acoustic or diver survey believed to see close to the whole stock (`q ≈ 1`),
or a `q` prior borrowed from a catchability meta-analysis or a companion
assessment. Put that belief on the fleet's catchability with a `~ 1` formula
and a prior on the intercept:

```{r}
build_catchability(linkages = list(
  # Survey q for fleet 2 believed near 1, with moderate uncertainty.
  q = linkage_spec(~ 1, by = ~ fleet, fleet = 2,
                   priors = list(`(Intercept)` = lognormal(log(1), 0.2)))))
```

The prior is on `q` on its natural scale, so a `lognormal(log(1), 0.2)` says
"`q` is near 1, and I'd be mildly surprised by a factor of ~1.5 either way."
`~ 1` alone gives an estimated `q` with a prior on its level and no
environmental effect; write `~ 1 + temp` to keep the level prior *and* link
`q` to a covariate. This replaces `Catchability = "Estimated-with-prior"`: both
put a prior on the same `q`, so `fit_mod()` refuses the pair, including on a `q`
shared through `Catchability_index`.

## Selectivity

Selectivity is also fleet-indexed (`by = ~ fleet`) and, unlike the other
processes, has several parameters. `build_selectivity()` accepts these
keys, which name the underlying slots shared across the parametric forms:

| key | logistic family | DoubleNormal |
|---|---|---|
| `slp_asc`, `slp_desc` | ascending / descending slope | `sigma_asc`, `sigma_desc` |
| `inf_asc`, `inf_desc` | ascending / descending inflection | `peak`, `right_floor` |
| `apical` | a multiplier on one sex's whole curve (every estimated form) | same |

The DoubleNormal aliases (`sigma_asc`, `peak`, ...) map to the same slots,
so you name the quantity your fleet's form actually has.

**A per-sex apical height.** Fishing mortality is one `log_F` per fleet and
year shared by the sexes, so a sex difference in F can only come from
selectivity, and no form has a height parameter of its own. `apical`
multiplies one sex's whole curve by `exp(log_sel_apical)`, bin by bin, after
the form and before normalization, so it works for every estimated form. Name
the fleet and the sex that holds it; the other sex is the reference, as in
Stock Synthesis's male-offset option:

```r
build_selectivity(linkages = list(
  apical = linkage_spec(~ 1, by = ~ fleet + sex, fleet = 3, sex = "male",
                        priors = list(intercept = lognormal(0, 0.5)))))
```

The multiplier equals the ratio of the sexes' peak heights only where their
shapes peak equally, which the logistic family does on an age axis; for a
dome with sex-specific shape it is the pointwise multiplier, and the peak ratio
is read from `fit$quantities$sel_at_age`. Only the contrast between the sexes
is identified, so the fit is refused when both sexes or no sex hold it, when
no fleet is named, on a one-sex species, on a `Fixed`, AR1 or mirror fleet,
with `link = "identity"`, and where `Sel_norm_scope = "WithinSex"`
normalization would divide the offset out (use `"AcrossSexes"`, under which
the more-selected sex peaks at 1, or turn `Sel_norm_bin` off). The contrast is
informed only by joint compositions (`comp_data$Sex = 3`); with single-sex
compositions it rests on its prior. On `GOAatf`'s fishery (Logistic,
across-sex normalization) the male multiplier fits at 2.15 (log offset 0.77,
SE 0.21), which is also the peak ratio there because both logistic curves
saturate.

```{r}
fit <- fit_mod(
  data_list = data_list,
  selFun    = build_selectivity(
    linkages = list(
      inf_asc = linkage_spec(~ cold_pool, by = ~ fleet)
    )
  ),
  estimateMode = 0,
  msmMode      = 0
)
```

As with every process, `link = "log"` makes the effect multiplicative on
the natural parameter and `link = "identity"` makes it additive, so you can
link the log-space slope and the natural-space inflection independently.

**Supported forms:** the parametric selectivities, `Logistic`,
`DoubleLogistic`, `DescendingLogistic`, `DoubleNormal`, and `LogisticPM`.
A linkage on any other form errors at fit time naming the fleet, rather than
being estimated to no effect. `apical` is the exception: it multiplies the
finished curve, so every estimated form takes it. The non-parametric forms (`coff`) are
excluded on principle: they mean-center their coefficients every year, so a
per-year offset applied across all bins cancels exactly, a per-bin
covariate would be needed, which the formula grammar does not express.
`Time_varying_sel` process-error deviates are a separate mechanism and
compose additively with a selectivity linkage.

## Data weighting (Dirichlet-multinomial)

The Dirichlet-multinomial (DM) likelihood (Thorson et al. 2017) is a
self-weighting alternative to Francis (2011) or McAllister-Ianelli (1997)
tuning: instead of
iterating an external weight, it estimates each composition dataset's
**effective sample size** as part of the fit. That works well when a fleet has
many comp years, but the weight can be poorly identified when a fleet has only
a handful, the fit may push the effective sample size to an implausible
extreme. A prior on the DM weight keeps it in a believable range without
reverting to hand-tuning:

```{r}
# Use the DM likelihood on the fleets you want it to self-weight ...
data_list$fleet_control$Comp_distribution[c(1, 2)] <- "DirichletMultinomial"

# ... and (optionally) constrain its weight with a prior.
fit <- fit_mod(
  data_list = data_list,
  compFun   = build_composition(
    linkages = list(
      theta_comp = linkage_spec(
        ~ 1, by = ~ fleet, fleet = c(1, 2),
        # Gamma(2, 0.5): mean 4, mildly informative. Keeps the weight off 0
        # (comps ignored) and off very large (comps fit as if unweighted).
        priors = list(`(Intercept)` = gamma(2, 0.5)))
    )
  ),
  estimateMode = 0,
  msmMode      = 0
)
```

Three keys, one per kind of composition data:

| key | applies to | DM likelihood set on |
|---|---|---|
| `theta_comp` | age / length composition | `Comp_distribution` |
| `theta_caal` | conditional age-at-length | `CAAL_distribution` |
| `theta_diet` | predator diet composition | `Diet_distribution` |

The prior is on the weight itself (a larger weight = a larger effective sample
size = the composition data trusted more). It only makes sense where the DM is
actually estimating that weight, so a `theta_*` linkage on a fleet or predator
whose likelihood is **not** `DirichletMultinomial` errors up front rather than
sitting inert. And because the weight is a single number per fleet, not a
time series, this is a **prior only**: a covariate (`~ temp`), an `init`, or an
`est_phase` on a composition spec is rejected, start or fix the weight
through the model's usual `inits` / `map` if you need to.

## A multi-process example

The same pattern composes when more than one parameter is linked:

```{r}
growth_spec <- build_growth(
  fun = "Richards",
  linkages = list(
    K    = linkage_spec(
      formula = ~ temp,
      by      = ~ species + sex,
      priors  = list(temp = normal(0, 1))
    ),
    Linf = linkage_spec(
      formula = ~ temp + PDO,
      by      = ~ species,
      priors  = list(temp = normal(0, 0.5),
                     PDO  = normal(0, 0.5))
    ),
    m    = linkage_spec(
      formula = ~ 1,             # intercept only, no env effect
      by      = ~ species
    )
  )
)
```

At fit time these specs are pooled. The result:

* one shared design matrix `X = model.matrix(~ temp + PDO, env_data)` of
  three columns,
* 16 linkage-table rows: 4 (sp × sex) + 3 (Linf cols) × 3 (sp) + 3
  (m intercepts),
* same-named columns (`(Intercept)`, `temp`) matched across specs, so
  nothing is duplicated in `X`.

Per-process logic (vB has no `m`; only Richards does) is enforced at
build time:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(m = linkage_spec(~ temp))
)
#> Error: linkages$m is only valid with fun = 'Richards'; ...
```

Growth, M, and recruitment can all be linked in the same fit,
each `build_*()` output holds its own linkages list and
`fit_mod()` pools them against a single shared design matrix:

```{r}
fit <- fit_mod(
  data_list   = data_list,
  growthFun   = build_growth(
    fun      = "vonBertalanffy",
    linkages = list(
      K = linkage_spec(formula = ~ temp)
    )
  ),
  M1Fun       = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      M1 = linkage_spec(formula = ~ temp)
    )
  ),
  recFun      = build_srr(
    srr_fun  = "mean",
    linkages = list(
      R0 = linkage_spec(formula = ~ temp)
    )
  ),
  estimateMode = 0,
  msmMode     = 0
)

# Linkage table contains rows for all three processes:
table(fit$data_list$linkage_table$process)
#>      growth           M  recruitment
#>           4           4            4
```

## Inspecting and debugging

A few diagnostics worth knowing:

```{r}
# The materialized table, one row per estimated coefficient.
fit$data_list$linkage_table

# The shared design matrix, one column per unique RHS term.
fit$data_list$linkage_X

# Per-(species, sex, growth_param, year) offsets actually applied to
# growth_parameters this iteration. Useful for verifying the linkage
# fired as you intended.
str(fit$quantities$growth_linkage_offset)
str(fit$quantities$M_linkage_offset)
str(fit$quantities$recruitment_linkage_offset)

# Per-row coefficient values in the same order as `linkage_table`.
fit$estimated_params$beta_linkage

# The prior contribution from the "Linkage-table priors" row of jnll_comp.
fit$quantities$jnll_comp["Linkage-table priors", ]
```

## Behind the scenes: automatic base-parameter handling

For three species × two sexes × `~ temp` (intercept + slope) the
pooled table has 12 rows, an `(Intercept)` and a `temp` slope per
(species, sex). Priors and inits supplied for `(Intercept)` flow to
the *base parameter* (`K`, `M1`, `R0`, ...); the slope rows hold the
year-by-year offset that combines with the base. The exact behavior
depends on whether the formula holds an intercept:

| Formula                      | Base parameter                            | Slope rows |
|------------------------------|-------------------------------------------|------------|
| `~ 1` (intercept only)       | estimable; init from spec or default      |n/a |
| `~ temp` (intercept + slope) | estimable; init from spec or default      | estimated |
| `~ 0 + temp` (slope only)    | fixed (not estimated); value from spec or default | estimated |

**Intercept-bearing formulas (`~ 1`, `~ temp`)**: the base parameter
remains estimable; for example `R0` for
species `s` lives on `rec_pars[s, 1]` exactly as it does without any
linkages. The `(Intercept)` row of the table exists for bookkeeping
and as a hook for `init`/`prior` (see below) but its `beta_linkage`
slot is held at `0` and not estimated. The recruitment, mortality, and
growth processes use `base + offset(yr)` each year, with the
intercept fixed at zero, the offset is exactly the slope contribution.

**Slope-only formulas (`~ 0 + temp`)**: there is no `(Intercept)`
row, so the base parameter is held fixed and stays at the
default initial value, or the value supplied by the `inits` argument in `fit_mod`, or
the value supplied by `init = list("(Intercept)"= ...)` on the spec.

### `init` and the base parameter

`init = list(`(Intercept)` = X)` on the linkage spec **flows to the
base parameter** rather than to the linkage row. The linkage row's
`beta_linkage` slot stays at `0`; the base parameter starts at `X`.
This is the recommended way to set a sensible starting value for
intercept-bearing linkages, it composes naturally with phasing
(the base parameter goes through the existing per-process phase
logic) and keeps the optimizer grounded. With `est_phase = 0` the base
parameter is held at `X` even when `fit_mod(inits = )` holds another
value; for an estimated intercept `X` is only a starting value, and
`inits` win.

```{r}
# Start M1 at 0.06 without estimating an extra intercept beta.
build_M1(
  M1_model = "sex_age_invariant",
  linkages = list(
    M1 = linkage_spec(
      formula = ~ 1,
      init    = list(`(Intercept)` = 0.06)
    )
  )
)
```

### Stratification

Stratification (`by`) determines *which* base-parameter slots are
affected: `by = ~ species` produces one `(Intercept)` row per species;
`by = NULL` (a fully shared linkage) produces a single shared row;
and `by = ~ species + sex` produces one per (species, sex). Each
`(Intercept)` row that the user supplies an `init` for pushes that
init into the matching base-parameter slot.

This rule applies uniformly to every linkage target (`K`, `L1`,
`Linf`, `m`, `M1`, `R0`, `alpha`, `beta`).

### Single-sex models and `by = ~ ... + sex`

If your model is single-sex (`nsex = 1` for every species in scope),
asking for `by = ~ species + sex` produces a warning: there is only
one sex level available, so the sex stratum collapses to a single
shared level. The result is equivalent to `by = ~ species`. Models
with mixed `nsex` (some single-sex, some two-sex species) emit one
sex level per species automatically, single-sex species get one row
per parameter, two-sex species get two.

## Random effects

A linkage formula may hold a **random effect**, so a parameter varies
year-to-year (or over any grouping) as a set of deviations damped by an
estimated density rather than fixed covariate slopes:

* `~ (1 | Year)`, **IID** deviations, `N(0, sigma)`;
* `rw(1 | Year)`: a **random walk**, `N(0, sigma)` on first differences
  (first deviate fixed so the base parameter holds the level, fixed at its
  `inits` value, 0 unless you supply another);
* `ar1(1 | Year)`: a stationary **AR1**, `SCALE(AR1(rho), sigma)` with
  `sigma` the marginal SD and `rho` the correlation.

These random-effect structures are the general way to express a
time-varying parameter, and cover what the `Time_varying_q`,
`Time_varying_sel`, and `M1_re` switches do. A survey catchability that used
`Time_varying_q = "RandomWalk"` becomes

```{r}
build_catchability(linkages = list(
  q = linkage_spec(~ rw(1 | Year), by = ~ fleet)))
```

time-varying selectivity becomes a linkage on the relevant selectivity parameter
(e.g. `inf_asc = linkage_spec(~ rw(1 | Year), by = ~ fleet)`), and a time-varying
natural-mortality `M1_re` model becomes
`build_M1(linkages = list(M1 = linkage_spec(~ (1 | Year))))`. The
`Time_varying_*` / `M1_re` switches are soft-deprecated; they still fit with
their exact numerics, and `fit_mod()` emits a nudge when it sees them, while
the grammar additionally lets you put a **prior** on the deviation SD (or
estimate it freely) rather than fixing it at an input value.

Note that `sigma` means different things across structures. For `rw()` it is
the **innovation** (step) SD, the SD of each `year-to-year` change, because
a random walk is non-stationary and has no overall spread. For `ar1()` it is
the **marginal** (stationary) SD, the overall spread of the deviations
(`sigma_marginal = sigma_innovation / sqrt(1 - rho^2)`), following the
`glmmTMB` convention. The two agree at `rho = 0` (where `ar1` reduces to `IID`),
but if you switch a group from `rw()` to `ar1()` the estimated `sigma` is not
directly comparable across the two.

Each grouping level gets one deviation; each group estimates its own SD. The
SD and (for `ar1`) the correlation route through the same `init` / `priors`
arguments as every other parameter: `init = list(sigma = 0.1)` fixes the SD
(the same as fixing `Time_varying_*_sd`), `init = list(rho =
0.7)` fixes the correlation, and `priors = list(sigma = lognormal(...))` /
`priors = list(rho = normal(...))` place priors. `rw()` / `ar1()` need a
numeric grouping variable (a real elapsed-time lag).

### Integrated or penalized

By default the deviations are **integrated out** by the Laplace approximation,
they are random effects in the usual sense. Many reference assessments instead
treat time-varying deviations as **penalized fixed effects**: the deviations stay
in the objective as a plain penalty with a *fixed* SD and are never integrated.
That is what ADMB/AMAK models do, and what Rceattle's own legacy
`Time_varying_sel` / `Time_varying_q` switches do. The two are genuinely
different models, the marginal likelihood holds a log-determinant term the
penalized form has no counterpart for, so an integrated `rw()` will not
reproduce a penalized reference fit.

`integrate = FALSE` asks for the penalized treatment:

```{r}
build_selectivity(linkages = list(
  slp_asc = linkage_spec(~ rw(1 | Year), fleet = "GOA_pollock_fishery",
                         init = list(sigma = 0.05), integrate = FALSE)))
```

It is permitted **only with a fixed SD** (and, for `ar1`, a fixed correlation).
A variance is consistently estimated only by integrating the deviations out;
estimating the deviations *and* their SD jointly as fixed effects is degenerate,
since the objective improves without bound as both go to zero. This is the same
reason `goa_pk` fixes `sigmaR` when it treats recruitment deviations as
penalized. It cannot be combined with `observe = `, an observed latent state
must stay integrated for the observation to identify it.

One model may mix the two treatments: an integrated state-space catchability
alongside a penalized selectivity walk is fine. Penalized deviations are ordinary
fixed effects, so `sdreport()` gives them standard errors, which an integrated
term does not get.

An `ar1(1 | Year)` term can additionally be a **state-space covariate** (the
Rogers et al. 2024 QAR1 catchability model): with `observe = "<env column>",
obs_sd = <value>` the AR1 latent is observed as an `env_data` column with a
**fixed** measurement SD and enters the linked parameter through an estimated
effect size (reported as `beta_linkage_obs` in `fit$quantities`, with a standard
error via `ADREPORT`). The effect size and the AR1 SD are separately identified
only when `obs_sd` is informative (small relative to the latent's SD); an
uninformatively large `obs_sd` lets the observation stop pinning the latent, so
use the measured series' actual measurement error. (`obs_sd` is held fixed by
default; set `obs_sd_est = TRUE` to estimate it instead, as the reference
`Estimate_q = 6` model does, but freely estimated it collapses toward 0 on a
smooth covariate, so keep it fixed unless the covariate is informative.)

The observed covariate need **not** span the whole series: if `env_data` starts
after the model start year (e.g. a survey that began mid-series), it is
auto-extended to `styr` with `NA` for the missing years and the observation is
applied **only** where the covariate is present, the latent AR1 still runs over
all years, but no observation is fabricated for the un-surveyed ones.

The AR1 latent is **zero-mean** (no estimated level), so its mean is confounded
with the linked parameter's intercept. **Standardize the observed covariate to
mean 0** (a z-score), or `build_catchability()` warns. One-step-ahead residuals
of the covariate observation come from `osa_residuals(fit, source = "ecov")`,
residualized first against its own series as in WHAM.

`rw()` / `ar1()` couple *consecutive* deviations with a unit lag, so their time
points must be **equally spaced**: a gap (a missing year) would be silently
treated as a single step. A grid with gaps errors rather than mis-specifying
the process.

`env_data` is applied to model years **by row position**: row 1 is the model
start year, and years beyond the last row get a zero offset (so `env_data` may
stop short of the projection horizon). When `env_data` holds a `Year` column
it must be contiguous and start at `styr`; a gappy or unsorted `Year`, or one
holding `NA`, errors rather than applying a covariate/deviate to the wrong
year. Rows *before* `styr` are dropped with a warning, since they would shift
every later row. Rows *after* the model years are kept: alignment runs from the
front, so they change no offset, but `cut()`, `poly()` and `scale()` are
computed over the rows `env_data` supplies, so discarding them would move a
design matrix that was already right.

## Current limitations

* **Time-variation switches are not auto-translated.** The soft-deprecated
  `Time_varying_q` / `Time_varying_sel` / `M1_re` switches (above) are not
  rewritten into the grammar for you, rewrite the spec by hand when you want the
  grammar's features. The switches estimate their deviations as penalized fixed
  effects with a fixed SD, so a faithful rewrite pairs `rw()` with
  `integrate = FALSE` and `init = list(sigma = )` set to the switch's
  `Time_varying_*_sd`; an integrated `rw()` is a different model. Two traps when
  rewriting a selectivity switch: the legacy penalty weights the *inflection*
  deviate at **4x** the *slope* deviate (and swaps which limb gets the 4x between
  `RandomWalk` and `RandomWalkAscending`), so read the sigmas off
  `ceattle.cpp`'s penalty block rather than assuming; and an inflection linkage
  needs `link = "identity"`, because the legacy deviate is added to the
  inflection whereas the default log link multiplies it. No grammar equivalent exists yet for the environmental /
  Rogers-AR1 catchability modes, the non-parametric selectivity forms, or the
  separable `M1_re = 6` (age × year), which keep using the switch-based path
  without a nudge.
* **Only the grammar's random effects are simulated.** `sim_mod(process = )` and
  `self_test(process = )` redraw catchability and selectivity where they are
  written as a random linkage; the legacy `Time_varying_q` / `Time_varying_sel`
  deviations hold densities in the model but no draw, so they keep their
  fitted values and `sim_mod()` warns that they did. Recruitment and M are drawn
  from their own switches (`M1_re`) as well as from linkages. Observed AR1
  (Rogers QAR1) linkages are the one grammar form that is deliberately not
  redrawn: the latent state is measured by an observed covariate series, and
  redrawing the state alone would leave the two describing different histories.
  A switch-based time-varying process therefore cannot have that process redrawn
  in a self-test.
* **`sigma` means different things in the two systems.** In a linkage,
  `init = list(sigma = )` under `ar1()` is the **marginal** standard deviation.
  The `M1_re` and 2DAR1/3DAR1 selectivity switches instead take the AR(1)
  **innovation** SD, and scale it to a marginal of `sigma / sqrt(1 - rho^2)`,
  the WHAM convention. The two coincide only at `rho = 0`.
* **Age-dependent linkages on processes other than M**: the
  `age_bin` stratum is consumed by the M accumulator today (a
  row's `age_bin == NA` broadcasts across ages; specific values
  pin the offset to that age slice). The schema is reserved for
  other processes that may eventually vary by age.
* **Non-parametric selectivity**: a `coff` linkage needs a per-bin
  covariate, since a per-year scalar cancels under the annual
  mean-centring; the formula grammar does not yet express that.
* **Legacy retirement**: `srr_fun %in% c(1, 3, 5)` and `srr_indices`
  are errors from 5.32.0: express the environmental effect as a
  recruitment linkage. `M1_indices` and `M1_model %in% c(4, 5)` remain
  deprecated. See the migration table in `NEWS.md`.

## See also

* `?linkage_spec`, `?build_growth`, `?build_M1`, `?build_srr`,
  `?build_catchability`, `?build_selectivity`, `?build_composition`,
  `?prior_normal`
* `vignette("model-parameterizations", package = "Rceattle")` for the
  full set of process options.
