---
title: "Growth estimation with linkages and priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Growth estimation with linkages and priors}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
  knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = identical(Sys.getenv("RCEATTLE_EVAL_VIGNETTES"), "true")
  )
```
  
## Overview
  
This vignette shows how to estimate growth with Rceattle's built-in
von Bertalanffy and Richards growth functions, and how to use the
formula-driven `linkage_spec()` API to add covariates and priors to
growth parameters.

The examples use the bundled `whamGrowthData` dataset, which supplies
conditional age-at-length data (`caal_data`) plus the stock assessment
inputs needed to estimate growth.

## Load data and inspect the growth dataset
  
```{r}
library(Rceattle)

data(whamGrowthData)
plot_data(whamGrowthData)

# The dataset includes CAAL data for the survey.
head(whamGrowthData$caal_data)

# Use logistic selectivity for all fleets in this example.
whamGrowthData$fleet_control$Selectivity <- "Logistic"
whamGrowthData$fleet_control$Selectivity_dimension <- "Length"
```
  
## Build a growth model with linkages
  
Growth linkages are built with `build_growth()` and `linkage_spec()`. A
linkage can act on any von Bertalanffy parameter, `K`, `Linf`, or `L1`
(natural-scale names; the default `link = "log"` keeps the underlying TMB
parameter on the log scale), and can hold priors on the linked
coefficients.
  
```{r}
# Create a simple environmental covariate for the example data.
set.seed(123)
yrs <- whamGrowthData$styr:whamGrowthData$endyr
whamGrowthData$env_data <- data.frame(
  Year = yrs,
  temp = rnorm(length(yrs), 0, 0.5)
)


growth_spec <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    Linf = linkage_spec(
      formula = ~ temp,             # 2 columns: (Intercept) and `temp`
      by      = ~ species,          # one beta per (species, column)
      priors  = list(
        # Intercept prior is on natural-scale Linf (log link back-transforms).
        # Use lognormal() if you want a Normal on log(Linf) instead.
        "(Intercept)" = normal(85, 5),
        temp = normal(0, 0.1)) # Tight prior on the slope
    )
  )
)

```

### What this linkage does

- `formula = ~ temp` fits an intercept and a temperature slope.
- `by = ~ species` gives each species its own intercept and slope
(add `+ sex` to further split by sex).
- `link = "log"` (the default) keeps the underlying TMB parameter on the
log scale, so the slope rows act multiplicatively on natural `Linf`.
- `priors` attaches a Normal prior to both the intercept and the
temperature slope.

The `(Intercept)` prior is a prior on the base `Linf` value (natural
scale, because the default log link is back-transformed inside the prior
block). An intercept-bearing linkage replaces the base parameter and
holds its level.

### Time-varying growth

The same grammar holds time variation, with a random-effect term on the
right-hand side instead of (or alongside) a covariate:

```{r}
# A random walk on the von Bertalanffy k. Use ar1(1 | Year) for a stationary
# autoregression, or (1 | Year) for independent annual deviations; under ar1()
# the `sigma` init is the MARGINAL standard deviation.
build_growth(
  fun = "vonBertalanffy",
  linkages = list(K = linkage_spec(~ rw(1 | Year)))
)
```

This is the only route to time-varying growth. The `growth_re` switch that
earlier versions held in the `data_list` was never implemented, nothing
consumed it, the deviation array was mapped off in every configuration, and the
model gave it no density, so it is removed as of 5.9.0 and setting it now
raises a deprecation message. Unlike the switch, a random linkage holds a real
density, so it is estimated, reported in the linkage table, and redrawn by
`sim_mod(process = "growth")`.

## Fit the model

```{r}
vbgf_prior_fit <- fit_mod(
  data_list   = whamGrowthData,
  growthFun   = growth_spec,
  estimateMode = 0,
  msmMode     = 0,
  fit_control = fit_control(verbose = 0, phase = FALSE)
)
summary(vbgf_prior_fit)
```

## Inspect the linkage table and prior contributions

```{r}
# The linkage table is stored on the fitted model.
vbgf_prior_fit$data_list$linkage_table

# The design matrix used by the linkages.
head(vbgf_prior_fit$data_list$linkage_X)

# Prior contribution appears in the joint NLL components.
vbgf_prior_fit$quantities$jnll_comp["Linkage-table priors", ]
```

## Compare with an unlinked baseline model and Richards model

```{r}
baseline_fit <- Rceattle::fit_mod(
  data_list = whamGrowthData,
  inits = NULL,       # Initial parameters at default
  estimateMode = 0,   # Estimate
  growthFun = build_growth(fun = "vonBertalanffy"), # Von Bert
  random_rec = FALSE, # No random recruitment
  msmMode = 0,        # Single species mode
  initMode = "NonEquilibrium",       # Unfished disequilibrium
  fit_control = fit_control(phase = FALSE, verbose = 1))


richards_model <- Rceattle::fit_mod(
  data_list = whamGrowthData,
  inits = NULL, # Initial parameters from above
  estimateMode = 0,   # Estimate
  growthFun = build_growth(fun = "Richards"), # Richards
  random_rec = FALSE, # No random recruitment
  msmMode = 0,        # Single species mode
  initMode = "NonEquilibrium",       # Unfished disequilibrium
  fit_control = fit_control(phase = FALSE, verbose = 1))



plot_biomass(
  list(baseline_fit, richards_model, vbgf_prior_fit),
  species = 1,
  model_names = c("Baseline VBGF", "Richards", "VBGF with Linf linkage and priors"),
  incl_proj = TRUE
)

baseline_fit$opt$AIC
vbgf_prior_fit$opt$AIC
```

## Notes on priors through linkages

- A prior on `"(Intercept)"` in a growth linkage constrains the base
growth parameter itself.
- A prior on a covariate column such as `temp` constrains that driver's
effect on the parameter.
- The linkage API accepts any R-style formula and supports per-species
or per-sex stratification.
- The length-at-age SD endpoints `sd_L1` and `sd_Linf` are also linkage
targets, so you can set `init` / `bounds` / `priors` on them. Only
intercept formulas (e.g. `~ 1`) are honored, the SDs do not vary by year,
so slope terms have no effect. See the "Growth SD endpoints" section of
`vignette("environmental-linkages-and-priors")` for an example.
