---
title: "2. HCRs and MSEs: an introduction"
author: "Grant Adams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{2. HCRs and MSEs: an introduction}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("RCEATTLE_EVAL_VIGNETTES"), "true")
)
```

## Fitting the operating model

A management strategy evaluation (MSE) runs a management strategy in closed loop: an operating model (OM) represents the "true" stock and generates data, an estimation model (EM) is fitted to that data and applies a harvest control rule (HCR), and the resulting catch feeds back into the OM. This example uses the northern rockfish assessment as the OM.

```{r, message = FALSE, warning = FALSE,echo = TRUE}
library(Rceattle)
data("NorthernRockfish2022")
nrdata <- NorthernRockfish2022
nrdata$fleet_control$Proj_F_proportion <- 1

om <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      M1 = linkage_spec(
        formula = ~ 1,
        priors = list(
          "(Intercept)" = lognormal(log(0.06), 0.05)
        )
      )
    )
  ),
  fit_control = fit_control(phase = TRUE)
)
```

------------------------------------------------------------------------

## Fitting the estimation model (Tier 3 HCR)

The `HCR = build_hcr()` argument to `fit_mod()` sets the harvest control rule and its biological reference points (BRPs). Here we use the NPFMC Tier 3 rule: fishing at F40%, the maxFABC rate (an Amendment-56 SPR proxy for FMSY), with an F35% overfishing limit (FOFL). Below the B40% target, F is ramped down linearly, reaching zero when spawning-stock biomass falls to `Alpha = 0.05` of that target. `Plimit = 0.2` is a separate shutoff: fishing stops when the stock is assessed as overfished relative to 20% of unfished SSB. See `?build_hcr` for the available rules and reference points.

```{r, message = FALSE, echo = TRUE}
em <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0, # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      M1 = linkage_spec(
        formula = ~ 1,
        priors = list(
          "(Intercept)" = lognormal(log(0.06), 0.05)
        )
      )
    )
  ),
  HCR = build_hcr(HCR = "NPFMC", # Tier3 HCR
                  Ftarget = 0.4, # F40%
                  Flimit = 0.35, # F35%
                  Plimit = 0.2,  # No fishing when SB<SB20
                  Alpha = 0.05),
  fit_control = fit_control(verbose = 1, phase = TRUE)
)
```

------------------------------------------------------------------------

## Running the MSE

`run_mse()` runs the closed loop: it projects the OM, samples data on the schedule you specify, re-fits the EM every `assessment_period` years, and applies the EM's HCR back to the OM. `nsim` sets the number of stochastic trajectories.

### Check `endyr` before you start

The OM and the EM must share a terminal year and a projection horizon; the MSE fills catch by row position across both models, so mismatched years are refused rather than run.

Then check `endyr` against the catch series. The MSE takes over at `endyr + 1`, and every year after that is a projection year whose catch the control rule sets, so catch recorded past `endyr` is blanked to `NA` at setup and `run_mse()` warns and names the years.

That warning almost always means `endyr` has fallen behind. A workbook can hold catch, index and composition data for years the model does not fit; the likelihood scores only `Year <= endyr`, so the data are added and `endyr` is left where it was. Conditioning the assessment on those years and projecting over them are different questions, and the warning is where you choose. Move `endyr` up and refit both models to condition on them; leave it and the MSE simulates those years instead.

```{r, message = FALSE, echo = TRUE}
mse1 <- run_mse(
  om = om, em = em,
  nsim = 10,
  assessment_period = 2,
  # Fishery = fleet 1 (annual), BTS = fleet 2 (biannual)
  sampling_period = c(1, 2),
  seed = 666 # default; per-sim seed = seed + sim
)
```

### Reproducibility

`run_mse()` accepts two seed arguments, both with default `seed = 666`:

* `seed` controls the random draws *inside* each simulated trajectory
  (recruitment, observation error, composition resampling). Each
  simulation worker calls `set.seed(seed + sim)`, so simulation `1`
  uses seed `667`, simulation `2` uses `668`, and so on. Re-running
  the same `run_mse()` call with the same `seed` reproduces the
  trajectory bit-for-bit.
* `regenerate_seed` (defaults to `seed`) controls the optional
  `regenerate_past = TRUE` step that re-fits the EM to OM-simulated
  historical data. Set this separately if you want to vary
  past-data realizations while holding the projection draws fixed.

You do not need to call `set.seed()` before `run_mse()`, the `seed`
arguments fully control reproducibility. (You may still want
`set.seed()` for plotting code that draws random colours or jitter.)

Reproducibility holds *within* a package version, not across one. In 5.9.0 the
observation draws moved from R into the TMB model, which displaces the
random-number stream, so a seeded `run_mse()` or `self_test()` will not
reproduce a run from 5.8.x. 5.18.0 is the same kind of boundary: it seeds each
assessment's draw on its own year and stops shortening the operating model's
projection horizon. Both move the stream for every seeded
`run_mse(simulate_data = TRUE)`.

5.13.0 is a boundary for two kinds of run rather than all of them: one using a
scalar `cap`, which that release rewrote, and one whose control rule reads
reference points from an estimated stock-recruit curve, which it corrected. A
run with neither reproduces across 5.13.0. `simulate_data = FALSE` is unaffected
by all three.

Two further points when comparing runs. The stream *position* is
data-dependent: a covariance survey fleet redraws until every row is positive,
and a multinomial composition draw stops early once the sample is exhausted, so
changing an unrelated covariance or sample size moves everything drawn after it.
And any model with an estimated suitability now varies its stomach contents,
which it previously did not.

For management-facing runs, record `packageVersion("Rceattle")` and
the `seed` alongside the saved `.rds` so a later reviewer can rerun
the exact same trajectories from the same package tag.

------------------------------------------------------------------------

## Evaluating the MSE with plots

Plotting the OM's spawning-biomass depletion across simulations shows the range of stock trajectories the strategy produces.

```{r, echo = TRUE}
plot_depletionSSB(lapply(mse1, function(x) x$OM))
```

```{r, echo = TRUE}
plot_depletionSSB(
  c(mse1$Sim_1$EM,           # EMs
    list(mse1$Sim_1$OM,      # OM
         mse1$Sim_1$OM_no_F  # OM with no fishing after the original OM's endyr
    )),    
  line_col = c(
    paste0("grey", round(seq(80, 50, length.out = 15))),
    1, 3
  )
)
```

------------------------------------------------------------------------

## Summary statistics

`mse_summary()` reduces the trajectories to performance statistics for comparing strategies.
It returns a **list**, because the metrics are computed on three different entities: `$species`
(one row per species), `$fleet` (one row per fleet), and `$total` (a named vector across the
whole system). `$meta` records what the summary was built from, the number of simulations,
species and fleets, and the harvest control rule.

```{r, warning = FALSE, echo = TRUE}
mse_summ <- mse_summary(mse1)
names(mse_summ)

knitr::kable(mse_summ$species, digits = 2)
knitr::kable(mse_summ$fleet, digits = 2)
```

Metric columns are named syntactically, `avg_catch`, `catch_iav`, `ssb_rmse_terminal`,
`om_p_overfishing`, `em_p_overfished`, with the `om_` and `em_` prefixes separating what the
operating model did from what the estimation model believed. `p_overfishing_false_pos` and its
siblings are where the two disagree.

------------------------------------------------------------------------

# Alternative scenarios

The runs below vary one element of the strategy at a time (the reference points, the survey frequency, the assessment interval, or the environment) so their performance statistics can be compared against `mse1`.

------------------------------------------------------------------------

## Different reference points

Here the EM uses a more aggressive rule (F30% target, F10% limit, and `Plimit = 0`, i.e. no SSB cutoff) to contrast with the Tier 3 run above.

```{r, message = FALSE, echo = TRUE}
em2 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0, # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      M1 = linkage_spec(
        formula = ~ 1,
        priors = list(
          "(Intercept)" = lognormal(log(0.06), 0.05)
        )
      )
    )
  ),
  HCR = build_hcr(HCR = "NPFMC", # Tier3 HCR
                  Ftarget = 0.30,# F30%
                  Flimit = 0.10, # F10%
                  Plimit = 0,    # No limit
                  Alpha = 0.1),
  fit_control = fit_control(verbose = 1, phase = TRUE)
)

mse2 <- run_mse(
  om = om, em = em2, 
  nsim = 10, 
  assessment_period = 2,
  sampling_period = c(1, 2) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

## Triennial bottom-trawl survey

`sampling_period = c(1, 3)` samples the fishery every year but the bottom-trawl survey (BTS) only every third year, testing the cost of a sparser survey.

```{r, message = FALSE, echo = TRUE}
mse3 <- run_mse(
  om = om, em = em, 
  nsim = 10, 
  assessment_period = 2,
  sampling_period = c(1, 3) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

## 4-year assessment cycle and reduced survey coverage

Here the EM is re-fitted only every fourth year (`assessment_period = 4`) and `fut_sample = 0.5` halves future sampling effort on both data streams: composition sample sizes are halved and the index log-SD is doubled (`Log_sd * 1 / fut_sample`). A lower-cost monitoring scenario.

```{r, message = FALSE, echo = TRUE}
mse4 <- run_mse(
  om = om, em = em, 
  nsim = 10, 
  fut_sample = 0.5,
  assessment_period = 4,
  sampling_period = c(1, 2) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

## A missed assessment, and a buffer while advice is stale

The scenarios above all set a *period*: an assessment every second or fourth year, for as long as the projection runs. A skipped assessment is a different thing (one gap in an otherwise regular cycle) and `assessment_period` also takes the schedule itself, as a vector of the years an assessment is completed.

```{r, message = FALSE, echo = TRUE, eval = FALSE}
biennial <- seq(om$data_list$endyr + 2, om$data_list$projyr, by = 2)
miss_yr  <- biennial[ceiling(length(biennial) / 3)]

mse6 <- run_mse(
  om = om, em = em,
  nsim = 10,
  assessment_period = setdiff(biennial, miss_yr), # every year but miss_yr
  sampling_period = c(1, 2)
)
```

The distinction matters for the answer, not just the bookkeeping. A triennial period is a permanent policy and its cost compounds; one missed assessment is a single episode of stale advice, and how much it costs depends on the state the stock was in when the gap opened.

To pair the gap with a precautionary reduction, give `catch_mult` a `data.frame` of `Year`, `Species` and `mult`. It applies only in the pairs listed, so the reduction can be held for the years advice is stale and lifted once the next assessment lands. `Species` is the species number, and any pair the table omits is multiplied by 1.

Be careful which years those are. The assessment in year `Y` sets catch for `Y+1` onward, so a missed assessment leaves the **two years after it** on the previous assessment's advice, not the missed year itself, which the previous assessment sets in the baseline too.

```{r, message = FALSE, echo = TRUE, eval = FALSE}
schedule <- setdiff(biennial, miss_yr)

# The years the missed assessment would have set: from the year after it,
# through the next assessment that was completed.
gap_yrs <- (miss_yr + 1):min(schedule[schedule > miss_yr])

# 10% reduction, held for exactly those years.
buffer <- expand.grid(Year = gap_yrs, Species = seq_len(om$data_list$nspp))
buffer$mult <- 0.90

mse7 <- run_mse(
  om = om, em = em,
  nsim = 10,
  assessment_period = schedule,
  sampling_period = c(1, 2),
  catch_mult = buffer
)
```

Run these against a baseline that differs *only* in the schedule (same `om`, same `em`, same `seed`) and compare them as paired within-replicate differences.

Two schedules on the same `seed` share their recruitment deviations, drawn once per replicate before the first assessment. Each assessment's observation draws are seeded on that assessment's own year rather than on wherever the previous assessments left the random stream, so an assessment in year `Y` starts from the same place in every schedule that assesses `Y`, and a divergence at one assessment does not displace every assessment after it.

Common random numbers are not complete, though, and it is worth knowing where they stop. One `sim_mod()` call draws every year in the assessment interval under that assessment's seed, so a year sitting *inside* a longer interval is drawn under a different seed than the same year in a schedule that assessed it directly. Two schedules that differ in which years they assess realize different observation error in the years between, even where the stock is in the same state. The paired difference is therefore mostly, but not entirely, the schedule, size the replicate count with that in mind rather than assuming all the noise has cancelled.

Once two schedules genuinely diverge (different catch taken, different years surveyed) their draws diverge too. That is a real difference between the runs rather than noise, and no seeding scheme removes it.

Make the last assessment year reach the projection horizon. Catch is filled only up to the last assessment, so a schedule that stops short leaves the trailing years at `NA`, and `mse_summary()` summarises over the whole projection, understating average catch, catch IAV and P(Closed) with nothing in the table to say so. A biennial cycle over an odd number of projection years lands here every time: `seq(2027, 2050, by = 2)` ends at 2049 against a 2050 horizon. `run_mse()` warns when it happens.

One caveat to take into any write-up: `catch_mult` multiplies catch, not ABC. A real ABC reduction changes removals only to the extent the fishery attains ABC, and for a stock whose catch sits well below ABC it changes them hardly at all. Either scale the multiplier by recent attainment, `1 - (1 - mult) * attainment`, or report the unscaled result as an upper bound on the effect of the reduction.

------------------------------------------------------------------------

## Climate-linked MSE

To test a strategy under environmental change, link a process in the OM to a covariate. Here recruitment (via `R0`) tracks a steadily rising environmental index, so the OM's productivity shifts over time while the EM keeps assuming stationary recruitment.

### Climate-linked OM

```{r, message = FALSE, echo = TRUE}

# Generate random env data
projyrs <- nrdata$styr:nrdata$projyr
projnyrs <- length(projyrs)

# Env data
nrdata$env_data <- merge(nrdata$env_data, data.frame(Year = projyrs, EnvIndex = seq(0,1, length.out = projnyrs)))

# Fit OM
om2 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      M1 = linkage_spec(
        formula = ~ 1,
        priors = list(
          "(Intercept)" = lognormal(log(0.06), 0.05)
        )
      )
    )
  ),
  recFun = build_srr(
    srr_fun = "mean",    # Mean R with covariates
    proj_mean_rec = FALSE,
    linkages = list(
      R0 = linkage_spec(formula = ~ EnvIndex)
    )),
  fit_control = fit_control(phase = TRUE)
)

plot_biomass(om2, incl_proj = TRUE)
```

### Run MSE

```{r, message = FALSE, echo = TRUE}
mse5 <- run_mse(
  om = om2, em = em, 
  nsim = 10, 
  fut_sample = 0.5,
  assessment_period = 4,
  sampling_period = c(1, 2) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

# Overall summary

Tagging each summary with an `MSE` label and stacking them puts the strategies side by side in
one table. Stack the entity you want to compare on, here the per-species metrics, rather
than the whole list.

```{r, echo = TRUE}
stack_species <- function(x, label) {
  out <- x$species
  out$MSE <- label
  out
}

knitr::kable(
  rbind(stack_species(mse_summary(mse1), 1),
        stack_species(mse_summary(mse2), 2)) |>
    dplyr::relocate(MSE),
  digits = 2)
```
