---
title: "1. Rceattle: An Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{1. Rceattle: An Introduction}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("RCEATTLE_EVAL_VIGNETTES"), "true")
)
```

## Overview

This vignette fits a sequence of five single- and multi-species models with `Rceattle`, applied to walleye pollock, arrowtooth flounder, and Pacific cod in the Gulf of Alaska and eastern Bering Sea. The model follows Holsman et al. (2016) and Adams et al. (2022). Every function has help available via `?`. The workflow is:

1.  Install `Rceattle`

2.  Load `Rceattle` and associated excel data

3.  Fit assessment models:

-   model 1: a single-species assessment model estimated with recruitment estimated as penalized effects;

-   model 2: three single-species assessment models jointly estimated with recruitment estimated as penalized effects;

-   model 3: as model 2, but with natural mortality estimated for each species;

-   model 4: as model 3, but with recruitment estimated as random effects;

-   model 5: full multi-species assessment models with time- and age- varying natural mortality due to predation.

4.  Compare models by AIC and Mohn's rho (retrospective analysis)

5.  Review plots of input data, diagnostics, and results.

## 1. Installation

`Rceattle` and associated dependencies and can be installed as follows:

```{r, eval = FALSE}
# Not run: this installs the package the vignette is already using, so
# executing it in CI would reinstall mid-render.
# Rceattle (pulls CRAN dependencies automatically; e.g. TMB, Matrix, dplyr)
install.packages("remotes")
remotes::install_github("afsc-assessments/Rceattle")

# Optional: TMBhelper provides richer optimization diagnostics.
# Rceattle falls back to plain nlminb + sdreport if it's not installed.
# devtools::install_github("kaskr/TMB_contrib_R/TMBhelper")
```

## 2. Data

`Rceattle` needs a data object to run. Data from the 2018 Gulf of Alaska and 2017 eastern Bering Sea single- and multi-species assessments are bundled with the package:

```{r, warning=FALSE, message=FALSE}
library(Rceattle)
data("GOApollock") # Data for GOA-pollock (1-species) ?GOApollock for more information on the data 
data("BS2017SS") # Combined data for 3 species. ?BS2017SS for more information on the data 
data("BS2017MS") # Multi-species data. Note: the only difference is the residual mortality (M1_base) is lower
```

Data can be visualized using the `plot_data` function:

```{r}
plot_data(GOApollock)
```

The data files can be written to excel, modified within excel, and read back into `R` as follows:

```{r, message = FALSE}
Rceattle::write_data(data_list = BS2017SS, file = "BS2017SS.xlsx")
mydata <- Rceattle::read_data( file = "BS2017SS.xlsx")
```

A description of the inputs and file structure can be found on the first sheet (`meta_data`) and the associated help function `?BS2017SS`.

## 3. Fit models

Models are fit with the `fit_mod` function. The first model is a single-species fit with no stock–recruit relationship to Gulf of Alaska pollock. It has 8 fleets: 1 fishery, 6 surveys, and 1 fleet turned off (its data excluded from the likelihood; see `model_1$fleet_control`).

```{r, message = FALSE}
model_1 <- Rceattle::fit_mod(data_list = GOApollock,
                            inits = NULL, # Initial parameters = 0
                            estimateMode = 0, # Estimate
                            random_rec = FALSE, # No random recruitment
                            fit_control = fit_control(
                              phase = TRUE,
                              verbose = 1))
summary(model_1)
```


The second model fits three single-species models jointly, with no stock–recruit curve and natural mortality fixed at the input values (`mydata$M1_base`). It has 7 fleets: 3 fisheries (one per species) and 4 surveys (see `mydata$fleet_control`). Single-species mode is the default (`msmMode = 0`):

```{r, message = FALSE}
model_2 <- Rceattle::fit_mod(data_list = mydata,
                            inits = NULL, # Initial parameters = 0
                            estimateMode = 0, # Estimate
                            random_rec = FALSE, # No random recruitment
                            msmMode = 0, # Single-species
                            fit_control = fit_control(
                              phase = TRUE,
                              verbose = 1))
summary(model_2)
```

Alternatively, natural mortality can be estimated with the `build_M1` function; `M1_model = "sex_age_invariant"` estimates a single M per species, constant across sex and age:

```{r, message = FALSE}
model_3 <- Rceattle::fit_mod(data_list = mydata,
                              estimateMode = 0, # Estimate
                              M1Fun = build_M1(M1_model = "sex_age_invariant"), # Estimate sex-invariant M)
                              random_rec = FALSE, # Penalized likelihood
                              msmMode = 0, # Single species mode
                              fit_control = fit_control(
                                phase = TRUE,
                                verbose = 1))
summary(model_3)
```

`model_4` treats recruitment as random effects via `random_rec`. It starts from the previous model's MLEs and skips phasing to reduce run-time:

```{r, message = FALSE}
model_4 <- Rceattle::fit_mod(data_list = mydata,
                              inits = model_3$estimated_params, # Start estimation at model 3's MLEs
                              estimateMode = 0, # Estimate
                              M1Fun = build_M1(M1_model = "sex_age_invariant"),
                              random_rec = TRUE, # Random recruitment
                              msmMode = 0, # Single species mode
                              fit_control = fit_control(
                                phase = FALSE,
                                verbose = 1))
summary(model_4)
```

Setting `msmMode = 1` estimates time- and age-varying natural mortality from predation, running the model in multi-species mode. This follows the MSVPA parameterization of [Magnusson (1995)](https://link.springer.com/article/10.1007/BF00179756).

```{r, message = FALSE}
model_5 <- Rceattle::fit_mod(data_list = BS2017MS,
                            inits = model_3$estimated_params, # Initial parameters from single species estimates
                            M1Fun = build_M1(M1_model = "sex_age_invariant"),
                            estimateMode = 0, # Estimate
                            niter = 3, # 3 iterations around population and predation dynamics
                            random_rec = FALSE, # No random recruitment
                            msmMode = 1, # MSVPA based
                            suitMode = 0, # Empirical suitability
                            fit_control = fit_control(
                              verbose = 1))
summary(model_5)
```

## 4. Reading the fitted object

`fit_mod()` returns an object of class `Rceattle`. Rather than reaching into it
by hand, most of what you need is reachable through standard S3 methods:

```{r}
summary(model_1)        # reference points, likelihood, convergence
coef(model_1)           # estimated parameters
head(vcov(model_1))     # covariance matrix (NULL when getsd = FALSE)
logLik(model_1)         # log-likelihood, holding its degrees of freedom
AIC(model_1)            # so stats::AIC() works
head(residuals(model_1))
head(as.data.frame(model_1))
```

`plot()`, `simulate()` and `profile()` dispatch on the fit too, and the
diagnostics return their own classes with `print()` methods that summarize the
result, `retrospective()` gives an `Rceattle_retro`, `jitter()` an
`Rceattle_jitter`, `self_test()` an `Rceattle_selftest`, `profile()` an
`Rceattle_profile`, and `osa_residuals()` an `rceattle_osa`. A data list read by
`read_data()` is an `Rceattle_data`, which also prints and summarizes.

The numbers themselves live in two places. Derived quantities on their natural
scale are in `fit$quantities`, biomass, `ssb`, `N_at_age`, `M_at_age`,
`sel_at_age`, reference points and the likelihood components. The parameters the
optimizer actually estimated are in `fit$estimated_params`, mostly on a log
scale, under abbreviated names. `parameter_dictionary()` explains those names:

```{r}
parameter_dictionary("R_log_sd")             # log_sd of recruitment -> sigma_R
parameter_dictionary(process = "selectivity")
```

Anything beginning `log_`, or ending `_log_sd`, needs `exp()` to reach the
quantity named in the dictionary's `natural` column.

`quantity_dictionary()` does the same job for the reported quantities, what
each one means, the units it is in, its dimensions, and whether `sdreport()`
gives it a standard error:

```{r}
quantity_dictionary("ssb_depletion")
quantity_dictionary(process = "reference_points")

# Everything that holds a standard error
dict <- quantity_dictionary()
dict[dict$se, c("quantity", "units")]
```

Units follow the model throughout: numbers-at-age are **thousands of fish** and
weight-at-age is **kg**, so their product is **mt**. Two things worth reading off
the dictionary before using a quantity. `ssb` is *female* spawning-stock biomass,
and every per-recruit reference point (`SPR0`, `SPRlimit`, `SPRtarget`) is
defined only in single-species mode, on a multispecies fit the model leaves them
at zero, which is not an estimate of zero.

## 5. Compare models

Models can be compared by marginal or joint negative log-likelihood, AIC, and retrospective analysis. AIC cannot compare a penalized-likelihood fit against a marginal-likelihood fit. The likelihood components are stored in the model objects built above:

```{r}
# Evaluate 1 model
AIC(model_1)                    # AIC, via the logLik() method below
model_1$opt$objective           # Negative log-likelihood
model_1$quantities$jnll_comp    # Negative log-likelihood components

# Compare AIC across 3-species penalized likelihood models
sapply(list(model_2, model_3, model_5), AIC)

```

`report_tables()` collects the quantities an assessment reports, the fit and
its likelihood decomposition, the time series with uncertainty, the reference
points, and the fits to index and catch, into one set of tidy tables. Every
table holds a `model` column, so several fits stack for a like-for-like
comparison:

```{r}
tabs <- report_tables(list(pollock = model_1, three_species = model_2))
tabs
tabs$reference_points
```

Read the `basis` column of `reference_points` before quoting anything from it. A
CEATTLE fit leaves a *number* behind for reference points it never estimated, and
each one looks like an estimate: `Ftarget` and `Flimit` sit at 1.0/yr unless the
harvest control rule estimates them, and in multispecies mode `SB0` is a
placeholder until a no-fishing projection derives it, which would make B40% a few
hundred tonnes. Those cells are returned as `NA` with the reason in `basis`,
rather than as the number the model happens to hold.

Diagnostics are attached rather than computed: a retrospective or a jitter is
tens to hundreds of refits, so `report_tables()` never runs one on your behalf.
Pass them in once you have them, and the matching sections appear:

```{r, eval = FALSE}
report_tables(model_1, retro = model_1_retro, osa = osa_residuals(model_1))
```

`standard_output()` relabels the result into the NOAA standardized assessment
output that the `stockplotr` and `asar` packages read, so Rceattle results can be
plotted and written into a report by the same tooling used for SS3, BAM, WHAM and
FIMS. That standard describes one stock and has no species dimension, so a fit
with more than one needs `species` to pick the stock to emit:

```{r}
head(standard_output(tabs, model = "pollock"))
head(standard_output(tabs, model = "three_species", species = 1))
```

The `retrospective` function measures retrospective bias, returning a list of refit models and a data.frame of Mohn's rho values.

```{r, message = FALSE, warning = FALSE}
model_1_retro <- retrospective(model_1, peels = 7)

plot_biomass(model_1_retro$Rceattle_list, model_names = paste("Pollock; mohns =", round(model_1_retro$mohns[1,2], 3)))
```

## 6. Model plots and diagnostics

`Rceattle` includes plots for hindcast and forecast time series of biomass, SSB, recruitment, biomass consumed as prey, mortality-at-age, and depletion, with uncertainty. Time-series plots accept a list of models (see above).

```{r}
model_list <- list(model_2, model_3, model_4, model_5)
model_names <- list("single-spp", "Single-spp M", "Single-spp M - RE", "Multi-spp")

# Biomass
plot_biomass(model_list, model_names = model_names, incl_proj = TRUE, add_ci = TRUE)

```

```{r}
# Recruitment
plot_recruitment(model_list, model_names = model_names, incl_proj = TRUE, add_ci = TRUE)

# SSB
plot_ssb(model_list, model_names = model_names, incl_proj = TRUE, add_ci = TRUE)

# Biomass depletion (total biomass / B0)
plot_depletion(model_list, model_names = model_names, incl_proj = FALSE, add_ci = FALSE)

# SSB depletion (female spawning biomass / SB0), the quantity a Tier 3 HCR
# compares against B40%
plot_depletionSSB(model_list, model_names = model_names, incl_proj = FALSE, add_ci = FALSE)

```

Diagnostic plots cover selectivity, composition fits, index fits, and catch fits. Composition plots take a single model; selectivity, index and catch plots take one or more.

Each fleet is drawn on the dimension its selectivity was estimated on, age, or length bin for a fleet whose `Selectivity_dimension` is `"Length"`. With one model, colour is the year, so time-varying selectivity reads as a fan. With several, colour separates the models and the fan moves to transparency.

```{r}
# Selectivity
plot_selectivity(model_2)

# Overlay models; `line_col` and `lwd` apply as in the other plotters
plot_selectivity(model_list, model_names = model_names, lwd = 2)

# Plot composition fit
plot_comp(model_2)

# Plot index
plot_index(model_list) # - All models

# Plot catch data
plot_catch(model_list)
```

Outputs of the model can queried using the `$` sign:

-   `model_1$initial_params` - list of initial parameter values prior to estimation

-   `model_1$bounds` - list of parameter bounds

-   `model_1$map` - list of parameter map in factor form (`model_1$map$mapFactor`) or list form (`model_1$map$mapList`)

-   `model_1$estimated_params` - list of estimated parameters post estimation

-   `model_1$quantities` - list of derived quantities (SSB, biomass, selectivity, etc)

-   `model_1$data_list` - the input data

-   `model_1$run_time` - model run time

-   `model_1$obj` - the TMB makeadfun object.

-   `model_1$opt` - the fit_tmb optimization object.

-   `model_1$sdrep` - the TMB sdreport object.
