---
title: "0. Model options and functionality"
author: "Grant Adams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{0. Model options and functionality}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("RCEATTLE_EVAL_VIGNETTES"), "true")
)
```

This vignette is an overview of the parameterization choices and capabilities
available in `Rceattle`. It complements:

- vignette 1, [Rceattle: An Introduction](introduction.html), which
  walks through fitting a model end-to-end;
- vignette 8, [Model parameterizations](model-parameterizations.html), the
  equation-level reference (in progress); and
- the per-topic vignettes 3 (diagnostics), 4 (projections / reference
  points), 5 (single- vs. multi-species), 6 (building a data object in R), and
  7 (Stock Synthesis migration).

If you're building a new application, this is the menu of options you
choose between when configuring `data_list` and the arguments to
`fit_mod()`.

## 1. Model dimensions and operating mode

> Equations: see vignette 8 [§Population dynamics](model-parameterizations.html#population-dynamics).

The top-level switches that decide the observation model live on
the `data_list` and the population model and estimation switches live on the call to `fit_mod()`:

| Switch | Where set | Role |
|---|---|---|
| `nspp`, `nages`, `nsex`, `nlengths` | `data_list` | Population structure |
| `styr`, `endyr`, `projyr` | `data_list` | Time horizon (hindcast + projection) |
| `msmMode` | `fit_mod()` | Single- vs. multi-species |
| `initMode` | `fit_mod()` | Initial age structure |
| `estimateMode` | `fit_mod()` | Estimate / project / debug |
| `random_rec`, `random_q`, `random_sel` | `fit_mod()` | Random-effects toggles |

### `msmMode`, single-species vs. multi-species

| Value | Meaning | Status |
|:---:|---|---|
| `0` | Single-species (M = M1 only) | Validated |
| `1` | Multi-species, MSVPA Type II predation | Validated |
| `2` | Multi-species, MSVPA Type III predation | Validated |
| `3:9` | Kinzey & Punt (2009) functional responses (Holling I/II/III, predator interference, predator preemption, Hassell-Varley, Ecosim) | **Blocked**: `data_check()` will refuse these until they are validated against the current parameter set |

### `initMode`, initial age structure

Codes (also accept the string aliases listed):

| Value | Alias | Description |
|:---:|---|---|
| `0` | `"FreeParams"` | Initial N-at-age estimated as free parameters |
| `1` | `"Equilibrium"` | Unfished equilibrium, no init devs, F<sub>init</sub> = 0 |
| `2` (default) | `"NonEquilibrium"` | Unfished equilibrium with initial deviations |
| `3` | `"FishedNonEquilibrium"` | Fished non-equilibrium with init devs; F<sub>init</sub> is added to M inside geometric series |
| `4` | `"FishedNonEquilibriumScaled"` | Fished non-equilibrium with init devs and R<sub>0</sub>-scaled by F<sub>init</sub>|
| `5` | `"OffsetEquilibrium"` | Unfished equilibrium, F<sub>init</sub> = 0, displaced by the year-1 recruitment deviation instead of sitting at R<sub>init</sub>; init devs off, no init-dev penalty (Cole Monnahan / AFSC GOA pollock convention) |
| `6` | `"FishedNonEquilibriumSelected"` | As `3`, but F<sub>init</sub> is weighted by the fishery's selectivity at age before it accumulates, so the first year decays with &sum;(M1 + F<sub>init</sub> s<sub>a</sub>). Stock Synthesis's InitF convention; F<sub>init</sub> is the apical initial F. Use this where the fishery is size-selective, since `3` charges unselected young ages the full F<sub>init</sub> and `4` applies it once |

Rceattle estimates one F<sub>init</sub> per species where SS3 carries one per fleet,
so under mode `6` a species with more than one fishery decays its initial age
structure at the **mean** selectivity of those fisheries. F<sub>init</sub> is then not
the apical initial F of any one of them, and `data_check()` warns — the initial state
sets the SSB scale.

#### The initial equilibrium catch

Modes `3`, `4` and `6` estimate F<sub>init</sub>, but nothing identifies it unless
you give it an observation: the catch the stock yielded under that initial fishing
mortality, in the year before the hindcast. Write it as an ordinary `catch_data`
row at `styr - 1`, with the fleet's `Catch` and `Log_sd`. No new columns.

**It is read only under mode `6`.** The prediction is Baranov at
F<sub>init</sub> &times; selectivity, and mode `6` is the only mode that builds the
initial age structure with that same mortality, so scoring it under `3` or `4`
would fit F<sub>init</sub> to a catch the population was never subject to. Under
any other mode the row is dropped, and `data_check()` says so.

That year also holds ordinary catch history — `GOA2018SS` carries 23 rows before
`styr`, two of them on `styr - 1` — so `data_check()` names the fleets whose rows
it reads whenever it reads any. It refuses two rows for one fleet, a non-positive
catch or `Log_sd`, a fleet that is not an active fishery, and a species with more
than one fishery (Rceattle carries one F<sub>init</sub> per species, so there is no
per-fleet split to predict with).

`clean_data()` moves the row into `data_list$equil_catch_data` so `catch_data`
keeps one row per fitted catch; `write_data()` writes it back into the catch
sheet. The prediction is reported as `quantities$equil_catch_hat` and scored on
its own `jnll_comp` row, `"Initial equilibrium catch"`.

### `estimateMode`, what does `fit_mod()` actually do?

| Value | Behavior |
|:---:|---|
| `0` | Estimate hindcast and projection parameters via `nlminb` / `TMB::sdreport` (default). |
| `1` | Estimate only the hindcast parameters via `nlminb` / `TMB::sdreport`. |
| `2` | Hold the hindcast fixed and run only the projection under the supplied `HCR`. |
| `3` | Evaluate the report once at `inits` (debug / mapping check). |
| `4` | Run through `MakeADFun` and `nlminb` with all parameters mapped out (debug / mapping check). |

### Random-effects toggles

`random_rec`, `random_q`, and `random_sel` move log-recruitment, the
log-catchability deviations, and the time-varying selectivity
deviations, respectively, from penalized fixed effects to random
effects. The deviations are marginalized via the Laplace approximation,
and their distribution parameters are estimated.

## 2. Predation: `suitMode` (per predator)

> Equations: see vignette 8 [§Predation sub-model](model-parameterizations.html#predation-sub-model) (suitability, predation mortality, consumption / bioenergetics).

`suitMode` chooses how prey preference is modeled. Length 1 (same for
all predators) or length `nspp`.

| Value | String | Description | Available? |
|:---:|---|---|:---:|
| `0` | `"Empirical"` | Empirical proportions from observed stomach data (MSVPA-style; Holsman et al. 2015) | Yes |
| `1` | `"GammaLength"` | Gamma suitability on prey length-at-age | **Not yet** |
| `2` | `"GammaWeight"` | Gamma suitability on prey weight-at-age | Yes |
| `3` | `"LognormalLength"` | Log-normal suitability on prey length-at-age | **Not yet** |
| `4` | `"LognormalWeight"` | Log-normal suitability on prey weight-at-age | Yes |
| `5` | `"NormalLength"` | Normal suitability on prey length-at-age | **Not yet** |
| `6` | `"NormalWeight"` | Normal suitability on prey weight-at-age | Yes |

Suitability is averaged across years `suit_styr:suit_endyr` (defaults to
the full hindcast) if year-specific diet data are not included (e.g. `Year = 0` in `diet_data`). 
Per-predator `suitMode` allows mixing parametric and  empirical suitability across species.

## 3. Selectivity (per fleet)

> Equations: see vignette 8 [§Selectivity parameterizations](model-parameterizations.html#selectivity-parameterizations).

Set the `Selectivity` column of `fleet_control` to one of the values
below (integers or strings both work):

| Value | String | Notes |
|:---:|---|---|
| `0` | `"Fixed"` | Selectivity supplied via the `emp_sel` data sheet |
| `1` | `"Logistic"` | Two-parameter ascending logistic |
| `2` | `"NonParametric"` | Age- or length-bin-indexed; penalties via `Sel_curve_pen1` / `Sel_curve_pen2` |
| `3` | `"DoubleLogistic"` | Dome-shaped (ascending + descending logistic) |
| `4` | `"DescendingLogistic"` | Descending only |
| `5` | `"Hake"` | Non-parametric à la Taylor et al. (Pacific hake) |
| `6` | `"2DAR1"` | 2D AR(1) bin × year (Cheng et al. 2024) |
| `7` | `"3DAR1"` | 3D AR(1) bin × year × cohort |
| `8` | `"DoubleNormal"` | Four-parameter double normal: peak, ascending and descending widths, and a right-tail floor (the Stock Synthesis pattern 24 shape without its initial-selectivity and plateau-width parameters) |
| `9` | `"NonParametricPM"` | Ianelli non-parametric with the ADMB AMAK ("pm") penalty; see `Sel_avgsel_pen`, `Sel_cap_bin` |
| `11` | `"LogisticPM"` | AMAK ("pm") logistic: multiplicative inflection/slope deviations plus a free age-1 log-selectivity |
| `13` | `"NonParametricIntegrable"` | The `NonParametric` base curve whose deviations have a proper density: the shape penalties are charged once on the base, so `random_sel = TRUE` integrates them. `Time_varying_sel` picks the structure (`Off`, `IID` or `RandomWalk`) |
| `15` | `"DoubleNormalSS3"` | Stock Synthesis size pattern 24: the six-parameter double normal, each parameter on SS3's own scale (peak; logit top width; log ascending and descending widths; logit initial and final selectivity) in its own array `sel_dn6`. An end left at SS3's -999 is unscaled and fixed. Not normalized, as in SS3. All six take selectivity linkages under the SS3 manual's names; `Time_varying_sel` must be `"Off"` |

There is no `10`: the integer codes are frozen, because renumbering one would
silently reinterpret every saved config and every fitted `.rds` that holds it.

The AR(1) forms reuse the `Sel_curve_pen` columns as logit-scale correlations,
one per dimension: `Sel_curve_pen1` across selectivity **bins** (ages or length
bins, per `Selectivity_dimension`), `Sel_curve_pen2` across **years**, and
`Sel_curve_pen3` across **cohorts** for `3DAR1`. Before 5.9.0 the 2DAR1 density
applied these two the wrong way round, so a fit that started from unequal
`Sel_curve_pen1` / `Sel_curve_pen2` (or that warm-starts from `inits` produced
by an earlier version) begins from a mirrored point. Refit from `inits = NULL`,
or transpose the two, when moving a 2DAR1 model across that boundary. Fits that
start both correlations at 0 are unaffected: `TMBphase()` holds them there, and
only the two reported estimates exchange names.

Time-varying behavior is set by `Time_varying_sel` (per fleet):

| Value | String | Behavior |
|:---:|---|---|
| `0` | `"Off"` | Time-invariant |
| `1` | `"IID"` | Independent annual deviations |
| `2` | `"AR1"` | **Removed in 5.16.0**: `data_check()` errors on it. It was never an AR1: the model scored it with the same independent normal penalty as `"IID"`, and the selectivity deviations hold no correlation parameter. Set `"IID"` for the fit it actually gave, or see the replacement below. |
| `3` | `"Block"` | Time block |
| `4` | `"RandomWalk"` | Random walk |
| `5` | `"RandomWalkAscending"` | Random walk on ascending portion of double logistic |

Not every mode applies to every selectivity form, because the model only scores
a deviation where it defines a density for it:

| Selectivity | Modes accepted |
|---|---|
| `Logistic`, `DoubleLogistic`, `DescendingLogistic`, `DoubleNormal` | `Off`, `IID`, `Block`, `RandomWalk`; `RandomWalkAscending` on `DoubleLogistic` and `DoubleNormal` |
| `NonParametric` | `Off`, `IID`, `RandomWalk` |
| `NonParametricPM` | `Off`, `RandomWalk`, its deviates *are* the walk increments, so `IID` would describe a different curve than the model builds |
| `NonParametricIntegrable` | `Off`, `IID`, `RandomWalk`: the base curve holds the shape penalties and the deviations have a proper density, so `random_sel = TRUE` integrates them. `Off` estimates no deviations |
| `Hake` | `Off`, `IID` |
| `2DAR1`, `3DAR1` | none; the field has its own correlations and sd through `Sel_curve_pen` |

The two non-parametric modes differ in what they are a deviation *of*. `"IID"`
scores each annual coefficient deviation about an estimated base curve;
`"RandomWalk"` scores the year-to-year change in the *realized* selectivity,
which is renormalized to mean 1 within each year.

**Neither is integrated out, and `fit_mod()` refuses `random_sel = TRUE` on
both**, fit them with `random_sel = FALSE`, the penalized formulation the AMAK
model intends. The walk's density says nothing about the level of a year's
coefficients, so those directions are improper and the deviation standard
deviation collapses to zero. The IID density is proper, but the AMAK shape
penalties (`Sel_curve_pen1`, `Sel_curve_pen2`) and the always-on average-selectivity
term are charged on each year's realized curve, so the prior actually integrated
is the IID normal *times* those penalties. None of them scales with
`Time_varying_sel_sd`, so the estimated sd is biased low whatever the penalty
weights are (measured at about 5% of the precision at sd 0.35 from the
average-selectivity term alone, more as the sd grows) and `Sel_curve_pen1` is
also one-sided, which leaves the Laplace objective only piecewise smooth.
Charging the shape penalties once on the base curve makes the mode
integrable without moving any penalized fit: that is what
`NonParametricIntegrable` (5.40.0) does, and it is the form to use with
`random_sel = TRUE`.
Their density is complete, but the Laplace marginal likelihood that
`random_sel = TRUE` maximizes still underestimates a variance component on small
multinomial samples: on `Atka2022`'s fishery (input sample sizes 2 to 236) a
true deviation SD of 0.35 comes back at 0.24 to 0.26, 0.70 near 0.61, and
0.35 near 0.32 with ten times the sample size (`tools/verify/verify-sim-recovery-np-integrable.R`).
Read the reported SD as a lower bound unless the compositions are well sampled.

The switch is **soft-deprecated as a whole**, and `data_check()` says so on use.
The linkage grammar expresses the same IID, random-walk and AR(1) deviations,
`(1 | Year)`, `rw(1 | Year)`, `ar1(1 | Year)`, per selectivity *parameter*
rather than per fleet, and additionally lets the deviation standard deviation be
estimated, given a prior, or held fixed:

```r
build_selectivity(linkages = list(
  inf_asc = linkage_spec(~ ar1(1 | Year), by = ~ fleet)))
```

Name whichever parameter should vary: `slp_asc`, `inf_asc`, `slp_desc`,
`inf_desc`, or a `DoubleNormal` alias. The GOA pollock 2025 assessment runs
that way. `apical` is the one linkage parameter every estimated form has: the
height of one sex's curve relative to the other's, see
[Environmental linkages and priors](environmental-linkages-and-priors.html#selectivity).

Bin indexing: with non-parametric and AR(1) forms, selectivity is
indexed over `N_sel_bins`.

`Sel_norm_bin` says where selectivity is normalized to 1. Set it to `"Max"` to
normalize by the largest value (non-differentiable), to `"Off"` to leave the
curve unnormalized, or to a bin to normalize there. Give `Sel_norm_bin_upper` as
well and the reference is mean selectivity over that range instead.

A numeric `Sel_norm_bin` is read on the fleet's own `Selectivity_dimension`: an
absolute **age** for an age-based fleet (`6` means age 6, not the sixth bin) or
a 1-based **length-bin ordinal** for a length-based one. Note `Bin_first_selected`
beside it uses the opposite convention and is always a bin ordinal. The value
must fall in the range the fleet is selected over: below `Bin_first_selected`
the curve is zeroed, so a reference taken there would divide by nothing, and a
value below it is read as `"Max"`, which is what a negative has always meant.
Legacy numbers still work: blank is `"Off"`, and `0`, `-1` or `-999` are `"Max"`.

### Sharing a selectivity between fleets

Fleets given the same `Selectivity_index` share one selectivity parameter block,
and the penalties and priors on it are accumulated once, on the group's first
non-`Off` fleet.

Sharing the parameters is not sufficient on its own. The columns that shape the
curve are read **per fleet**, so they have to agree across the group or the
fleets end up with different selectivities:

| Column | If it differs within the group |
|---|---|
| `Selectivity`, `Selectivity_dimension`, `Bin_first_selected`, `N_sel_bins`, `Sel_norm_bin`, `Sel_norm_bin_upper`, `Sel_norm_scope`, `Sel_cap_bin` | The fleets get **different curves**: mirroring fails |
| `Time_varying_sel` | Resolved to the lead fleet's setting |
| `Sel_start_year` | Resolved to the group's earliest year |
| `Time_varying_sel_sd` | Honoured per fleet while it is **fixed**. Under `random_sel = TRUE` the group estimates one deviation sd, and TMB averages the initial values of parameters mapped together, this one on the log scale, so the group starts at the *geometric* mean of its estimated members' values and no fleet keeps its own. `build_map()` warns |

A blank counts as a value in the first row: an unset `Sel_norm_bin` (written
back as `"Off"`) means "do not
normalize", which is a different curve from normalizing at a bin. `data_check()`
warns in all three cases, distinguishing divergence from resolution.

The reliable way to mirror a fleet is to copy its `fleet_control` row and change
only the identity and catchability columns:

```r
mirror <- fleet_control[fleet_control$Fleet_name == "Fishery", ]
mirror$Fleet_name         <- "Fishery_CPUE"
mirror$Fleet_code         <- nrow(fleet_control) + 1L
mirror$Fleet_type         <- "Survey"
mirror$Catchability_index <- mirror$Fleet_code
fleet_control <- rbind(fleet_control, mirror)
```

### Sex structure and relative selectivity

Populations are modelled as one or two sexes via `nsex`. **Every selectivity
form is sex-specific by default** when a species has two sexes: each sex gets
its own free copy of every selectivity parameter, for all forms.

What differs between configurations is the *relative* scale between the sexes,
whether males and females can be selected at genuinely different levels, or only
with different shapes. Normalization makes two independent decisions, held by
two columns:

* **`Sel_norm_bin`, where the reference is taken.** `"Off"` normalizes
  nothing; a bin takes it there (or, with `Sel_norm_bin_upper`, the mean over
  that range); `"Max"` takes it at the curve's maximum over bins.
* **`Sel_norm_scope`, whose scale it sets.** `"AcrossSexes"` (default) pools one
  reference over the sexes, so the less-selected sex stays below 1 and
  **relative sex-specific selectivity is retained**. `"WithinSex"` divides each
  sex by its own reference, so both reach 1 and only the *shape* differs by sex.

| `Sel_norm_bin` | `Sel_norm_scope` | Result |
|---|---|---|
| `"Off"` |, | nothing normalized; the relative scale is free |
| `"Max"` | `"AcrossSexes"` | maximum over bins **and sexes**: one sex peaks at 1, the other keeps whatever relative level its form gives it |
| `"Max"` | `"WithinSex"` | each sex scaled at **its own plateau**, wherever that falls |
| a bin | `"AcrossSexes"` | anchored at that bin, sexes kept comparable |
| a bin | `"WithinSex"` | both sexes forced to 1 at that bin; only the shape differs |

For strongly dimorphic stocks the relative sex selectivity is usually the
quantity of interest, so the `"AcrossSexes"` default is the one to keep. If
instead the sexes plateau at different ages and you want each scaled at its own
plateau, use `Sel_norm_bin = "Max"` with `Sel_norm_scope = "WithinSex"`: that is
more robust than naming a bin, since a named bin stops being the plateau once
selectivity is time-varying while the maximum tracks it.

Note that `Sel_norm_scope` has no effect on a one-sex species, or where
`Sel_norm_bin` is `"Off"`. Before Rceattle 5.8.0 a named bin always implied
`"WithinSex"` and max-normalization always implied `"AcrossSexes"`; a two-sex
fleet normalizing at a named bin should set `Sel_norm_scope = "WithinSex"`
explicitly to keep its old behaviour.

#### Which forms can hold a sex difference in level

Normalization *preserves* a level ratio; it cannot create one. `"AcrossSexes"`
divides both sexes by one pooled scalar, so whether the sexes can reach
different levels at all is a property of the **form**, not of these two columns:

| Form | Sex difference in level, from the form alone |
|---|---|
| `Logistic`, `DescendingLogistic`, `LogisticPM` | **none**: the curve tends to 1 for either sex, whatever its parameters |
| `NonParametric`, `NonParametricPM`, `NonParametricIntegrable` | **none**: each sex is re-centred so its own mean selectivity-at-age is exactly 1, every year |
| `Hake` | **none**: normalizes within sex by its own maximum; `Sel_norm_scope` is not read (`LogisticPM` likewise: its `Sel_norm_bin` is a penalty range) |
| `DoubleNormal` | **old-age plateau only**: the right-tail floor is sex-specific, but both sexes peak at exactly 1 |
| `DoubleLogistic`, `2DAR1`, `3DAR1` | **free**: the height is a free function of that sex's own parameters |

**Every row above describes the form on its own.** Since 5.38.0 an `apical`
selectivity linkage adds a per-sex height multiplier to any estimated form, so a
"none" here means the shape cannot hold the difference, not that the model cannot.
The multiplier is applied after the form's own centring or normalization.

**The form's own parameters are already per sex, and stay that way.** Each sex
gets its own copy: a two-sex `Logistic` fleet estimates a slope and an
inflection for each. An `apical` linkage adds a height on top of two
already-free shapes, so that fleet estimates five parameters -- slope and
inflection per sex, plus one multiplier on the sex named. This is where the
Stock Synthesis analogy stops. An SS male offset is usually an offset from the
female *shape*, so the shape is shared and one parameter carries the sex
difference; here both shapes stay free and the offset carries only the height.
The distinction matters when the sexes' compositions are thin: each shape is
informed by every row for that sex, but the height is informed only by the
joint-sex rows (`Sex = 3`). Sharing a shape across the sexes is not a switch
today -- it needs a hand-edited `map` that points one sex's cells at the
other's.

Measured on `GOAatf`'s two-sex fishery, as the ratio of male to female maximum
selectivity: `Logistic` 1.00, `DoubleLogistic` 1.97, `2DAR1` 4.53. The same
`DoubleLogistic` fit under `"WithinSex"` returns to 1.00, which is the column
doing its job, not the form losing the difference.

Going non-parametric does **not** open this up: those forms pin each sex's mean
to 1, so a level ratio is a by-product of shape rather than something the data
can inform. `inst/dev/two-sex-selectivity-example.R` fits the
comparison.

Two things constrain what is actually estimable:

* **The composition data must be joint.** `comp_data$Sex = 3` stacks female and
  male composition into a single vector that is normalized to sum to 1 as a
  whole (as Stock Synthesis does), so the sex ratio is fit and the relative
  selectivity is informed. `Sex = 1` or `2` put each sex in its own row, each
  summing to 1 independently: those hold **no** sex-ratio information, and
  `Sel_norm_bin` then makes no difference however it is set. There is no
  separate sex-ratio likelihood component.
* **`Sel_norm_scope` is not read by `Selectivity = "Hake"` or `"LogisticPM"`**,
  because neither goes through the shared normalizer; `data_check()` says so for
  a two-sex fleet on the default `"AcrossSexes"`. The reason differs by form.
  Hake normalizes each sex to its own maximum in its own block, so both sexes
  reach 1 at their own peaks and the normalization cannot express a level
  difference. LogisticPM does not normalize at all, and reuses `Sel_norm_bin` as
  a penalty bin range. On either form an `apical` linkage can still set the
  per-sex level: the multiplier is applied after each form's own rescaling, so
  Hake's within-sex normalization does not remove it.

Related settings elsewhere: the sex ratio at recruitment is **fixed input**, not
estimated, it comes from the `sex_ratio` sheet at the first modelled age.
Sex-specific natural mortality is `build_M1(M1_model = )` (§7) and sex-specific
growth is `build_growth()` (§8). The per-sex hooks in the linkage grammar are
`linkage_spec(by = ~ sex)` and its `sex =` filter.

Note that `fleet_control` also holds a `Sex` column. It is **inert**: read
nowhere in the model, and is retained only so older workbooks still read. Sex
is set per observation on `comp_data`, not per fleet.

## 4. Catchability (per fleet)

> Equations: see vignette 8 [§Catchability parameterizations](model-parameterizations.html#catchability-parameterizations).

Set the `Catchability` column of `fleet_control`:

| Value | String | Description |
|:---:|---|---|
| `0` | `"Fixed"` | q supplied directly |
| `1` | `"Estimated"` | q estimated as a free parameter |
| `2` | `"Estimated-with-prior"` | Estimated with a lognormal prior on q: mean `Catchability_init` (median when `bias_adjust_proc = FALSE`), log-scale SD `Catchability_prior_sd` |
| `3` | `"Analytical"` | Closed-form analytical q (concentrated likelihood) |
| `4` | `"PowerEquation"` | Power-function catchability (q · B<sup>α</sup>). **Not implemented**: `data_check()` rejects it. |
| `5` | `"Environmental"` | q linked to an environmental index. **Deprecated in 4.9.0** (recorded in the 5.8.1 notes), express it as a covariate linkage instead: `build_catchability(linkages = list(q = linkage_spec(~ temp, by = ~ fleet)))`. The series is named by `Time_varying_q`, not `Catchability_index`. |
| `6` | `"AR1"` | **Removed in 5.12.0**: `data_check()` errors on it. It never worked: the log-q deviates were gated on `Time_varying_q`, which under this form holds an `env_data` column index rather than a mode, so q came back constant. Express the Rogers et al. (2024) form as a q linkage instead, see below. |
| `7` | `"AnalyticalArith"` | q solved from the fleet's own index observations using the AMAK arithmetic-mean estimator; pair with `Index_distribution = "MVN"` for a supplied covariance |

### Index likelihood (`Index_distribution`)

How an index observation is scored, per fleet. This is the one switch whose
choice also changes how residuals and plots are computed, so it is worth setting
deliberately rather than inheriting.

| Value | String | Scale | Description |
|:---:|---|---|---|
| `0` | `"Lognormal"` | log | Independent lognormal on `log(obs)`. The historical default. `Log_sd` is a log-scale CV. |
| `1` | `"MVN"` | natural | The AMAK/`ebswp` `DoCovBTS = 1` bare quadratic form `0.5 · r' Σ⁻¹ r` on the residual `obs − q·pred`. Drops the normalizing constant, so the reported value matches ADMB. Needs `index_cov`. |
| `2` | `"MVNORM"` | natural | The full multivariate-normal negative log-density. Identical fit to `"MVN"` (the extra term is constant), but properly normalized. Needs `index_cov`. |
| `3` | `"Normal"` | natural | Normal on `obs − q·pred`, with `Log_sd` read as an **absolute** natural-scale sd, not a CV. No lognormal bias correction. |
| `4` | `"TruncatedNormal"` | natural | The same normal, left-truncated at zero. An index cannot be negative, so this is the only natural-scale family whose simulator and likelihood are the same distribution. **Prefer it over `"Normal"`** unless you need an exact ADMB comparison. |

The `MVN` families take a user-supplied variance-covariance matrix (e.g. a
VAST-derived Σ) through `index_cov`, and pair naturally with
`Catchability = "AnalyticalArith"` for the AMAK arithmetic-mean q.

**A family's scale is recorded in two places.** Adding one to
`index_distribution_map` is not enough: it must also be classified in
`.index_rows_natural_scale()`, which is what `residuals(type = "pearson")`,
`plot_index()`'s observation interval and `plot_indexresidual()` read. A
natural-scale family that misses it does not error, it silently gets the
log-scale formula, whose `σ²/2` term is then a number the size of the index
squared.

Time-varying behavior is set by `Time_varying_q` (per fleet):

| Value | String | Behavior |
|:---:|---|---|
| `0` | `"Off"` | Time-invariant |
| `1` | `"IID"` | Independent annual deviations |
| `2` | `"AR1"` | **Removed in 5.16.0**: `data_check()` errors on it. It was never an AR1: the model scored it with the same independent normal penalty as `"IID"`, no correlation parameter exists for this switch (the one that did belonged to the removed `Catchability = "AR1"` form). Set `"IID"` for the fit it actually gave, or see the replacement below. |
| `3` | `"Block"` | Time block |
| `4` | `"RandomWalk"` | Random walk |

`Time_varying_q` is **soft-deprecated as a whole**, as `Time_varying_sel` is,
and `data_check()` says so on use. The linkage grammar expresses the same IID,
random-walk and AR(1) deviations, and additionally lets the deviation standard
deviation be estimated, given a prior, or held fixed:

```r
build_catchability(linkages = list(
  q = linkage_spec(~ ar1(1 | Year), by = ~ fleet)))
```

The removed `Catchability = "AR1"` (QAR1, Rogers et al. 2024) is the same
formula plus the series the deviations are *observed* against: that is what
makes it the state-space form rather than a free AR(1) on q:

```r
build_catchability(linkages = list(
  q = linkage_spec(~ ar1(1 | Year), by = ~ fleet, fleet = <Fleet_code>,
                   observe = "<that fleet's env_data column>",
                   obs_sd  = <its measurement SD>)))
```

`obs_sd` is that series' measurement standard deviation, which the old switch
never asked for. GOA pollock 2025 is a worked example.

If `Catchability` is `Environmental`, `Time_varying_q` can be an integer or character of integers
specifying the environmental series in `env_data` to use (e.g. `Time_varying_q = 1` or `Time_varying_q = "1,2,3"`).

### Which fleets get a catchability

Catchability follows the **data**, not the fleet type. A fleet is given a q if it
holds `index_data` the model fits, a fishery with a CPUE series as much as a
survey. A fishery's index is predicted from that fishery's own selectivity, which
is what CPUE should share; an index needing a *different* selectivity has to be
its own fleet, since `sel_at_age` is one block per fleet (the same is true of
`Observation_units`).

A fleet with no fitted index rows gets **no** q, whatever `Catchability` says,
a catchability with no index to inform it is a flat direction in the likelihood.
This applies to a composition-only survey exactly as it does to an ordinary
fishery. `print()` marks the case so the column does not read as authoritative:

```
[2] Fishery │ Fishery │ sel: Logistic │ q: Estimated (fixed: no index data)
```

The exception is sharing. Fleets with the same `Catchability_index` share one q
parameter, so the group can only hold one answer to "is it estimated?", and the
**lead** fleet (the group's first non-`Off` fleet) decides for all of them,
regardless of type. A fleet with no index of its own is therefore still estimated
if its group's lead is, and a fleet whose own `Catchability` is `"Estimated"` is
still fixed if its lead is `"Fixed"`. Give a fleet its own `Catchability_index`
when it should be estimated independently.

The exception is `"Analytical"` and `"AnalyticalArith"`, which solve q from each
fleet's own index observations rather than from the shared parameter. A group
containing one does **not** share a catchability, even when every fleet in it
holds the same setting, since each still solves separately. Give those fleets
their own `Catchability_index`, or use an estimated q for the whole group.

`"Environmental"` *does* share: it rebuilds q from the group's shared parameters
(`index_log_q`, `index_q_beta`), so the whole group follows the lead fleet,
including when the fleets name different environmental series in
`Time_varying_q`, where the lead's series is the one used.
A `"Fixed"` lead is a third case: nothing is estimated for the group and nothing
is shared, so each fleet sits at its own `Catchability_init`. Where a q *is*
estimated for the group, its starting value is the mean of the group's
log `Catchability_init` (TMB averages the initial values of parameters mapped
together) rather than the lead's; only the prior centre uses the lead's.
`build_map()` warns when those inits differ, and names the geometric mean the
group will start from. Worth heeding: a shared q starting at the mean scales a
survey's whole predicted index by a constant factor, which no residual pattern
distinguishes from a real change in abundance.

`Time_varying_q_sd` is shared the same way once it is estimated, which is
`random_q = TRUE` with a `Time_varying_q` of `"IID"` or `"RandomWalk"`:
the group gets one deviation sd starting at the geometric mean of its estimated
members' values, and `build_map()` warns when they differ. While the sd is
fixed, each fleet keeps its own. `Catchability_prior_sd` is a prior sd the
assessor sets and is never estimated, but in a shared group only the **lead**
fleet's is read: the prior is scored once, on the lead, so a differing value on a
non-lead fleet has no effect at all.

`data_check()` warns on the analytical case, on a `Fixed` lead that leaves fleets
on different inits, and on a `Catchability` or `Time_varying_q` that differs
within a group.

### How an index is timed against mortality

An index is predicted from the numbers its fleet sees, and that depends on the
kind of observation. The model keys this on `Fleet_type`; there is no column to
set.

A **survey** index is a snapshot at the month given in `index_data`:

$$\hat{I}_y = q \sum_a s_a\, N_{a,y}\, e^{-(\text{Month}/12) Z_{a,y}}\, w_a$$

A **fishery** index is CPUE, which integrates over the year alongside the catch.
Baranov gives $C_a = F_a \bar{N}_a$, and effort cancels the $F$ (since
$F_a = q_e E s_a$):

$$\hat{I}_y = q \sum_a s_a\, \bar{N}_{a,y}\, w_a,
\qquad \bar{N}_{a,y} = N_{a,y}\frac{1 - e^{-Z_{a,y}}}{Z_{a,y}}$$

$\bar{N}$ is the mean-numbers term the catch equation uses, and the one a
fishery's age composition is already built on, so a fleet's index, catch, and
comps are consistent by construction. `Month` is not read for a fishery's index
rows.

This affects trend, not just scale, so a constant $q$ cannot absorb it. At
`Month = 6`, $e^{-Z/2}$ approximates $\bar{N}/N$ to within 1% at $Z = 0.5$ but is
9.6% low at $Z = 1.5$; at `Month = 0` the snapshot factor is $e^0 = 1$, against
$\bar{N}/N$ of 0.906 at $Z = 0.2$ and 0.432 at $Z = 2.0$. A seasonal fishery
needs no special treatment, its exact window predictor differs from the
year-average by a near-constant factor $q$ takes up (1–3% trend error across
$F$ 0.05–0.8 for $M$ between 0.1 and 0.5, against 29–33% for a snapshot).

### A snapshot-timed index on a fishery's selectivity

This is the AMAK/ebswp CPUE form, and what an ADMB bridge needs. Give the index
its own `Survey` fleet and point that fleet's `Selectivity_index` at the fishery:
fleets sharing a `Selectivity_index` share one selectivity block, so the index
gets the fishery's selectivity along with its own `Month` and its own q.

Build the mirror fleet by copying the fishery's `fleet_control` row, see
[Sharing a selectivity between fleets](#sharing-a-selectivity-between-fleets)
for the columns that have to agree, and why a hand-authored row can silently get
a different curve.

## 5. Composition likelihoods

Set `Comp_distribution` (or `CAAL_distribution`) per fleet:

| Value | String |
|:---:|---|
| `-1` | `"MultinomialAFSC"` (legacy AFSC convention) |
| `0` | `"Multinomial"` |
| `1` | `"DirichletMultinomial"` |

Diet composition uses an analogous switch on the bioenergetics control
sheet: `Diet_distribution = 0` (multinomial) or `1` (Dirichlet-multinomial).

Conditional age-at-length (CAAL) data go through their own data path
and are weighted via `CAAL_weights` per fleet.

### Ageing error, per fleet

`age_error` holds one matrix per species by default, and every fleet reads its
own species'. A stock whose ageing protocol changed part way through a series
needs more than one — GOA Pacific cod reads its pre-2007 survey otoliths about a
year older than its post-2007 ones — so both `age_error` and `fleet_control`
take an optional `Ageing_error_index`: give the matrices index values, then
point each fleet at the one that aged its samples. Omit the column, or leave it
`NA`, and the index is the species, which is what a model with one matrix per
species has always done.

The two eras of a survey are then two fleets sharing a `Selectivity_index` and a
`Catchability_index`, so they still estimate one selectivity and one
catchability between them.

Each index belongs to one species, because a matrix's rows are sized and offset
by that species' age range. `data_check()` refuses an index that names no matrix,
an index whose rows give more than one `Species`, and a fleet pointed at a matrix
belonging to another species — which is the mistake the species default invites,
since without the column `2` means "species 2's matrix" and not "the second
matrix". Coverage is reported per matrix, so an incomplete one is named even
where the species' other matrices are complete.

## 6. Recruitment / stock-recruit (`recFun = build_srr()`)

| `srr_fun` | String | Description |
|:---:|---|---|
| `0` | `"mean"` | Mean recruitment with deviations |
| `2` | `"BevertonHolt"` | Beverton-Holt |
| `4` | `"Ricker"` | Ricker |

Priors and environmental relationships can be added to stock-functions
via `build_srr()`. See `?build_srr` for options and
`vignette("environmental-linkages-and-priors")` for details.

Under multispecies mode (`msmMode > 0`) spawning biomass per recruit is
undefined, because mortality includes predation. A curve fitted in the
hindcast then estimates its initial recruitment level instead of deriving it,
and the choice of `initMode` can move the fitted stock substantially; the curve
can also enter as a recruitment penalty (`srr_fun = "mean"` with the curve as
`srr_pred_fun`). A prior on Beverton-Holt steepness is refused.

## 7. Natural mortality (`M1Fun = build_M1()`)

### `M1_model`, fixed-effects structure of M1

| Value | String | Description |
|:---:|---|---|
| `0` | `"fixed"` | use the input `M1_base` (no estimation).
| `1` | `"sex_age_invariant"` | estimate one `M1_{spp}`.
| `2` | `"sex_specific"` | estimate `M1_{spp, sex}`.
| `3` | `"sex_age_specific"` | estimate `M1_{spp, sex, age}`.

Priors and environmental relationships can be added to M1 
via `build_M1()`. See `?build_M1` for options and 
`vignette("environmental-linkages-and-priors")` for details. 

### `M1_re`, random-effects structure on M1

| Value | String | Description |
|:---:|---|---|
| `0` (default) | `"none"`  | No random effects |
| `1` | `"iid_age"` | IID by age, constant over years |
| `2` | `"iid_year"` | IID by year, constant over ages |
| `3` |  `"iid_age_year"` | IID across both year and age |
| `4` | `"ar1_age"`  | AR(1) by age, constant over years |
| `5` | `"ar1_year"`| AR(1) by year, constant over ages |
| `6` | `"ar1_age_year"` | 2D AR(1) over both year and age |

`M1_re > 0` requires an `M1_model` of `0`, `1` or `2`, and is refused
for `3`, `4` and `5` -- those levels have no deviation block, so the
request would silently give constant M. Under `M1_model = 0` the input
schedule is held fixed and the deviations carry all of M's variation,
which makes them the most identifiable of any level. Under `M1_model = 3` the
age-specific level and an age-varying deviation are not separately
identifiable; for `4` and `5` the linkage grammar expresses the same
model as `linkage_spec(~ temp + (1 | Year))`.

Variance and correlation parameters are species-specific but
sex-invariant. `M1_dev_log_sd` is the AR(1) **innovation** standard deviation,
the marginal is `sigma / sqrt(1 - rho^2)`, as in WHAM. (The linkage grammar uses
the opposite convention: `linkage_spec(~ ar1(1 | Year), init = ...)` takes
`sigma` as the marginal SD.) `M2_use_prior = TRUE` adds a log-normal prior with
mean `M_prior` (the median when `bias_adjust_proc = FALSE`) and log-scale SD
`M_prior_sd` on total M (M1 + M2) in multi-species mode.

Modes `3` and `6` estimate a full age × year deviation field. Before 5.9.0 a map
defect gave them one deviation per *year* laid on a stride pattern instead (42
free deviations where 882 were meant, on `GOA2018SS`) so fits in these two modes
change. They are also the two to treat carefully: a free age-by-year mortality
field with a single standard deviation and no prior on M is strongly confounded
with selectivity and recruitment in a single-species assessment. Read
`fit$convergence` and the estimability table before trusting one, and consider a
prior on M. Modes `1`/`2`/`4`/`5` are unaffected; they hold the deviation
constant along one dimension deliberately.

## 8. Growth and weight-at-age (`growthFun = build_growth()`)

### `growth_model`, fixed-effects structure of growth

| Value | String | Description |
|:---:|---|---|
| `0` (default) | `"empirical"` |  Empirical weight-at-age supplied via the `weight` data sheet. Forecasts re-use the terminal-year weight schedule. |
| `1` | `"vonBertalanffy"` | von Bertalanffy length-at-age + length-weight allometry (`alpha_wt_len`, `beta_wt_len` on the data control sheet) |
| `2` | `"Richards"` |Sex-specific Richards growth |

### Growth variability, plus group, length bins and maturity

| Argument / column | Values | Description |
|---|---|---|
| `sd_form` | `"SD"` (default), `"CV"` | Whether `sd_L1` / `sd_Linf` are SDs of length-at-age (cm) or CVs (SD = CV x length) -- SS3 `CV_Growth_Pattern` 2 or 0 |
| `sd_plus_group` | `"WHAM"` (default), `"SS3"` | Plus-group SD pinned to the upper endpoint, or interpolated by length. SS3 with `Growth_Age_for_L2 = 999` is `"WHAM"` |
| `plus_group_length` | `"M1"` (default), `"none"`, `"SS3.24"`, `"decay"` | Plus-group mean length: M1-weighted mean toward L-infinity, or SS3's `Linf_decay` -998 / -999 / `plus_group_decay` |
| `pop_lengths` | lower edges (cm); default the data bins | Population length bins on which the age-length key, weight and maturity are integrated |
| `L50_mat_len`, `slope_mat_len` (control sheet) | cm, per cm; default NA | Logistic maturity-at-length; spawning output integrates maturity x weight over length. NA keeps the age-based `maturity` sheet |

Priors and environmental relationships can be added to growth
via `build_growth()`. See `?build_growth` for options and
`vignette("environmental-linkages-and-priors")` for details.

Time-varying growth is specified through the linkage grammar rather than a
switch of its own, `build_growth(linkages = list(K = linkage_spec(~ rw(1 | Year))))`
puts a random walk on the von Bertalanffy `k`, and `ar1()` / `(1 | Year)` give
the autoregressive and independent alternatives.

The length-based suitability modes (`suitMode = 1/3/5`) are not yet
enabled; use a weight-based mode (`2/4/6`) for parametric suitability.
When the length-based modes are enabled, prey length-at-age is taken
from the estimated growth model (`growth_model > 0`).

## 9. Harvest control rules (`HCR = build_hcr()`)

| `HCR` | String | Description | Multi-species? |
|:---:|---|---|:---:|
| `0` | `"NoFishing"` | No fishing, estimate hindcast only | Yes |
| `1` | `"CMSY"` | Maximize joint catch across species (option to keep depletion ≥ `Plimit`) | Yes |
| `2` | `"ConstantF"` | Constant F set at `Ftarget` per species | Yes |
| `3` | `"ConstantFSSB"` | F that achieves `Ftarget`% of SB<sub>0</sub> at the end of the projection | Yes |
| `4` | `"ConstantFSPR"` | F<sub>SPR</sub> at `Ftarget` (NEFSC convention; scale via `Fmult`) | No |
| `5` | `"NPFMC"` | NPFMC Tier 3 SPR-based HCR (`Ftarget`, `Flimit`, `Ptarget`, `Plimit`, `Alpha`) | No |
| `6` | `"PFMC"` | PFMC Category 1 `Ptarget`-`Plimit` ACL with uncertainty buffer around `Flimit` (`Pstar`, `Sigma`) | Yes |
| `7` | `"SESSF"` | SESSF Tier 1 SPR-based HCR | No |

`DynamicHCR = TRUE` switches from static to dynamic SB<sub>0</sub>
reference points. `HCRorder` controls the order in which species F is
solved for iterative multi-species HCRs (e.g. predators before prey).
See `?build_hcr` for more details.

## 10. Projection and MSE

| Function | Purpose |
|---|---|
| `fit_mod(estimateMode = 2, HCR = build_hcr(...))` | Forward projection under a fixed HCR |
| `sample_rec()` | Bootstrap historical recruitment deviations (with optional `rec_trend`) into the projection |
| `remove_F()` | Refit a model with F = 0 from `styr` (default: the year after `endyr`); `run_mse()` uses it for the no-fishing run behind the collapse metrics |
| `run_mse()` | Closed-loop MSE: refit EM on the `assessment_period` schedule (a period, or the years themselves), sample new data at `sampling_period` |
| `load_mse()` | Reload per-simulation `.rds` files written when `dir =` is supplied |
| `check_mse()` | Validate which OM/EM simulations converged |
| `mse_summary()` | Per-fleet performance metrics: mean catch, IAV, P(closed), P(F > Flimit), P(SSB < SSBlimit), terminal depletion |

`run_mse()` parallelizes over simulations on a PSOCK cluster. See the
[MSE vignette](hcrs-and-mses.html) for the
seed plumbing and reproducibility guarantees.

## 11. Diagnostics and inference

| Function | Purpose |
|---|---|
| `retrospective()` | Peel terminal years and refit; reports Mohn's ρ for each quantity / species, optionally with forecast-skill peels (`nyrs_forecast`) |
| `jitter()` | Refit from random-perturbed starting values; check optimum stability |
| `sim_mod()` / `simulate()` | Simulate data from a fitted model. Observations always; process error (recruitment, M, growth, catchability, selectivity) via `process =`. `simulate()` is the `stats` generic and draws `nsim` replicates at once |
| `self_test()` | Simulate and refit; check the model recovers what generated the data |
| `compare_sim()` | Estimation-vs-truth comparison across simulated replicates |
| `model_average()` | Weighted average of derived quantities across multiple fits, with optional bootstrap uncertainty |
| `calc_mcall_ianelli()` / `calc_mcall_ianelli_diet()` | McAllister-Ianelli composition-data reweighting |

## 12. Plotting

Pop / mortality / fit / predation / biology plots, all named after the
quantity they show: `plot_biomass`, `plot_ssb`, `plot_recruitment`,
`plot_depletion`, `plot_depletionSSB`, `plot_exploitable_biomass`,
`plot_catch`, `plot_f`, `plot_mortality`, `plot_m_at_age`,
`plot_m2_at_age_prop`, `plot_index`,
`plot_indexresidual`, `plot_comp`, `plot_data`, `plot_form`,
`plot_b_eaten`, `plot_b_eaten_prop`, `plot_diet_comp`, `plot_ration`,
`plot_selectivity`, `plot_selectivity_vs_maturity`, `plot_maturity`,
`plot_stock_recruit`, `plot_timeseries`. Most accept a single fitted
model or a list of models with `model_names`.

These are built with **ggplot2** (colorblind-safe Okabe-Ito palette for
discrete series, viridis for continuous ones, `theme_classic`) and **return
the `ggplot` object**, so they print when called interactively and can be
further customized by adding ggplot2 layers
(e.g. `plot_biomass(fit) + ggplot2::ggtitle("My title")`). Pass `file = "stem"`
to also write the figure to `stem_<plot>.png`.

**Shared arguments.** The timeseries plotters (`plot_biomass`, `plot_ssb`,
`plot_recruitment`, the depletions, `plot_exploitable_biomass`, `plot_f`), the
predation plotters (`plot_b_eaten`, `plot_b_eaten_prop`, `plot_m_at_age`,
`plot_m2_at_age_prop`, `plot_ration`) and `plot_selectivity` take one common
vocabulary, documented together in `?"rceattle-plot-args"`: `line_col`, `lwd`,
`lty`, `alpha`, `species`, `spnames`, `minyr`, `maxyr`, `incl_proj`,
`incl_mean`, `add_ci`, `model_names`, `file`, `width` and `height`. Each word
means the same thing wherever it appears, but not every plotter takes every one:
`incl_mean` is on the predation plotters, `add_ci` only where the quantity
holds standard errors, and `alpha` only where the figure has a ribbon or a
fan. Check `args()`. The other plotters still take their own arguments.

* `line_col` takes colour names, hex codes, or base-graphics palette indices
  (`line_col = 1`), and `lwd` keeps the base-graphics scale where the default
  `3` is a standard-weight line.
* `line_col` and `lty` supply values for whichever variable the *figure*
  separates by colour or line type, which is not always the model. In
  `plot_b_eaten_prop()` colour separates predators; in `plot_ration()` line type
  separates the sexes. Each function's help says which. Too few colours are
  recycled, with a warning naming what they coloured. Where colour is a
  magnitude (the year fan in `plot_selectivity()`) `line_col` gives the ramp
  anchors, so `line_col = "black"` draws the fan in black.
* `species` selects (by index, name, logical mask, or `"all"`, in the order
  given) and `spnames` labels.
* `minyr` and `maxyr` narrow the data, not just the axis, so a panel with a free
  y scale rescales to the window.
* Arguments left over from the base-graphics version, `right_adj`, `top_adj`,
  `mod_cex`, `legend.pos`, `single.plots`, `theta`, are still accepted, and
  ignored, by the plotters that had them; they were never added to the ones that
  did not. Set the equivalent on the returned ggplot instead, e.g.
  `plot_biomass(fit) + ggplot2::theme(legend.position = "right")`.

**Display units.** The model works in mt and thousands of fish: numbers-at-age
are in thousands and weight-at-age in kg, so biomass and SSB come out in mt and
recruitment in thousands of fish. The timeseries plotters rescale to
`million mt` and `millions` for the axis, and derive the recruitment age from
`minage` rather than assuming age 1. Supply the model's inputs on that
convention, catch and index in mt, weight-at-age in kg, or the axis labels
will not describe what is plotted. Pass `ylab` to override the derived label.

**Confidence intervals.** `add_ci = TRUE` draws a 95% interval. Strictly
positive series take it on the log scale, `exp(log(x) ± 1.96 · sd_log)`, using
the model's `log_biomass` / `log_ssb` / `log_R` where they exist and the
delta-method identity `sd(log x) = sd(x)/x` where they do not. The interval is
therefore right-skewed and cannot reach zero for a weak year class or a
depleted stock. See `?plot_timeseries` for the full derivation.

`plot_selectivity()` takes it too, but only on a fit run with
`fit_control(selectivity_se = TRUE)` and only for estimated, age-based fleets;
otherwise it warns and draws none. Selectivity is identified only up to a
scalar, so `Sel_norm_bin` fixes the curve's level rather than measuring it: the
pinned bin has a standard error of 0 and the band widens away from it. Quote
that anchor with the interval.

The error comes from whichever `sdreport` the fit ends on, and that is not
always the one that estimated the curve. Under `estimateMode = "Estimate"` with
an estimating HCR, `fit_mod()` re-optimizes the projection with every hindcast
parameter fixed, so the standard error of every selectivity is exactly 0.
`estimateMode = "Projection"` estimates no selectivity at all, and
`projection_uncertainty` cannot rescue it. Fit the hindcast alone, or ask for
uncertainty in the projection:

```{r sel-ci, eval = FALSE}
# Errors from the hindcast fit, the usual choice for a selectivity figure.
mod <- fit_mod(data_list = data, estimateMode = "Hindcast",
               fit_control = fit_control(selectivity_se = TRUE))
plot_selectivity(mod, add_ci = TRUE, minyr = 2015, maxyr = 2015)

# Or keep the HCR projection and reuse the hindcast parameters in it.
mod <- fit_mod(data_list = data, HCR = build_hcr(HCR = 5),
               fit_control = fit_control(selectivity_se = TRUE,
                                         projection_uncertainty = TRUE))
```

`fit_mod()` warns when the combination would report zeros, and
`plot_selectivity()` declines to draw a band that is 0 wide at every age rather
than render it as certainty. Give `minyr` / `maxyr` a single year on a
time-varying fleet: every year drawn gets its own band.

Two further conditions belong with any interval you report:

- **The normalization.** On `Atka2022` the survey pins age 4, and its standard
  deviation runs 0.33, 0.18, 0.12, 0.00, 0.14 over ages 1--5; the fishery pins
  nothing, normalizes to a mean of one, and is flattest mid-range instead. Same
  fit, same data, the shape of the band is a property of `Sel_norm_bin`, not of
  the information in the data.
- **The smoothing.** Under `random_sel = FALSE` the annual deviations are
  penalized rather than integrated, so these are penalized-likelihood errors
  conditional on `Sel_curve_pen1` / `Sel_curve_pen2` and on a fixed
  `Time_varying_sel_sd`. Change the smoothing and both the curve and its band
  move, with nothing in the data to arbitrate.

The band is marginal. Selectivity error is strongly correlated with F and q
through the same confounding, so propagating it into a derived quantity needs the
covariance (`fit_control(getReportCovariance = TRUE)`), not this interval alone.

## References

- Adams, G.D., Holsman, K.K., Barbeaux, S.J., Dorn, M.W., Ianelli, J.N.,
  Spies, I., Stewart, I.J., Punt, A.E. (2022). An ensemble approach
  to understand predation mortality for groundfish in the Gulf of
  Alaska. *Fisheries Research*, 251, 106303.
- Holsman, K.K., Ianelli, J., Aydin, K., Punt, A.E., Moffitt, E.A.
  (2016). A comparison of fisheries biological reference points
  estimated from temperature-specific multi-species and single-species
  climate-enhanced stock assessment models. *Deep Sea Research Part II*,
  134, 360-378.
- Wassermann, S.N., Adams, G.D., Haltuch, M.A., Kaplan, I.C., Marshall,
  K.N., Punt, A.E. (2025). Even low levels of cannibalism can bias
  population estimates for Pacific hake. *ICES Journal of Marine
  Science*, 82(1), fsae064.
- Cheng, M.L.H., et al. (2024). 2D / 3D AR(1) selectivity (citation in
  vignette 8).
- Rogers, L.A., et al. (2024). AR(1) catchability for GOA pollock
  (citation in vignette 8).
