---
title: "3. Model diagnostics"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{3. Model diagnostics}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("RCEATTLE_EVAL_VIGNETTES"), "true"),
  fig.width = 7,
  fig.height = 5
)
```

Rceattle provides a full set of model diagnostics:

- **convergence diagnostics** — attached automatically to every fit
- **S3 methods** — `residuals()`, `logLik()`, and the like
- **fit plots** — compare predicted to observed data
- **retrospective analysis** — peel off recent years to detect systematic bias
- **jitter testing** — perturb starting values to check the optimizer found a global minimum
- **self-testing** (simulation–estimation) — check the model recovers its own estimates from simulated data
- **likelihood profiling** — check how informative the data are about a given parameter

**They all report the same way.** Printing any of them opens with one line — the
object, an overall status, and a verdict — before the detail:

```
<Rceattle convergence>  status: OK
  all checks OK

<Rceattle OSA diagnostics>  status: WARN
  7 of 7 source(s) outside a null interval; overall SDNR 1.82 (0.91-1.09)
          source  fleet  n     sdnr   null
  [WARN]  index      4     36  2.52   0.77-1.23
  ...

<Rceattle retrospective>  status: WARN
  3 peel(s); 3 of 12 terminal Mohn's rho outside +/-0.2
          quantity  N     Pollock  Cod      Arrowtooth flounder
  [WARN]  R            3  0.511    0.283    -0.221

<Rceattle jitter>  status: OK
  4 of 4 start(s) reached the best optimum (within 0.01)

<Rceattle self-test>  status: NOTE
  47 of 50 simulation(s) converged
          convergence  n
  [OK  ]  OK           41
  [NOTE]  NOTE          6
  observations redrawn, processes held at their fitted values: the spread is a lower bound

<Rceattle profile>  status: WARN
  sigmaR over 29 grid point(s); all converged
  minimum -log L : 1274.8831  at slot_1 = 1.5
  within 1.92 of the minimum : 1.3 to >=1.5
  the minimum is at a grid edge (slot_1), so the grid does not bracket it -- widen `values`
```

The severity words are the same four everywhere — `OK`, `NOTE`, `WARN`, `FAIL` —
and the overall status is the worst present. **Adding a print method did not
change any return value**: each object is still the data frame or list it always
was, so `jit$nll`, `diag$sdnr`, `prof$grid$slot_1` and `sims[["Sim_1"]]` index
exactly as before, and `c(sims, list(fit))` still gives a plain list of fits to
plot. Only the printing has an opinion.

`retro$mohns` is the exception, and not because of printing: it was reshaped from
wide to long in 5.23.0.9007, so a read by position or by species-column name
needs updating. See *Retrospective analysis* below, and `NEWS.md` for the
one-call way back to the old layout.

Four of these carry a threshold you can set. `print(retro, band = )` is Mohn's
rho threshold — the default `0.2` is the rule of thumb below, and Hurtado-Ferro
et al. (2015) give asymmetric, life-history-dependent alternatives (−0.15 to
0.20 long-lived, −0.22 to 0.30 short-lived), which is why it is an argument.
`print(jit, tol = )` is how close an objective must be to count as having
reached the same optimum. `print(sims, rate = )` is the fraction of simulations
that must converge before a self-test counts as clean. `print(prof, cutoff = )`
bounds the reported interval; the default 1.92 is the 95% profile-likelihood
cutoff for one parameter — half the 0.95 quantile of a chi-square on one degree
of freedom.

Two of them report something the numbers alone cannot. A self-test drops its
non-converged runs before returning, so the length of what comes back is not the
number attempted — the header gives the rate against `nsim`, and the last line
says what generated the replicates, since with processes held fixed the spread
carries observation error only. A profile whose minimum sits at the first or
last grid value has run out of grid rather than found the optimum; it plots as
an ordinary curve, so the print says so and asks for a wider `values`.

## Setup and plotting data

```{r}
library(Rceattle)

# Fit a single-species model for the 2022 Northern Rockfish assessment.
data("NorthernRockfish2022")
nrdata <- NorthernRockfish2022
nrdata$fleet_control$Proj_F_proportion <- 1
plot_data(nrdata)
```

Fit the model.
```{r}
model_1 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    updateM1 = TRUE,
    M1_model = 1,
    M1_use_prior = TRUE,
    M_prior = 0.06,
    M_prior_sd = 0.05),
  fit_control = fit_control(phase = TRUE, verbose = 0)
)
summary(model_1)
```

## S3 Methods

```{r}
summary(model_1)              # same compact summary
coef(model_1)                 # estimated fixed-effect parameters
logLik(model_1)               # logLik with df attribute (AIC works)
AIC(model_1)
vcov(model_1)                 # fixed-effect covariance from sdreport
residuals(model_1)                    # response residuals, all data sources
residuals(model_1, type = "pearson")  # Pearson residuals (glm-style `type`)
as.data.frame(model_1)
```


## Convergence diagnostics

After optimization, `fit_mod()` runs a battery of convergence checks and attaches
the result as `model$convergence`. Each check is one record with a `severity`
(`"OK"`, `"NOTE"`, `"WARN"`, or `"FAIL"`); the object's `status` is the worst
severity present. Non-OK checks are reported via `message()` during the fit, and
`print(model)` shows the overall status. A non-converged model is never turned
into an error — the fit is always returned with its diagnostics attached, so you
decide whether to trust it.

```{r}
model_1$convergence            # overall status + any non-OK checks
print(model_1$convergence, all = TRUE)   # show every check, including OK ones
```

Re-run the battery on any fit with `convergence_diagnostics()`:

```{r}
convergence_diagnostics(model_1)
```

The checks cover:

- **`max_gradient`** — maximum absolute marginal gradient and the parameter
  carrying it (`WARN` > 1e-3, `FAIL` > 1).
- **`pdHess`** — whether the Hessian is positive definite (`FAIL` if not).
- **`sdreport_failed`** — `FAIL` when an `sdreport` was requested but the
  Hessian could not be inverted.
- **`hessian_conditioning`** — the Hessian condition number and, when poorly
  conditioned, the parameters loading on the least-identified direction (the
  unidentified linear combination). Complements `TMBhelper::check_estimability`
  (a per-parameter verdict) with a continuous severity and direction.
- **`parameters_on_bounds`** — parameters that hit a configured `build_bounds()`
  limit (often unidentified or mis-scaled).
- **`phasing`** — phases that ended with a high gradient, localizing which
  parameter block is hard to fit.
- **`estimability`** — surfaces `TMBhelper::check_estimability` when it ran.

## Fit plots

### Composition data

`plot_comp()` draws composition Pearson-residual bubbles and overlays the
observed (shaded area) and predicted (line) age or length compositions for every
fleet, both annually and aggregated across years. Joint-sex data are drawn on a
single bin axis with females above and males below zero.

```{r}
plot_comp(model_1)
```

### Survey indices

`plot_index()` shows observed vs. predicted survey biomass; pass `log = TRUE` for the log scale. `plot_indexresidual()` plots the residuals as observed minus predicted, on whichever scale the fleet is fitted on (log for a `Lognormal` fleet, the natural scale for the others) — a useful first check for time trends or heteroscedasticity. A positive residual means the survey saw more than the model predicted. These return `ggplot` objects.

```{r}
plot_index(model_1)
plot_index(model_1, log = TRUE)
plot_indexresidual(model_1)
```

### Catch

```{r}
plot_catch(model_1)
```

## One-step-ahead (OSA) residuals

Pearson residuals on composition data are *not* standard normal even when the
model is correct: the age/length bins are correlated (more fish in one bin means
fewer in another), and random effects induce correlation across years. One-step-ahead
(OSA) residuals remove both problems. Each observation is residualized conditional
on the observations added before it, integrating out the random effects, so under a
correctly specified model the residuals are iid standard normal (Thygesen et al. 2017;
Trijoulet et al. 2023). This makes them suitable for objective goodness-of-fit testing.

`osa_residuals()` computes them *post hoc* for any combination of the fitted data
types -- survey indices (`"index"`), fishery catch (`"catch"`), age/length
composition (`"comp"`), conditional age-at-length (`"caal"`), predator diet
(`"diet"`), and, where a linkage carries an observed state-space covariate, the
covariate observation itself (`"ecov"`). Composition data use the conditional
binomial / beta-binomial decomposition. Because OSA re-optimizes the random
effects for each observation it is expensive, so run it on a converged fit
(`estimateMode < 3`). All sources are available from any fit; `osa_residuals()`
builds the required observation data on demand and skips a requested source the
model has no observations for (it errors only if *none* of them do). `"diet"` is
the one source not in the default set -- it applies only to multispecies models
and can be slow -- so pass `source = "all"` to include it.

```{r osa}
# Aggregate (catch + index) residuals
osa <- osa_residuals(model_1, source = c("index", "catch"))
head(osa)

# Statistical diagnostics (Stewart & Monnahan 2025): SDNR and lower/upper tail
# statistics, each with the interval expected under the standard-normal null.
osa_diagnostics(osa)

# Q-Q plot (with SDNR / tail annotation) + residual-by-year
plot(osa)

# residuals() also exposes them in the common residual schema (`source` selects
# the data, like `type` in stats::residuals.glm()):
residuals(model_1, type = "osa", source = c("index", "catch"))

# Composition OSA residuals -- no special fit required
plot_comp(model_1, residual_type = "osa")              # Q-Q + signed bubbles
residuals(model_1, type = "osa", source = "comp")
```

### Fleets with tail accumulation

Where a fleet folds its composition tails into a boundary bin
(`Comp_accum_young` / `Comp_accum_old`), both the OSA and the Pearson residuals
describe the bins the likelihood actually fit, not the raw bins in `comp_data`.
A fleet on 10 ages with `Comp_accum_young = 3` reports bins 3 to 10: bin 3 holds
ages 1 to 3 combined, and the `accumulated` column (`Accumulated` from
`residuals()`) marks it, since it stands for a range rather than the single age
it is named for. Expect fewer residual bins than raw age bins for such a fleet.

The two differ in one place. The one-step-ahead decomposition drops each
composition's last bin, which is fixed by the sum-to-sample-size constraint.
Under an *old*-tail accumulation that dropped bin is the upper accumulated one,
so the fleet's upper boundary bin appears in the Pearson residuals and not in
the OSA residuals.

Following Stewart and Monnahan (2025), a practical workflow is:

1. **Inspect the aggregate fit** across years (`plot_comp(model_1)`): systematic
   lack of fit here points to structural misspecification (e.g. the wrong
   selectivity shape).
2. **Compare the SDNR** to its null interval (from `osa_diagnostics()`): a value
   outside the interval flags over- or under-dispersion, often a data-weighting
   or effective-sample-size issue.
3. **Compare the lower/upper tail statistics** to their null intervals: these
   localize departures to the tails of the residual distribution.
4. **Inspect Pearson bubble plots** (`plot_comp(model_1)`, the default): patterns
   in the sign of residuals across age/year can identify *where* lack of fit
   occurs -- Stewart and Monnahan recommend keeping these alongside OSA
   residuals.
5. **Revise the model or data weighting** where misspecification is found, using
   this together with the other diagnostics rather than as an accept/reject test.

When the diagnosis is a weighting problem rather than a structural one,
`reweight_comps()` runs the McAllister-Ianelli (1997) tuning loop: it refits the
model the way it was fitted, updating each fleet's `Comp_weights` until the
implied weights settle.

```{r}
tuned <- reweight_comps(model_1, n_iter = 10, tol = 0.01)
tuned$reweight$history    # weight per fleet, per iteration
```

Restrict it to particular fleets with `fleets =`, taking either `Fleet_code`
values or names matching `fleet_control$Fleet_name` — give ids or names, not a
mix, since R coerces `c(1, "Survey")` to character.

```{r}
tuned <- reweight_comps(model_1, fleets = c("Survey_1", "Fishery_1"))
```

Only multinomial fleets are tuned. A Dirichlet-multinomial fleet estimates its
own weight inside the likelihood, so tuning it externally would compete with
that estimate; those fleets are named and skipped.

The alternatives are Francis (2011) reweighting, applied by hand, and estimating
the weights directly with a Dirichlet-multinomial likelihood
(`Comp_distribution = "DirichletMultinomial"`), which avoids the iteration
entirely — note that under that likelihood `Comp_weights` is read as the **log**
of the starting weight, so the model default of `1` is a starting weight of
*e*, and a weight of 1 is written as `0`.

Because OSA residuals for discrete composition data are randomized quantile
residuals, they are stochastic; `osa_residuals()` takes a `seed` for
reproducibility.

### Process residuals

`process_residuals()` is the complementary check on the *process* model. It
standardizes the model's random-effect deviations (recruitment, initial
abundance, catchability) against their process prior, drawing once from the
joint posterior (SAM's `procres`; Nielsen and Berg 2014). Under a correct
process model these too are approximately iid standard normal.

```{r procres}
pr <- process_residuals(model_1, process = "recruitment")
osa_diagnostics(pr)
plot(pr)

# residuals() exposes them too:
residuals(model_1, type = "process")
```

## Retrospective analysis

A retrospective analysis removes the most recent years of data one peel at a time and re-fits the model. Mohn's rho summarizes how the terminal estimates shift across peels; for SSB, a rho outside ±0.2 generally signals a problem worth investigating.

```{r}
model_1_retro <- retrospective(model_1, peels = 5)

# Mohn's rho for each quantity
model_1_retro$mohns

# Plot historical trajectories across peels
plot_biomass(model_1_retro$Rceattle_list)

# Include the projection period to see how the forecast changes
plot_biomass(model_1_retro$Rceattle_list, incl_proj = TRUE)
```

Each peel carries its own terminal year as `data_list$endyr`, so the first plot draws every peel only as far as it was fit and the peels fan out.

A peel still estimates the years it dropped: they are its retrospective forecast, fit to the observed catch with recruitment held at the peel's mean and the survey and composition data withheld. Fishing mortality is estimated through the unpeeled terminal year and the harvest control rule takes over after it, so each peel carries three years:

| Field | Meaning |
|---|---|
| `endyr`, `endyr_peel` | the peel's terminal year — what it was fit through |
| `endyr_full` | the unpeeled terminal year, where the retrospective forecast ends |
| `projyr` | the end of the projection |

`incl_proj = TRUE` plots the forecast and projection years. Take the forecast years as `endyr_peel + seq_len(endyr_full - endyr_peel)` — empty for the unpeeled model, where `(endyr_peel + 1):endyr_full` would count down instead.

The `retrospective()` function returns a list with three elements:

| Element | Description |
|---|---|
| `Rceattle_list` | List of fitted models, ordered from most-peeled to full run, each named for its own terminal year (`Year_2013`, …, `Year_2017`) |
| `mohns` | Retrospective bias. Long: `Object` (the quantity — `"biomass"`, `"ssb"`, `"R"`, `"F_spp"`), `Forecast year` (a horizon in steps ahead of the peel's terminal year; 0 = terminal-year bias), `N` (number of peels), `species`, `rho` (mean signed relative error, Mohn's rho) |
| `mase` | Forecast skill. Same leading columns as `mohns` — `Object`, `Forecast year`, `N`, `species` — then `mase`, its two halves `mae_forecast` / `mae_naive`, `peels_used`, and `forecast_rec`. Always computed from the peels already fitted, so it costs no extra fits |

**The two tables share their key columns, so they merge directly:**

```{r, eval = FALSE}
merge(model_1_retro$mohns, model_1_retro$mase,
      by = c("Object", "Forecast year", "species"))
```

Merge on those three, **not** on `N` as well. `N` means the same thing in both
— the number of observations the statistic was computed from — but the two are
computed from different peels, since MASE additionally needs a usable naive
baseline. They agree for a well-behaved species and diverge precisely where a
term was unusable, so joining on `N` would discard the rows you most want to
see. Merging on the keys brings both `N` columns through, and a disagreement
between them is a finding.

`Forecast year` is a horizon in both — steps ahead of the peel's terminal year, not a calendar year. They overlap only above horizon 0: `mohns` carries horizon 0, the terminal-year bias the ±0.2 rule is about, and `mase` does not, because a zero-step forecast is the terminal estimate compared with itself. `mohns` also scores `F_spp`, which `mase` does not.

Read the two together, because they do not measure the same thing. **Mohn's rho is signed, so it is a measure of BIAS**: a model whose peels miss high as often as they miss low averages to a rho near zero and can still forecast badly. **MASE is a scaled absolute error, so it is a measure of SKILL** -- `< 1` beats the naive baseline, `> 1` is worse than it. A model can pass one and fail the other, and that combination is informative rather than contradictory.

```{r, eval = FALSE}
model_1_retro$mase
```

Three things to read off the table rather than assume:

* **`forecast_rec` is a column**, because the table scores whichever projection rule the peels were forecast under. `retrospective()`'s default is `"mean"` -- mean recruitment, *not* the fitted model's own process -- so a DSEM or stock-recruit model scored at the default is being told how well mean recruitment forecasts. `print()` on the object says so. Use `hindcast_skill()` to score a fit's own process; the next section does exactly that.
* **`N` varies down the table, so read it per row.** A horizon of `h` years needs a peel at least `h` deep, so the count falls as the horizon grows and the deepest row can rest on a single peel. `peels_used` names which depths contributed. Join on `Forecast year` rather than comparing positionally. (It can differ by species in both tables: a species whose relative error or forecast is not finite — a zero base, or input numbers-at-age — contributes no observation at that horizon.)
* **A species with fixed numbers-at-age (`estDynamics > 0`) is `NA`, not 0.** Its peel reproduces the parent exactly, so there is no forecast error to scale, and a 0 there would read as perfect skill.

The `nyrs_forecast` argument (default 3) also evaluates rho for years projected beyond each terminal peel, so a single call quantifies forecast skill alongside retrospective bias.

A peel that did not converge is dropped and a message reports how many, so `Rceattle_list` can be shorter than one plus the number of peels asked for. Index it by name rather than position — with a peel missing, `Rceattle_list[[3]]` is not the 3-year peel. If no peel converges, `mohns` comes back with no rows and the call warns.

`peels` can also be a **vector of peel depths** rather than a count, so a run can skip the shallow peels:

```{r, eval = FALSE}
retro <- retrospective(model_1, peels = 2:10)   # nine peels, no 1-year peel
retro$peel_depths                               # the depths asked for
```

Each element is a number of years removed from `endyr`, not a year, and depths are sorted and de-duplicated. Mohn's rho is averaged over whichever peels are fitted, so a rho from `peels = 2:10` is not comparable to one from `peels = 10` — subset deliberately, and say which depths a reported rho covers.

## Forecast skill across peels

`hindcast_skill()` scores how well each peel's projection recovers the full model's estimate over the years the peel did not see, given the catch actually taken. Where Mohn's rho measures estimation consistency, this measures forecast skill, so it is the natural way to compare recruitment-projection assumptions: mean recruitment against the stock-recruit curve against a DSEM.

```{r}
skill <- hindcast_skill(model_1, peels = 5)
skill$mase
```

Skill is a mean absolute scaled error (MASE), following Kell et al. (2016). The baseline is the peel's own terminal estimate carried forward — "assume nothing changes" — so `MASE < 1` beats persistence and `MASE > 1` is worse than it, which usually means the recruitment assumption is pulling the projection the wrong way.

Two things to know when reading it. `reference = "observed"` scores the peel's predicted survey index against the index actually observed, which is classic hindcast cross-validation; `reference = "model"` (the default) scores SSB, biomass or recruitment against the full model. Note the `"observed"` MASE is **not** on the same scale as published values: Kell et al. (2021) average the denominator over a wider window than the numerator, and this averages both over the peels fitted, which moves the denominator by up to 2.6× on the bundled data — enough to flip a `MASE < 1` verdict. Comparisons between two models scored the same way here are unaffected. And read `N` alongside each MASE, per row rather than down the column: the scoring averages across peels at a fixed steps-ahead. On an annual series a horizon `h` needs a peel at least `h` deep, so `N` falls as the horizon grows and the deepest row rests on one peel. That is not guaranteed — under `reference = "observed"` an irregular survey is scored only where it has observations both at the peel's terminal year and after it, so some horizons carry no rows at all and `$mase` is ragged. Join on `Forecast year` rather than comparing positionally.

`hindcast_skill()` defaults to `forecast_rec = "model"`, which hands the peeled years to the model's own projection rule; `retrospective()` defaults to `"mean"`, the convention Mohn's rho has always been computed under. Without that difference every model forecasts identically and the comparison says nothing.

**`forecast_rec = "model"` needs `build_srr(proj_mean_rec = FALSE)`.** `proj_mean_rec` takes precedence, and it defaults to `TRUE`, so a model fitted without naming it projects the peeled years at mean recruitment whatever process it carries — and `hindcast_skill()` then cannot tell projection methods apart, which is the one thing it exists to do. Measured on the GOA arrowtooth DSEM: `forecast_rec = "mean"` and `"model"` returned identical Mohn's rho to every digit. `retrospective()` warns when it happens, once per call and before any peel is fitted.

### Does a DSEM's correlation structure improve the projection?

This is the comparison `hindcast_skill()` exists for: a DSEM's lagged and covariate paths carry information into the peeled years that mean recruitment throws away, and the MASE says whether that information helps.

**Do it from one fit, not two.** Fit the DSEM with `proj_mean_rec = FALSE`, then score it both ways. The two `forecast_rec` settings differ only in the *forecast* refit — the peeled hindcast is the same fit either way — so this isolates the projection rule and nothing else:

```{r, eval = FALSE}
dsem_fit <- fit_mod(data_list = BS2017SS, dsem = build_DSEM(sem = my_sem),
                    recFun = build_srr(srr_fun = 0, proj_mean_rec = FALSE),
                    random_rec = TRUE, estimateMode = 1)

retro_mean  <- retrospective(dsem_fit, peels = 3:10, forecast_rec = "mean")
retro_dsem  <- retrospective(dsem_fit, peels = 3:10, forecast_rec = "model")

# Score RECRUITMENT -- see below for why not SSB
hindcast_skill(dsem_fit, retro = retro_mean, quantity = "R")$mase
hindcast_skill(dsem_fit, retro = retro_dsem, quantity = "R")$mase
```

`"mean"` is the mean-recruitment projection and `"model"` the DSEM's; the lower MASE wins. `peels = 3:10` is about cost rather than noise -- more peels add terms to the mean inside each MASE rather than adding rows, so dropping the two shallowest pays for two fewer model fits and costs a little precision at the short horizons.

**Score recruitment, not SSB, over a short horizon.** The recruits the two rules disagree about are not mature yet, so SSB hardly responds and the MASE cannot see the projection rule. Measured on a BS2017SS DSEM (lag-1 on the recruitment deviations plus a temperature covariate) at a 3-year peel: the two rules differed by **46.8%** in forecast recruitment — `"mean"` flat at 35.2 million, the SEM decaying 35.2 → 25.8 → 18.7 million as the lagged state relaxed — and by **0.03%** in forecast SSB. The SSB MASE agreed to four significant figures for two of the three species. So `quantity = "R"` is the sharp test of a recruitment projection; SSB becomes informative only once the horizon reaches the age at maturity, which for these stocks means deeper peels than three years.

Use the same run to confirm the comparison is clean: the peeled *hindcast* is identical between the two settings (measured: `max |difference|` in hindcast recruitment exactly 0), so any difference you see is the projection rule and nothing else.

**Don't do it by refitting with `proj_mean_rec = TRUE` and comparing the two fits.** That changes the projection rule *and* the fit, so a MASE difference mixes the two and neither number isolates the correlation structure. `proj_mean_rec` is a property of the configuration you are fitting; `forecast_rec` is the knob for asking what a *given* fit would have forecast, which is the question here.

One setting removes the choice entirely: `build_DSEM(estimate_projection = TRUE)` makes the SEM span the projection, and `fit_mod()` then sets `proj_mean_rec = FALSE` for you and says so. Such a fit reads `FALSE` here and is scored on its own process without further arrangement.

## Jitter testing

Jitter testing re-fits the model from many randomly perturbed starting values and checks whether the optimizer keeps returning the same minimum negative log-likelihood (NLL). A spread of NLL values across jitters points to multiple local minima, so the results should be treated with caution.

```{r}
jitters <- jitter(model_1, njitter = 10, phase = TRUE)  # default njitter = 50

# Histogram of NLL differences relative to the best run
hist(log(jitters$nll - min(jitters$nll)),
     main = "Jitter NLL spread (log scale)",
     xlab = "log(NLL - min NLL)")

# Overlay biomass trajectories — tight overlap indicates a stable optimum
plot_biomass(jitters$Rceattle_list)

# Number of runs that converged
length(jitters$Rceattle_list)
```

Non-converging runs are dropped from `Rceattle_list` and reported in a message, so if `length(jitters$Rceattle_list)` is well below `njitter`, convergence is fragile. Read that count as part of the result, not as a nuisance: the dropped starts are the ones that wandered off, so a jitter test that returns only its successes understates the spread.

A run is dropped when its maximum absolute gradient exceeds 1 — the same threshold `convergence_diagnostics()` calls `FAIL` — or when an `sdreport` that was requested did not come back.

`retrospective()` and `jitter()` both fit peels / starts in parallel by default. Pass `cores = 1` to force sequential execution; otherwise the call uses a PSOCK cluster sized at `parallel::detectCores() - 6` (capped at 2 under `R CMD check`).

## Self-test (simulation–estimation)

`self_test()` simulates `nsim` datasets from a fitted model — holding its estimated parameters fixed — and re-fits the model to each. If the refit estimates cluster around the values used to simulate them, the model is at least self-consistent: it can recover its own estimates from data it generated. Persistent bias or a wide spread in a quantity means the data are not informative about it.

```{r}
sims <- self_test(model_1, nsim = 10)            # default nsim = 50

# Number of simulations that converged (non-converged runs are dropped)
length(sims)

# Overlay biomass / SSB trajectories across simulations — the original fit's
# trajectory should sit inside the spread of the refits.
plot_biomass(c(list(model_1), sims), model_names = c("fit", names(sims)))
plot_ssb(c(list(model_1), sims),     model_names = c("fit", names(sims)))
```

Each simulation uses `seed + i` as its RNG seed, so results are reproducible whether or not `cores > 1`. Set `simulate = FALSE` to refit against the model's *expected* values (no observation error) rather than draws from the observation likelihood — useful for confirming that the estimator returns the generating parameters in the noise-free limit. `cores` behaves the same way as in `retrospective()` and `jitter()`.

Every refit starts from the fitted model's *starting* values, not its estimates, so it covers the same ground the original fit did. `phase` therefore defaults to whatever the input model was fit with — pass `phase = TRUE` explicitly if you fit the model some other way and the refits are being dropped, since a model that needed phasing once needs it for every simulated data set too. `start = "estimated"` starts from the fitted estimates instead: much faster, but the optimizer then begins inside the basin containing the generating values and recovery is close to guaranteed by construction, so read it as optimistic about recovery rather than simply less powerful.

Three things to keep in mind when reading the result. Non-converged refits are dropped and reported in a message, and they are dropped *because* they behaved badly — so the returned replicates are a favourable subset, and any spread or bias computed over them (including `compare_sim()`'s CV) understates the truth. By default only the observations are redrawn: every replicate shares the operating model's single realization of recruitment and mortality, so the spread across replicates is observation error alone — a lower bound on estimation uncertainty in SSB and recruitment, not comparable to the model's own uncertainty bands.

Third, watch for a truncation warning on a natural-scale index. A fleet with `Index_distribution = "Normal"`, `"MVN"` or `"MVNORM"` is drawn on the natural scale with an absolute sd, so a draw can come out negative — impossible for an index, and rejected by `data_check()`. Those draws are redrawn, which samples the fleet from a normal *truncated at zero* rather than the untruncated normal the likelihood maximises. Occasional truncation is harmless; a lot of it is not. An acoustic index observed at 0.70 with an absolute sd of 0.80 rejects a fifth of its draws and the truncated mean sits 39% above the untruncated one, so the self-test is measuring recovery against a data-generating process the estimator never assumed. `sim_mod()` warns when any row puts more than 2% of its probability at or below zero — `Phi(-mu/sd)`, read off the fit rather than counted from the draw, so it does not vary between replicates.

The fix for a `Normal` fleet is `Index_distribution = "TruncatedNormal"`, which fits the truncated density — renormalized by `log Φ(µ/σ)` over the `(0, ∞)` the data can actually occupy — and draws from it by inverse CDF. Draw and likelihood then agree exactly and no draw is ever rejected. `"Normal"` is kept untruncated because it reproduces the AMAK `avo_like` / `cpue_like` term for term, which is what an ADMB bridge compares against; use it there and `"TruncatedNormal"` elsewhere. `MVN`/`MVNORM` have no such option: a multivariate normal truncated to the positive orthant has no closed-form sampler, so rejection is the only route and the gap stays. Either way the underlying issue is the same — an absolute sd that large relative to the index means the natural-scale normal was a poor observation model to begin with.

`"TruncatedNormal"` OSA residuals need one thing said about them. The truncation constant is a function of the prediction, not of the observation, so it drops out of any method that reads the curvature of the density in the observation — including `oneStepPredict()`'s default `oneStepGaussianOffMode`, which would return the untruncated `(obs - µ)/σ`. `osa_residuals()` therefore residualizes this family in its own call, integrating the density over `(0, ∞)`, so the residual is `qnorm(F(x))` for the truncated `F`. The correction is the size of the truncated mass: on a fleet predicting 100 with an absolute sd of 150, an observation exactly at the prediction moves from a residual of 0 to −0.44. You do not need to pass anything for this; `method` is overridden for those rows only.

### Redrawing process error

`process` widens the question. The default (`FALSE`) asks whether the estimator recovers its own parameters from new observations; naming a process redraws that process too, and asks whether it recovers something it was not shown.

```{r}
# Recruitment deviations redrawn as well as the observations
sims <- self_test(model_1, nsim = 10, process = "recruitment")

# "dynamics" covers recruitment, M and growth; "observation" covers
# catchability and selectivity; TRUE covers everything.
sims <- self_test(model_1, nsim = 10, process = "dynamics")
```

Recruitment covers the initial age structure as well as the hindcast deviations, since the initial numbers-at-age are recruitment from before `styr`. Natural mortality covers the `M1_re` random effects, and growth, catchability and selectivity are redrawn wherever they are written as a random linkage (see `vignette("environmental-linkages-and-priors")`). A process the model gives no distribution to — asking for M when `M1_re = 0`, say — is left alone and reported in a warning rather than silently returned unchanged.

Recruitment has one such case worth knowing about. Under `srr_fun = 0` with `srr_pred_fun > 0` — the AMAK/Ianelli configuration, where recruitment is estimated annually and the stock-recruit curve is fitted as a *penalty* on those same deviations — the deviations are scored by two densities at once.

Two terms on one quantity do not compose into a distribution to draw from. Drawing at `R_sd` from the first alone ignores the second, and the second is not a density on the deviations alone either: it couples them to the stock-recruit parameters, which are estimated too. Either way the self-test would measure recovery against a process the model does not assume, so the deviations are not redrawn and `sim_mod()` says why. A random linkage on a recruitment parameter is a separate latent and is still drawn. Set `srr_fun` to the stock-recruit function itself to redraw the deviations.

The fitted deviations are no longer what generated the data, so the values that *did* come back attached to each replicate, each with a mask of the cells the draw touched:

```{r}
truth <- attr(sims, "process_sim")[["Sim_1"]]
rec   <- truth$rec_dev[truth$rec_dev_drawn]
```

Compare estimates against those, not against the operating model. The mask matters: the process draws cover the hindcast only, and `beta_linkage_re` spans every random linkage whether or not its process was asked for, so a correlation taken over the whole array scores fitted values as though they had been simulated. For the same reason `compare_sim()`, which compares against the operating model, is not valid for `process`-drawn replicates.

One caveat before reading a redrawn M as validation. On `BS2017SS` with a year-varying M random effect, refits recover the simulated deviations (correlation around 0.5, positive in every replicate) but recover their standard deviation poorly — roughly 70% low, and collapsing to zero in a third of replicates. That is the usual behaviour of a variance component estimated from little information, not a fault in the simulation, and a year-varying M is weakly informed because it trades off against selectivity and catchability. Read a self-test that redraws M as a test of deviation recovery rather than of variance recovery.

When the returned list is shorter than `nsim`, `debug = TRUE` returns *every* simulation so the failures can be read directly rather than inferred:

```{r}
sims <- self_test(model_1, nsim = 10, debug = TRUE)

failed <- names(which(!attr(sims, "converged")))
sims[[failed[1]]]$convergence          # which check failed, and on which parameter
```

## Likelihood profile

`profile()` re-fits the model across a grid of fixed values for a chosen parameter and returns the NLL surface. A flat profile means the data carry little information about the parameter; a minimum away from the MLE means the fit has not truly settled there. It supports the recruitment standard deviation, the stock–recruit parameters, and natural mortality.

Specify the parameter with a natural-scale alias: `"sigmaR"` (recruitment
standard deviation), `"M1"` (natural mortality), or the stock–recruit parameters
`"R0"`, `"alpha"`, and `"beta"`. Aliases take natural-scale values directly.
`slots` gives the cell(s) to profile — usually just the species index — and
`values` gives the grid of values for each. Supplying more than one slot
profiles over the full grid of combinations.

```{r}
# 1-D profile: sigmaR for species 1 (natural-scale alias)
prof_sigmaR <- profile(
  fitted = model_1,
  param    = "sigmaR",
  slots    = list(1),
  values   = list(seq(0.1, 1.5, by = 0.1))
)

plot(prof_sigmaR)   # components, with the total overlaid; see below

# 1-D profile: SRR alpha for species 1
# (alias fills in the rec_pars column; slot is just the species index)
prof_alpha <- profile(
  fitted = model_1,
  param    = "alpha",
  slots    = list(1),
  values   = list(seq(2, 80, length.out = 20))
)

# 2-D cross-profile: M1 across sex for species 1, age 1
prof_M_sex <- profile(
  fitted = model_1,
  param    = "M1",
  slots    = list(c(1, 1, 1), c(1, 2, 1)),  # males and females
  values   = list(seq(0.10, 0.40, by = 0.05),
                  seq(0.10, 0.40, by = 0.05))
)
```

### Moving a set of cells together

`slots` cross, so *k* slots give a *k*-dimensional profile. That is wrong for a
parameter supplied as a vector — an empirical age-based M profiled over 13
values across ten ages would be `13^10` fits. `joint` puts every named cell on
**one** grid instead: `"multiply"` scales each cell's current value, preserving
the shape of the schedule (`1` is the fitted model), and `"add"` shifts it,
preserving the increments (`0` is the fitted model).

```{r}
sp <- 1
prof_M_all <- profile(
  fitted = model_1,
  param  = "M1",
  slots  = lapply(seq_len(model_1$data_list$nages[sp]),
                  function(a) c(sp, 1, a)),   # every age bin of species 1
  values = list(seq(0.7, 1.3, by = 0.05)),
  joint  = "multiply"
)

plot_profile(prof_M_all)
```

### Catchability

`param = "q"` profiles a fleet's base catchability, and `slots` takes a fleet
name. Fleets sharing a `Catchability_index` share one q, so naming any member
profiles the whole group — the others are added with a message.

```{r}
prof_q <- profile(
  fitted = model_1,
  param  = "q",
  slots  = list("BT_Pollock"),
  values = list(seq(0.5, 2.0, by = 0.1))
)
```

`profile()` returns `Rceattle_list` (one fit per grid row, `NULL` where the fit failed), `grid` (the user-scale value grid — a multiplier or offset under `joint`), `nll` (joint NLL aligned with `grid`, `NA` for non-converged fits), `alias` (the natural-scale name you profiled under, if you used one), `joint`, and echoes of `param` and `slots`. To cross-profile across multiple species, supply one slot per species (e.g. `slots = list(1, 2, 3)` with `param = "sigmaR"`); to cross-profile M1 across sex, supply one slot per sex as in the third example above. `cores` behaves the same way as in the other diagnostic functions.

### Where the data disagree

The total says how well the data determine the parameter, not whether the data
sources agree about it — a smooth quadratic total can be two surveys pulling in
opposite directions. `plot_profile()` draws the components separately, in the
layout `r4ss::SSplotProfile()` uses.

```{r}
prof_M <- profile(
  fitted = model_1,
  param  = "M1",
  slots  = list(c(1, 1, 1)),
  values = list(seq(0.20, 0.45, by = 0.025))
)

plot_profile(prof_M, xlab = "M at age 1")
```

Read it by **where each curve bottoms out**, not by how deep it is: every series
is re-zeroed at its own minimum and a point marks it, so the spread of those
points along the x axis is the disagreement. The heavy black line is the total.

When one component dwarfs the rest the others flatten onto the axis and where
they prefer the parameter cannot be read. `relative = "scaled"` puts every curve
on 0 to 1 so the minima can be compared directly; it discards magnitude, so
raise `minfraction` alongside it.

`profile_components()` returns the same numbers, one row per grid point per
component, each labelled with the fleet or species it belongs to.

```{r}
comps <- profile_components(prof_M)
head(comps)
```

Under `random_rec = TRUE` the total is the Laplace-approximated marginal
likelihood while the components are the inner joint negative log-likelihood, so
they will not sum; compare the shapes, not the sums. See `?plot_profile` for
`relative`, `minfraction`, model overlays, and what `weighted = FALSE` does and
does not include.

## Diagnostics on a DSEM fit

A model fitted with `fit_mod(dsem = build_DSEM(...))` makes the recruitment deviations the latent states of a Gaussian Markov random field rather than independent deviations (see `vignette("dsem")`). Every diagnostic on this page runs on one, but three of them mean something slightly different, and one is not available.

| | On a DSEM fit |
|---|---|
| `osa_residuals()`, `osa_diagnostics()` | unchanged — same observations, same residuals |
| `retrospective()`, `hindcast_skill()` | work; peeled-year latent states are integrated out rather than pinned |
| `jitter()` | works; perturbs the SEM path coefficients and the latent recruitment deviations, not the covariate columns |
| `self_test()` | works, with or without `process = TRUE` |
| `process_residuals()` | works; recruitment residuals are whitened by the process precision |
| `profile()` | works; `param = "sigmaR"` resolves to the SEM's variance path |
| `summary()`, `AIC()`, `convergence_diagnostics()` | work; `summary()` also prints the SEM path coefficients |
| `sample_rec()` | needs `build_DSEM(estimate_projection = TRUE)` |
| `run_mse()` | not yet supported |

**OSA residuals cover the assessment data, not the environmental data.** The one-step-ahead residuals are computed for catch, survey indices, compositions, conditional age-at-length and diet — exactly as without a DSEM, and the count is the same. The SEM's covariate series is not among them. Under `family = "fixed"`, which is the usual setting, that is not a gap: the covariates carry no measurement density at all, so there is no residual to compute. Under any other family there is one, and it is not yet exposed.

**Recruitment process residuals are whitened, not divided.** A DSEM's deviations are correlated, so there is no per-year SD to standardize by. `process_residuals(process = "recruitment")` instead conditions on the states the model was given, takes one draw from the joint posterior, and multiplies by the Cholesky factor of the process precision. Under a correctly specified SEM the result is approximately iid N(0, 1), which is the same claim the non-DSEM path makes.

```{r}
pr <- process_residuals(dsem_model, process = "recruitment")
osa_diagnostics(pr)
plot(pr)
```

**`profile()` finds the recruitment SD where it actually lives.** Under a DSEM, `R_log_sd` is mapped out and the SD comes from the SEM's two-headed self-loop, so `param = "sigmaR"` resolves to `dsem_beta_z[rec_sd_idx[sp]]`. `slots` still indexes by species. The values are on the natural scale — `R_sd` is the absolute value of a Cholesky factor, whose sign is not identified.

**`self_test(process = TRUE)` re-realizes the SEM.** The latent field is drawn from the same precision that scored the fit, so an AR1's autocorrelation, a covariate's effect and any cross-species path are all preserved across replicates — which is what makes the test able to say whether the SEM structure is identified. Every estimated cell is drawn, conditional on the ones the map pins; the deviations behind each replicate come back in `attr(sims, "process_sim")`, which under a DSEM carries the latent field itself (`dsem_x_tj`) alongside `rec_dev`. Score recovery against those, not against the operating model — its deviations are no longer what generated the data, and `compare_sim()` warns if you try.

**A covariate is drawn according to how it was declared.** Under `family = "fixed"` it is data: no measurement density, pinned in the map, identical in every replicate. Under any other family it is a latent state observed with error, so both halves are simulated — the state with the rest of the field, and the observation from the family that scores it, written back into `env_data` so the refit is fitted to it. That is not gated on `process`: an observation is redrawn whenever the observations are. A self test of a SEM whose covariate carries measurement error will therefore look less optimistic than one whose covariate is `fixed`, and it should.

**Recruitment scored twice is not redrawn, DSEM or not.** `srr_fun = 0` with `srr_pred_fun > 0` — the AMAK/Ianelli configuration — adds a stock-recruit penalty on the same deviations, so there is no single distribution to draw from and `sim_mod()` says so rather than picking one. Under `cores > 1` those warnings come back from `self_test()` once, collected from the replicates; if nothing at all was redrawn, `self_test()` says that too rather than leaving an absent `process_sim` to be interpreted.

## Comparing runs

Overlaying sensitivity runs is itself a useful diagnostic — large divergences in biomass or mortality deserve scrutiny.

```{r}
mod_list  <- list(model_1, model_1_retro$Rceattle_list$Year_2019)
mod_names <- 1:length(mod_list)

plot_biomass(Rceattle = mod_list, model_names = mod_names)
plot_recruitment(Rceattle = mod_list, model_names = mod_names, add_ci = TRUE)
plot_depletionSSB(Rceattle = mod_list, model_names = mod_names)
```


## Model average

To average estimates across model variants, see `?model_average`.

```{r}
mod_avg <- model_average(list(model_1, model_1), weights = c(1,1))
plot_biomass(mod_avg)
```
