---
title: "3. Model diagnostics"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{3. Model diagnostics}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("RCEATTLE_EVAL_VIGNETTES"), "true"),
  fig.width = 7,
  fig.height = 5
)
```

Rceattle provides a full set of model diagnostics:

- **convergence diagnostics**: attached automatically to every fit
- **S3 methods**: `residuals()`, `logLik()`, and the like
- **fit plots**: compare predicted to observed data
- **retrospective analysis**: peel off recent years to detect systematic bias
- **jitter testing**: perturb starting values to check the optimizer found a global minimum
- **self-testing** (simulation–estimation), check the model recovers its own estimates from simulated data
- **likelihood profiling**: check how informative the data are about a given parameter

**They all report the same way.** Printing any of them opens with one line (the
object, an overall status, and a verdict) before the detail:

```
<Rceattle convergence>  status: OK
  all checks OK

<Rceattle OSA diagnostics>  status: WARN
  7 of 7 source(s) outside a null interval; overall SDNR 1.82 (0.91-1.09)
          source  fleet  n     sdnr   null
  [WARN]  index      4     36  2.52   0.77-1.23
  ...

<Rceattle retrospective>  status: WARN
  4 peel(s); 3 of 12 terminal Mohn's rho outside +/-0.2
          quantity  N     Pollock  Cod      Arrowtooth flounder
  [WARN]  R            3  0.511    0.283    -0.221

<Rceattle jitter>  status: OK
  4 of 4 start(s) reached the best optimum (within 0.01)

<Rceattle self-test>  status: NOTE
  47 of 50 simulation(s) converged
          convergence  n
  [OK  ]  OK           41
  [NOTE]  NOTE          6
  observations redrawn, processes held at their fitted values: the spread is a lower bound

<Rceattle profile>  status: WARN
  sigmaR over 29 grid point(s); all converged
  minimum -log L : 1274.8831  at slot_1 = 1.5
  within 1.92 of the minimum : 1.3 to >=1.5
  the minimum is at a grid edge (slot_1), so the grid does not bracket it, widen `values`
```

The severity words are the same four everywhere, `OK`, `NOTE`, `WARN`, `FAIL`,
and the overall status is the worst present. The **return values are unchanged**:
each object is still the data frame or list it always was, so `retro$mohns`,
`jit$nll`, `diag$sdnr`, `prof$grid$slot_1` and `sims[["Sim_1"]]` index exactly as
before, and `c(sims, list(fit))` still gives a plain list of fits to plot. Only
the printing has an opinion.

Four of these hold a threshold you can set. `print(retro, band = )` is Mohn's
rho threshold; the default `0.2` is the rule of thumb below, and Hurtado-Ferro
et al. (2015) give asymmetric, life-history-dependent alternatives (−0.15 to
0.20 long-lived, −0.22 to 0.30 short-lived), which is why it is an argument.
`print(jit, tol = )` is how close an objective must be to count as having
reached the same optimum. `print(sims, rate = )` is the fraction of simulations
that must converge before a self-test counts as clean. `print(prof, cutoff = )`
bounds the reported interval; the default 1.92 is the 95% profile-likelihood
cutoff for one parameter, half the 0.95 quantile of a chi-square on one degree
of freedom.

Two of them report something the numbers alone cannot. A self-test drops its
non-converged runs before returning, so the length of what comes back is not the
number attempted; the header gives the rate against `nsim`, and the last line
says what generated the replicates, since with processes held fixed the spread
holds observation error only. A profile whose minimum sits at the first or
last grid value has run out of grid rather than found the optimum; it plots as
an ordinary curve, so the print says so and asks for a wider `values`.

## Setup and plotting data

```{r}
library(Rceattle)

# Fit a single-species model for the 2022 Northern Rockfish assessment.
data("NorthernRockfish2022")
nrdata <- NorthernRockfish2022
nrdata$fleet_control$Proj_F_proportion <- 1
plot_data(nrdata)
```

Fit the model.
```{r}
model_1 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    updateM1 = TRUE,
    M1_model = 1,
    M1_use_prior = TRUE,
    M_prior = 0.06,
    M_prior_sd = 0.05),
  fit_control = fit_control(phase = TRUE, verbose = 0)
)
summary(model_1)
```

## S3 Methods

```{r}
summary(model_1)              # same compact summary
coef(model_1)                 # estimated fixed-effect parameters
logLik(model_1)               # logLik with df attribute (AIC works)
AIC(model_1)
vcov(model_1)                 # fixed-effect covariance from sdreport
residuals(model_1)                    # response residuals, all data sources
residuals(model_1, type = "pearson")  # Pearson residuals (glm-style `type`)
as.data.frame(model_1)
```


## Convergence diagnostics

After optimization, `fit_mod()` runs a battery of convergence checks and attaches
the result as `model$convergence`. Each check is one record with a `severity`
(`"OK"`, `"NOTE"`, `"WARN"`, or `"FAIL"`); the object's `status` is the worst
severity present. Non-OK checks are reported via `message()` during the fit, and
`print(model)` shows the overall status. A non-converged model is never turned
into an error: the fit is always returned with its diagnostics attached, so you
decide whether to trust it.

```{r}
model_1$convergence            # overall status + any non-OK checks
print(model_1$convergence, all = TRUE)   # show every check, including OK ones
```

Re-run the battery on any fit with `convergence_diagnostics()`:

```{r}
convergence_diagnostics(model_1)
```

The checks cover:

- **`max_gradient`**: maximum absolute marginal gradient and the parameter
  holding it (`WARN` > 1e-3, `FAIL` > 1). When an `sdreport` of the same
  parameters is available, the message also says how far the estimates are from
  the optimum in standard errors. A quadratic approximation puts the optimum a
  Newton step `cov.fixed %*% gradient` away, and each element is divided by its
  own standard error. That size is comparable across log, logit and
  natural-scale parameters, which the raw gradient is not: 0.02 means finishing
  the optimization would move no estimate by more than 2% of its standard error.
  The value is in `$data$newton_step_se`. It is reported only; the severity is
  still read on the gradient.
- **`pdHess`**: whether the Hessian is positive definite (`FAIL` if not).
- **`sdreport_failed`**: `FAIL` when an `sdreport` was requested but the
  Hessian could not be inverted.
- **`hessian_not_run`**: `NOTE` when the fit was made with `getsd = FALSE`, so
  `pdHess`, `hessian_conditioning`, `sdreport_failed` and `estimability` had
  nothing to read and never ran. It separates "every check passed" from "the
  strongest checks were skipped"; refit with `getsd = TRUE` before reading the
  status as convergence. Each refitting diagnostic resolves its own `getsd`, so
  `run_mse()`'s refits carry the note and a reweighted default fit does not.
- **`zero_n_penalty`**: the "Zero n-at-age penalty" likelihood row is non-zero
  for a species: its numbers-at-age, Ricker intercept or (under an identity-link
  recruitment linkage) recruitment sat on the 0.001 floor, so the fit is of a
  floored model rather than the one specified. `WARN` when the root-sum-square
  excursion is under 1 (a thousand fish), `FAIL` above.
- **`hessian_conditioning`**: how much better the data pin down some parameter
  combinations than others. Inverting the Hessian gives the covariance of the
  estimates; standardising it by the standard errors gives their correlation,
  whose eigenvalues are variances along directions in parameter space. The
  condition number is the largest over the smallest. This is the multi-parameter
  version of a 0.999 pairwise correlation, and catches confounding spread over
  many parameters at once. On a fishery it usually means F traded against
  selectivity: scale F up, shift selectivity down, and predicted catch-at-age
  barely moves.

  The message leads with the square root, a ratio of standard errors. Because
  the matrix is scaled by those errors, the ratio is measured against each
  parameter's own: 13,634 means the least-determined combination is 13,634 times
  more uncertain than the best-determined one, in those units. `WARN` above 1e6,
  `FAIL` above 1e10.

  Read on the **correlation** and not the covariance, because the covariance
  condition number is not scale-invariant: `log_F` sits near -2 and `sel_inf`
  near 10, so rescaling a parameter would move it without changing the model.
  Standardising costs sensitivity (the same three models read 4.6e4, 1.9e5 and
  2.5e9 on the covariance against 2.0e4, 1.6e4 and 1.9e8 here) so a model that
  is merely badly scaled no longer trips the check. The covariance number is
  still reported, as `covariance_condition_number`, for the numerical cost of
  inverting the Hessian; `pdHess` and `sdreport_failed` are what catch an
  inversion that actually fails. The thresholds were held at 1e6/1e10 on three
  models: `BS2017SS` sits at 2.0e4, roughly 1.7 orders below `WARN`.

  The parameters holding the top 90% of each block's loading are then named by
  fleet, bin and year, with a count: a ridge over ten years and one over a single
  year read differently. Complements `TMBhelper::check_estimability`, which gives
  a per-parameter verdict, with a continuous severity and a direction.
- **`parameters_on_bounds`**: parameters that hit a configured `build_bounds()`
  limit (often unidentified or mis-scaled).
- **`variance_collapse`**: a deviation standard deviation estimated to zero
  (`WARN` below 1e-3), meaning the process it governs is not varying: a model
  configured as time-varying has fitted something time-invariant. Covers the
  recruitment, selectivity, catchability, M1 and linkage random-effect standard
  deviations. `WARN` rather than `FAIL` because a variance at the boundary is a
  well-posed optimum (a verdict about the model, not the optimizer) and it can
  occur with a perfectly clean gradient, which is the case nothing else catches.
- **`stock_recruit`**: whether the fitted curve is one the data informed. In a
  single-species model it is read against the replacement line through
  steepness. Under predation steepness does not exist (it needs spawning biomass
  per recruit), so the curve is read over the observed SSB range instead: `WARN`
  when it is flat there (recruitment independent of SSB), linear (no density
  dependence), when a Ricker peaks below the lowest observed SSB, when log alpha
  or log beta sits at its +/-30 bound, or when a log-scale standard error exceeds
  10. A curve held at its inputs is a `NOTE`.
- **`phasing`**: phases that ended with a high gradient, localizing which
  parameter block is hard to fit.
- **`estimability`**: surfaces `TMBhelper::check_estimability` when it ran.

### Naming the parameter, not just the block

TMB names every element of `obj$par` after its parameter block, so on their own
`max_gradient`, `estimability`, `parameters_on_bounds` and
`hessian_conditioning` could say `sel_coff_dev` was flagged but not which of its
392 elements. Each now reports the fleet, sex, age, bin and year instead, from
[`parameter_index()`], and follows the block name with the quantity it
estimates, from [`parameter_dictionary()`]:

```
  [WARN] max_gradient     Maximum absolute marginal gradient = 0.00222 on log_M1 (M1): Pollock, age 1.
  [WARN] hessian_conditioning Condition number = 2.4e+06; loads on: log_F (F): 67% + rec_dev (recruitment deviations): 26%.
  log_F (F)                        Hake_fishery, 38 years in 1980-2021  (38 of 44 parameters; 67% of the direction)
  rec_dev (recruitment deviations) 24 years in 1981-2013  (24 of 44 parameters; 26% of the direction)
```

A scattered set of years or ages is counted against its span ("38 years in
1980-2021"); a consecutive run prints as a range. The count in parentheses is
how many of the block's parameters carry 90% of its loading.

The checks do not all count positions in the same vector. `max_gradient`,
`estimability` and `parameters_on_bounds` are captured from the hindcast
optimization; `hessian_conditioning` reads the `sdreport`, which under an
estimating HCR belongs to the projection fit and holds only `log_Ftarget` /
`log_Flimit`. Each check therefore verifies that the index describes the vector
it is about to label, and prints block names alone when it does not: a
coordinate naming the wrong parameter is worse than no coordinate.

`parameter_index()` itself describes `fit$obj`, which under any HCR but
`"NoFishing"` is that projection object. Pass a fit run with
`estimateMode = "Hindcast"` to browse the hindcast parameters directly.

## Fit plots

### Composition data

`plot_comp()` draws composition Pearson-residual bubbles and overlays the
observed (shaded area) and predicted (line) age or length compositions for every
fleet, both annually and aggregated across years. Joint-sex data are drawn on a
single bin axis with females above and males below zero.

```{r}
plot_comp(model_1)
```

### Survey indices

`plot_index()` shows observed vs. predicted survey biomass; pass `log = TRUE` for the log scale. `plot_indexresidual()` plots the residuals as observed minus predicted, on whichever scale the fleet is fitted on (log for a `Lognormal` fleet, the natural scale for the others), a useful first check for time trends or heteroscedasticity. A positive residual means the survey saw more than the model predicted. These return `ggplot` objects.

```{r}
plot_index(model_1)
plot_index(model_1, log = TRUE)
plot_indexresidual(model_1)
```

`plot_catchability()` draws the other half of the same fit: the catchability the model scaled each survey by, faceted by fleet. It reads the realized `q`, so one figure covers an estimated or fixed mean, annual deviations under `Time_varying_q`, an environmental regression, a `q` linkage such as `rw(1 | Year)`, and the closed-form `"Analytical"` value. A fleet with a time-invariant `q` draws a flat line. Hindcast years only, the model does not project catchability.

```{r}
plot_catchability(model_1)
plot_catchability(model_1, log = TRUE)
```

### Catch

```{r}
plot_catch(model_1)
```

## One-step-ahead (OSA) residuals

Pearson residuals on composition data are *not* standard normal even when the
model is correct: the age/length bins are correlated (more fish in one bin means
fewer in another), and random effects induce correlation across years. They are
at least standardized by the variance the fleet's own likelihood assumes: the
composition weight enters as an effective sample size under a multinomial, and a
Dirichlet-multinomial's overdispersion is in the denominator, so a weighted or
DM fleet is not inflated on that account alone. One-step-ahead
(OSA) residuals remove both problems. Each observation is residualized conditional
on the observations added before it, integrating out the random effects, so under a
correctly specified model the residuals are iid standard normal (Thygesen et al. 2017;
Trijoulet et al. 2023). This makes them suitable for objective goodness-of-fit testing.

`osa_residuals()` computes them *post hoc* for any combination of the fitted data
types, survey indices (`"index"`), fishery catch (`"catch"`), age/length
composition (`"comp"`), conditional age-at-length (`"caal"`), predator diet
(`"diet"`), and, where a linkage holds an observed state-space covariate, the
covariate observation itself (`"ecov"`). Composition data use the conditional
binomial / beta-binomial decomposition. Because OSA re-optimizes the random
effects for each observation it is expensive, so run it on a converged fit
(`estimateMode < 3`). All sources are available from any fit; `osa_residuals()`
builds the required observation data on demand and skips a requested source the
model has no observations for (it errors only if *none* of them do). `"diet"` is
the one source not in the default set, it applies only to multispecies models
and can be slow, so pass `source = "all"` to include it.

```{r osa}
# Aggregate (catch + index) residuals
osa <- osa_residuals(model_1, source = c("index", "catch"))
head(osa)

# Statistical diagnostics (Stewart & Monnahan 2025): SDNR and lower/upper tail
# statistics, each with the interval expected under the standard-normal null.
osa_diagnostics(osa)

# Q-Q plot (with SDNR / tail annotation) + residual-by-year
plot(osa)

# residuals() also exposes them in the common residual schema (`source` selects
# the data, like `type` in stats::residuals.glm()):
residuals(model_1, type = "osa", source = c("index", "catch"))

# Composition OSA residuals, no special fit required
plot_comp(model_1, residual_type = "osa")              # Q-Q + signed bubbles
residuals(model_1, type = "osa", source = "comp")
```

A triangle marks an outlier. On the OSA panel the cut is `qnorm(1 - 0.05 / (2 n))`
for the panel's `n` residuals, so under the model a panel shows 0.05 triangles on
average whatever its size. The Pearson panel keeps `|resid| > 3`: those residuals
are sum-constrained and overdispersed, not N(0, 1).

### Choosing the one-step-ahead method

`method` selects how the conditional distribution is read. The default,
`"oneStepGaussianOffMode"` (WHAM's and SAM's), approximates it as Gaussian: the
observation is treated as a free variable, the mode of the conditional density
is found in that coordinate, and the observation is standardized against it.

`method = "cdf"` instead asks the model for the conditional CDF and returns
`qnorm(F(x))`. For a composition bin the conditional is a binomial, written
through the regularized incomplete beta so it is defined at the fractional
counts composition data hold; for the aggregate series it is `pnorm`. On a
fixed-effect model nothing is approximated; with random effects see below.

The two do not agree on composition data, and the disagreement is measurable.
Simulating observations from a fitted model and residualizing them at the
parameters that generated them makes the right answer exactly standard normal,
so the methods can be scored (`tools/verify/verify-osa-cdf.R`, BS2017SS, 20
replicates of 4538 composition bins):

| method | mean | sd | lag-1 acf within a composition | KS rejects |
|---|---|---|---|---|
| `oneStepGaussianOffMode` | +0.103 | 0.918 | +0.060 | every replicate |
| `cdf`, `discrete = FALSE` | +0.610 | 1.262 | +0.434 | every replicate |
| `cdf`, `discrete = TRUE` | −0.007 | 0.995 | +0.001 | none |

The null standard error on the autocorrelation is 0.015, so only the last row is
consistent with the residuals being iid standard normal. A composition bin holds
a count, so its conditional CDF is a step function and `qnorm(F(x))` inherits
the step; the randomized quantile residual `qnorm(F(x) − U f(x))` (Dunn and
Smyth 1996, as Trijoulet et al. 2023 prescribe) removes it. That is why
`discrete` defaults to `TRUE` under `"cdf"`, and why `"cdf"` with
`discrete = FALSE` is the worst of the three, not the best. On the aggregate
index and catch series, which really are Gaussian, all three agree to 5e-5
wherever the 8.04 ceiling below does not bind.

The step it randomizes over is the **model's**, not the data's: the counts the
likelihood reads are `proportion × Sample_size`, so on a fleet whose sample size
has been tuned they are not whole fish. That is the right choice: an OSA
residual asks whether the observation is consistent with the model as specified,
but it is worth knowing that a heavily down-weighted fleet is being
residualized against a multinomial it was never sampled from.

```{r osa-cdf}
osa_cdf <- osa_residuals(model_1, source = c("index", "comp"), method = "cdf")
osa_diagnostics(osa_cdf)
```

Three things to know before switching:

* **`predicted` is `NA` on every row.** The method never forms a conditional
  mean, which is the point, under the Gaussian methods that mean is an expected
  bin count that can come back negative on a sparse bin, and the residual on
  those rows is then biased positive. Fitted values are in `model_1$quantities`
  and in `residuals(model_1, type = "pearson")`. (`sd` is `NA` under the default
  method as well; only `method = "oneStepGaussian"` returns one.)
* **A Dirichlet-multinomial fleet cannot use it.** Its conditional is a
  beta-binomial, which has no closed-form CDF and cannot be summed at a
  fractional count. Those fleets are residualized with
  `"oneStepGaussianOffMode"`, announced in a message, and recorded in the
  object's `method` attribute. Their `predicted` is blanked too, so the column
  keeps one meaning across the object.
* **`|residual|` is censored at 8.04, in both directions.** The upper end is
  forced by double precision, `F` is recovered as `1/(1 + exp(.))`, which
  saturates at the last double below one. The lower end is not: that same
  expression takes a small `F` down to about 1e-308. It is censored to match
  anyway, because an asymmetric ceiling would show as a long left tail against a
  wall on the right, which is what skewness in the residuals looks like. The
  cost is real, `osa_diagnostics()` computes SDNR and the tail statistics on
  the censored values, so it bites hardest on a short series where one
  observation holds the statistic. `osa_residuals()` warns when any residual
  sits there. Which method to reach for then is measured, not obvious
  (`tools/verify/verify-osa-cdf-accuracy.R`, a 12-year survey with one
  observation multiplied by 200):

  | method | min | max | non-finite | SDNR |
  |---|---|---|---|---|
  | `oneStepGaussianOffMode` | −11.77 | 0.19 | 1 | `NA` |
  | `oneStepGaussian` | −11.77 | 38.98 | 0 | 12.89 |
  | `oneStepGeneric` | −3.95 | 3.33 | 0 | 2.21 |
  | `cdf` | −8.04 | 8.04 | 0 | 4.59 |

  Only `"oneStepGaussian"` reports the uncensored value here: the package
  default returns `NaN` on that row, so its SDNR is unusable rather than merely
  large, and `"oneStepGeneric"` compresses the outlier harder than the ceiling
  does. The default's failure is magnitude-dependent (at a ×20 outlier it is
  finite and matches `"oneStepGaussian"`) so reach for that method
  specifically, and on the fleet in question: it costs an `nlminb` per
  observation.

### With random effects, the choice reverses for the aggregate series

`"cdf"` is exact only on a fixed-effect model. With random effects
`oneStepPredict()` integrates the CDF over the latent states by Laplace, and
that integrand is a Gaussian times a sigmoid rather than a density, where, for
a Gaussian observation, the Gaussian methods integrate a density and are exact.
Measured against the exact Kalman innovations of a linear-Gaussian state space
model, as the latent state becomes more informative relative to the observation
(reproduced by `tools/verify/verify-osa-cdf-accuracy.R`):

| latent/observation sd | `fullGaussian` | `oneStepGaussian` | `cdf` |
|---|---|---|---|
| 0.56 | 4e-16 | 3e-14 | 7e-04 |
| 1.12 | 1e-15 | 4e-14 | 6e-03 |
| 2.24 | 1e-15 | 3e-14 | 4e-02 |

**That is a result about a linear-Gaussian model, and it does not transfer
wholesale.** The Gaussian methods are exact when the one-step-ahead *predictive*
is Gaussian, which needs the model to be linear in the random effects, and
Rceattle's index and catch are `exp()` of cumulated log recruitment deviations
pushed through the population dynamics. `fullGaussian` and `oneStepGaussian`
cannot differ for a Gaussian conditional, so their disagreement measures the
departure: on a 17-deviation fixture they differ by 0.091 on both index and
catch, while `"cdf"` differs from `oneStepGaussian` by 0.017 on index. **No
method is exact for index or catch under random effects.**

**`"ecov"` is the exception, and there a Gaussian method is right.** Its
conditional (a Gaussian measurement of an AR1 latent, every other data term
unconditional) genuinely is linear-Gaussian: the two Gaussian methods agree to
4e-14 on the QAR1 fixture, while `"cdf"` sits 0.139 away, about a quarter of the
residual sd. It does not reverse for compositions: their conditional is discrete
and skewed, which is what the Gaussian methods get wrong, and by much more.
Simulating from a 22-random-effect model with the recruitment deviations redrawn
and residualizing at the generating parameters (1680 residuals),
`oneStepGaussianOffMode` gives mean +0.513, sd 0.404 and KS rejection in every
one of 120 replicates, against +0.006, sd 1.002 and 6 of 120, the nominal 5%,
for `"cdf"` with `discrete = TRUE`.

What limits `"cdf"` on compositions is the depth of the conditioning rather than
the random effects themselves. On `BS2017SS` with `random_rec = TRUE` (159 random
effects, 4538 composition bins) it returns 1879 non-finite residuals, against 0
for `"oneStepGaussianOffMode"` on the same fit, and takes hours; the same rows
residualized on their own come back clean, and the 22-random-effect model above
is residualized correctly. So try it, read the warning it issues, and where the
residuals come back non-finite in bulk fall back to a Gaussian method for that
source, remembering that its composition residuals are the under-dispersed ones
in the comparison above.

Both halves of that split are what the SAM authors do, in two packages.
`stockassessment::residuals.sam()` names no method, so it takes
`oneStepPredict()`'s `"oneStepGaussianOffMode"` default with `discrete = FALSE`,
on a model that is random effects throughout, whose observations are Gaussian
on the log scale. Their composition package `compResidual::resMulti()`, the
implementation Trijoulet et al. (2023) cite, instead hardcodes
`method = "cdf", discrete = TRUE`. Rceattle fits both kinds of data, so it makes
the same choice per observation type inside one call rather than across two
packages.

One caveat that applies to every method: `Estimate_index_sd = "Analytical"` and
the analytical `Catchability` forms concentrate their parameter out using **all**
the observations, including the one being residualized, so the conditioning is
not strictly one-step-ahead and the residuals are approximate.

### Fleets with tail accumulation

Where a fleet folds its composition tails into a boundary bin
(`Comp_accum_young` / `Comp_accum_old`), both the OSA and the Pearson residuals
describe the bins the likelihood actually fit, not the raw bins in `comp_data`.
A fleet on 10 ages with `Comp_accum_young = 3` reports bins 3 to 10: bin 3 holds
ages 1 to 3 combined, and the `accumulated` column (`Accumulated` from
`residuals()`) marks it, since it stands for a range rather than the single age
it is named for. Expect fewer residual bins than raw age bins for such a fleet.

The two differ in one place. The one-step-ahead decomposition drops each
composition's last bin, which is fixed by the sum-to-sample-size constraint.
Under an *old*-tail accumulation that dropped bin is the upper accumulated one,
so the fleet's upper boundary bin appears in the Pearson residuals and not in
the OSA residuals.

Following Stewart and Monnahan (2025), a practical workflow is:

1. **Inspect the aggregate fit** across years (`plot_comp(model_1)`): systematic
   lack of fit here points to structural misspecification (e.g. the wrong
   selectivity shape). Counts are pooled across the fitted hindcast years, and
   the interval holds 95% of the data the model predicts, not a confidence
   interval on the mean, so observations outside it are the ones to explain.
   The panel is annotated with the input sample size, the effective sample size
   the likelihood assumed (`ESS (likelihood)`), and the one this fit's residuals
   imply (`ESS (McAllister-Ianelli)`). The gap between those two is the
   reweighting decision: they agree when the fleet is already weighted the way
   its residuals suggest.
2. **Compare the SDNR** to its null interval (from `osa_diagnostics()`): a value
   outside the interval flags over- or under-dispersion, often a data-weighting
   or effective-sample-size issue.
3. **Compare the lower/upper tail statistics** to their null intervals: these
   localize departures to the tails of the residual distribution. They are
   reported as order statistics, `lower_r` / `upper_r` name which one, and
   `lower_p` / `upper_p` the nominal probability it sits at, which is near but
   not exactly 2.5%, against the exact `Beta(r, n - r + 1)` null. The
   statistic and its interval must come from the same estimator, so
   `osa_diagnostics()` no longer simulates and its `nsim` / `seed` arguments are
   ignored. The residuals themselves remain randomized-quantile residuals: pass
   a seed to `osa_residuals()` to control those.
4. **Inspect Pearson bubble plots** (`plot_comp(model_1)`, the default): patterns
   in the sign of residuals across age/year can identify *where* lack of fit
   occurs, Stewart and Monnahan recommend keeping these alongside OSA
   residuals.
5. **Revise the model or data weighting** where misspecification is found, using
   this together with the other diagnostics rather than as an accept/reject test.

When the diagnosis is a weighting problem rather than a structural one,
`reweight_comps()` runs the McAllister-Ianelli (1997) tuning loop: it refits the
model the way it was fitted, updating each fleet's `Comp_weights` until the
implied weights settle.

```{r}
tuned <- reweight_comps(model_1, n_iter = 10, tol = 0.01)
tuned$reweight$history    # weight per fleet, per iteration
```

Restrict it to particular fleets with `fleets =`, taking either `Fleet_code`
values or names matching `fleet_control$Fleet_name`, give ids or names, not a
mix, since R coerces `c(1, "Survey")` to character.

```{r}
tuned <- reweight_comps(model_1, fleets = c("Survey_1", "Fishery_1"))
```

Only multinomial fleets are tuned. A Dirichlet-multinomial fleet estimates its
own weight inside the likelihood, so tuning it externally would compete with
that estimate; those fleets are named and skipped.

The alternatives are Francis (2011) reweighting, applied by hand, and estimating
the weights directly with a Dirichlet-multinomial likelihood
(`Comp_distribution = "DirichletMultinomial"`), which avoids the iteration
entirely, note that under that likelihood `Comp_weights` is read as the **log**
of the starting weight, so the model default of `1` is a starting weight of
*e*, and a weight of 1 is written as `0`.

Because OSA residuals for discrete composition data are randomized quantile
residuals, they are stochastic; `osa_residuals()` takes a `seed` for
reproducibility.

### Process residuals

`process_residuals()` is the complementary check on the *process* model. It
standardizes the model's random-effect deviations (recruitment, initial
abundance, catchability) against their process prior, drawing once from the
joint posterior (SAM's `procres`; Nielsen and Berg 2014). Under a correct
process model these too are approximately iid standard normal.

```{r procres}
pr <- process_residuals(model_1, process = "recruitment")
osa_diagnostics(pr)
plot(pr)

# residuals() exposes them too:
residuals(model_1, type = "process")
```

## Retrospective analysis

A retrospective analysis removes the most recent years of data one peel at a time and re-fits the model. Mohn's rho summarizes how the terminal estimates shift across peels; for SSB, a rho outside ±0.2 generally signals a problem worth investigating.

```{r}
model_1_retro <- retrospective(model_1, peels = 5)

# Mohn's rho for each quantity
model_1_retro$mohns

# Plot historical trajectories across peels
plot_biomass(model_1_retro$Rceattle_list)

# Include the projection period to see how the forecast changes
plot_biomass(model_1_retro$Rceattle_list, incl_proj = TRUE)
```

Each peel holds its own terminal year as `data_list$endyr`, so the first plot draws every peel only as far as it was fit and the peels fan out.

A peel still estimates the years it dropped: they are its retrospective forecast, fit to the observed catch with recruitment projected at the single deviation `sample_rec(sample_rec = FALSE)` uses, computed over the peel, and the survey and composition data withheld. Fishing mortality is estimated through the unpeeled terminal year and the harvest control rule takes over after it, so each peel has three years:

| Field | Meaning |
|---|---|
| `endyr`, `endyr_peel` | the peel's terminal year, what it was fit through |
| `endyr_full` | the unpeeled terminal year, where the retrospective forecast ends |
| `projyr` | the end of the projection |

`incl_proj = TRUE` plots the forecast and projection years. Take the forecast years as `endyr_peel + seq_len(endyr_full - endyr_peel)`, empty for the unpeeled model, where `(endyr_peel + 1):endyr_full` would count down instead.

The `retrospective()` function returns a list with two elements:

| Element | Description |
|---|---|
| `Rceattle_list` | List of fitted models, ordered from most-peeled to full run, each named for its own terminal year (`Year_2013`, …, `Year_2017`) |
| `mohns` | Data frame with columns `Object` (quantity, e.g. `"Biomass"`, `"SSB"`), `Forecast year` (0 = terminal year bias; 1+ = forecast skill), `N` (number of peels), and one column per species with mean relative error (Mohn's rho) |

The `nyrs_forecast` argument (default 3) also evaluates rho for years projected beyond each terminal peel, so a single call quantifies forecast skill alongside retrospective bias.

A peel that did not converge is dropped and a message reports how many, so `Rceattle_list` can be shorter than `peels + 1`. Index it by name rather than position, with a peel missing, `Rceattle_list[[3]]` is not the 3-year peel. If no peel converges, Mohn's rho is `NaN` and the call warns.

## Jitter testing

Jitter testing re-fits the model from many randomly perturbed starting values and checks whether the optimizer keeps returning the same minimum negative log-likelihood (NLL). A spread of NLL values across jitters points to multiple local minima, so the results should be treated with caution.

```{r}
jitters <- jitter(model_1, njitter = 10, phase = TRUE)  # default njitter = 50

# Histogram of NLL differences relative to the best run
hist(log(jitters$nll - min(jitters$nll)),
     main = "Jitter NLL spread (log scale)",
     xlab = "log(NLL - min NLL)")

# Overlay biomass trajectories, tight overlap indicates a stable optimum
plot_biomass(jitters$Rceattle_list)

# Number of runs that converged
length(jitters$Rceattle_list)
```

Non-converging runs are dropped from `Rceattle_list` and reported in a message, so if `length(jitters$Rceattle_list)` is well below `njitter`, convergence is fragile. Read that count as part of the result, not as a nuisance: the dropped starts are the ones that wandered off, so a jitter test that returns only its successes understates the spread.

A run is dropped when its maximum absolute gradient exceeds 1 (the same threshold `convergence_diagnostics()` calls `FAIL`) or when an `sdreport` that was requested did not come back.

`retrospective()` and `jitter()` both fit peels / starts in parallel by default. Pass `cores = 1` to force sequential execution; otherwise the call uses a PSOCK cluster sized at `parallel::detectCores() - 6` (capped at 2 under `R CMD check`).

## Self-test (simulation–estimation)

`self_test()` simulates `nsim` datasets from a fitted model (holding its estimated parameters fixed) and re-fits the model to each. If the refit estimates cluster around the values used to simulate them, the model is at least self-consistent: it can recover its own estimates from data it generated. Persistent bias or a wide spread in a quantity means the data are not informative about it.

```{r}
sims <- self_test(model_1, nsim = 10)            # default nsim = 50

# Number of simulations that converged (non-converged runs are dropped)
length(sims)

# Overlay biomass / SSB trajectories across simulations, the original fit's
# trajectory should sit inside the spread of the refits.
plot_biomass(c(list(model_1), sims), model_names = c("fit", names(sims)))
plot_ssb(c(list(model_1), sims),     model_names = c("fit", names(sims)))
```

Each simulation uses `seed + i` as its RNG seed, so results are reproducible whether or not `cores > 1`. Set `simulate = FALSE` to refit against the model's *expected* values (no observation error) rather than draws from the observation likelihood, useful for confirming that the estimator returns the generating parameters in the noise-free limit. `cores` behaves the same way as in `retrospective()` and `jitter()`.

Every refit starts from the fitted model's *starting* values, not its estimates, so it covers the same ground the original fit did. `phase` therefore defaults to whatever the input model was fit with, pass `phase = TRUE` explicitly if you fit the model some other way and the refits are being dropped, since a model that needed phasing once needs it for every simulated data set too. `start = "estimated"` starts from the fitted estimates instead: much faster, but the optimizer then begins inside the basin containing the generating values and recovery is close to guaranteed by construction, so read it as optimistic about recovery rather than simply less powerful.

Three things to keep in mind when reading the result. Non-converged refits are dropped and reported in a message, and they are dropped *because* they behaved badly (so the returned replicates are a favourable subset, and any spread or bias computed over them (including `compare_sim()`'s CV) understates the truth. By default only the observations are redrawn: every replicate shares the operating model's single realization of recruitment and mortality, so the spread across replicates is observation error alone) a lower bound on estimation uncertainty in SSB and recruitment, not comparable to the model's own uncertainty bands.

Third, watch for a truncation warning on a natural-scale index. A fleet with `Index_distribution = "Normal"`, `"MVN"` or `"MVNORM"` is drawn on the natural scale with an absolute sd, so a draw can come out negative (impossible for an index, and rejected by `data_check()`. Those draws are redrawn, which samples the fleet from a normal *truncated at zero* rather than the untruncated normal the likelihood maximises. Occasional truncation is harmless; a lot of it is not. An acoustic index observed at 0.70 with an absolute sd of 0.80 rejects a fifth of its draws and the truncated mean sits 39% above the untruncated one, so the self-test is measuring recovery against a data-generating process the estimator never assumed. `sim_mod()` warns when any row puts more than 2% of its probability at or below zero) `Phi(-mu/sd)`, read off the fit rather than counted from the draw, so it does not vary between replicates.

The fix for a `Normal` fleet is `Index_distribution = "TruncatedNormal"`, which fits the truncated density, renormalized by `log Φ(µ/σ)` over the `(0, ∞)` the data can actually occupy, and draws from it by inverse CDF. Draw and likelihood then agree exactly and no draw is ever rejected. `"Normal"` is kept untruncated because it reproduces the AMAK `avo_like` / `cpue_like` term for term, which is what an ADMB bridge compares against; use it there and `"TruncatedNormal"` elsewhere. `MVN`/`MVNORM` have no such option: a multivariate normal truncated to the positive orthant has no closed-form sampler, so rejection is the only route and the gap stays. Either way the underlying issue is the same, an absolute sd that large relative to the index means the natural-scale normal was a poor observation model to begin with.

`"TruncatedNormal"` OSA residuals need one thing said about them. The truncation constant is a function of the prediction, not of the observation, so it drops out of any method that reads the curvature of the density in the observation, including `oneStepPredict()`'s default `oneStepGaussianOffMode`, which would return the untruncated `(obs - µ)/σ`. `osa_residuals()` therefore residualizes this family in its own call, integrating the density over `(0, ∞)`, so the residual is `qnorm(F(x))` for the truncated `F`. The correction is the size of the truncated mass: on a fleet predicting 100 with an absolute sd of 150, an observation exactly at the prediction moves from a residual of 0 to −0.44. You do not need to pass anything for this; `method` is overridden for those rows only.

Under `method = "cdf"` none of that applies: the model supplies the truncated `F` in closed form, so the family is residualized in the main call, exactly, and the separate integration (with its cost and its occasional failure to converge on a random-effects model) is not needed.

### Redrawing process error

`process` widens the question. The default (`FALSE`) asks whether the estimator recovers its own parameters from new observations; naming a process redraws that process too, and asks whether it recovers something it was not shown.

```{r}
# Recruitment deviations redrawn as well as the observations
sims <- self_test(model_1, nsim = 10, process = "recruitment")

# "dynamics" covers recruitment, M and growth; "observation" covers
# catchability and selectivity; TRUE covers everything.
sims <- self_test(model_1, nsim = 10, process = "dynamics")
```

Recruitment covers the initial age structure as well as the hindcast deviations, since the initial numbers-at-age are recruitment from before `styr`. Natural mortality covers the `M1_re` random effects, and growth, catchability and selectivity are redrawn wherever they are written as a random linkage (see `vignette("environmental-linkages-and-priors")`). A process the model gives no distribution to (asking for M when `M1_re = 0`, say) is left alone and reported in a warning rather than silently returned unchanged.

Recruitment has one such case worth knowing about. Under `srr_fun = 0` with `srr_pred_fun > 0` (the AMAK/Ianelli configuration, where recruitment is estimated annually and the stock-recruit curve is fitted as a *penalty* on those same deviations) the deviations are scored by two densities at once.

Two terms on one quantity do not compose into a distribution to draw from. Drawing at `R_sd` from the first alone ignores the second, and the second is not a density on the deviations alone either: it couples them to the stock-recruit parameters, which are estimated too. Either way the self-test would measure recovery against a process the model does not assume, so the deviations are not redrawn and `sim_mod()` says why. A random linkage on a recruitment parameter is a separate latent and is still drawn. Set `srr_fun` to the stock-recruit function itself to redraw the deviations.

The fitted deviations are no longer what generated the data, so the values that *did* come back attached to each replicate, each with a mask of the cells the draw touched:

```{r}
truth <- attr(sims, "process_sim")[["Sim_1"]]
rec   <- truth$rec_dev[truth$rec_dev_drawn]
```

Compare estimates against those, not against the operating model. The mask matters: the process draws cover the hindcast only, and `beta_linkage_re` spans every random linkage whether or not its process was asked for, so a correlation taken over the whole array scores fitted values as though they had been simulated. For the same reason `compare_sim()`, which compares against the operating model, is not valid for `process`-drawn replicates.

One caveat before reading a redrawn M as validation. On `BS2017SS` with a year-varying M random effect, refits recover the simulated deviations (correlation around 0.5, positive in every replicate) but recover their standard deviation poorly, roughly 70% low, and collapsing to zero in a third of replicates. That is the usual behaviour of a variance component estimated from little information, not a fault in the simulation, and a year-varying M is weakly informed because it trades off against selectivity and catchability. Read a self-test that redraws M as a test of deviation recovery rather than of variance recovery.

When the returned list is shorter than `nsim`, `debug = TRUE` returns *every* simulation so the failures can be read directly rather than inferred:

```{r}
sims <- self_test(model_1, nsim = 10, debug = TRUE)

failed <- names(which(!attr(sims, "converged")))
sims[[failed[1]]]$convergence          # which check failed, and on which parameter
```

## Likelihood profile

`profile()` re-fits the model across a grid of fixed values for a chosen parameter and returns the NLL surface. A flat profile means the data say little about the parameter; a minimum away from the MLE means the fit has not truly settled there. It supports the recruitment standard deviation, the stock–recruit parameters, and natural mortality.

Specify the parameter with a natural-scale alias: `"sigmaR"` (recruitment
standard deviation), `"M1"` (natural mortality), or the stock–recruit parameters
`"R0"`, `"alpha"`, and `"beta"`. Aliases take natural-scale values directly.
`slots` gives the cell(s) to profile (usually just the species index) and
`values` gives the grid of values for each. Supplying more than one slot
profiles over the full grid of combinations.

```{r}
# 1-D profile: sigmaR for species 1 (natural-scale alias)
prof_sigmaR <- profile(
  fitted = model_1,
  param    = "sigmaR",
  slots    = list(1),
  values   = list(seq(0.1, 1.5, by = 0.1))
)

plot(prof_sigmaR)   # components, with the total overlaid; see below

# 1-D profile: SRR alpha for species 1
# (alias fills in the rec_pars column; slot is just the species index)
prof_alpha <- profile(
  fitted = model_1,
  param    = "alpha",
  slots    = list(1),
  values   = list(seq(2, 80, length.out = 20))
)

# 2-D cross-profile: M1 across sex for species 1, age 1
prof_M_sex <- profile(
  fitted = model_1,
  param    = "M1",
  slots    = list(c(1, 1, 1), c(1, 2, 1)),  # males and females
  values   = list(seq(0.10, 0.40, by = 0.05),
                  seq(0.10, 0.40, by = 0.05))
)
```

### Moving a set of cells together

`slots` cross, so *k* slots give a *k*-dimensional profile. That is wrong for a
parameter supplied as a vector, an empirical age-based M profiled over 13
values across ten ages would be `13^10` fits. `joint` puts every named cell on
**one** grid instead: `"multiply"` scales each cell's current value, preserving
the shape of the schedule (`1` is the fitted model), and `"add"` shifts it,
preserving the increments (`0` is the fitted model).

```{r}
sp <- 1
prof_M_all <- profile(
  fitted = model_1,
  param  = "M1",
  slots  = lapply(seq_len(model_1$data_list$nages[sp]),
                  function(a) c(sp, 1, a)),   # every age bin of species 1
  values = list(seq(0.7, 1.3, by = 0.05)),
  joint  = "multiply"
)

plot_profile(prof_M_all)
```

### Catchability

`param = "q"` profiles a fleet's base catchability, and `slots` takes a fleet
name. Fleets sharing a `Catchability_index` share one q, so naming any member
profiles the whole group; the others are added with a message.

```{r}
prof_q <- profile(
  fitted = model_1,
  param  = "q",
  slots  = list("BT_Pollock"),
  values = list(seq(0.5, 2.0, by = 0.1))
)
```

`profile()` returns `Rceattle_list` (one fit per grid row, `NULL` where the fit failed), `grid` (the user-scale value grid, a multiplier or offset under `joint`), `nll` (joint NLL aligned with `grid`, `NA` for non-converged fits), `alias` (the natural-scale name you profiled under, if you used one), `joint`, and echoes of `param` and `slots`. To cross-profile across multiple species, supply one slot per species (e.g. `slots = list(1, 2, 3)` with `param = "sigmaR"`); to cross-profile M1 across sex, supply one slot per sex as in the third example above. `cores` behaves the same way as in the other diagnostic functions.

### Where the data disagree

The total says how well the data determine the parameter, not whether the data
sources agree about it, a smooth quadratic total can be two surveys pulling in
opposite directions. `plot_profile()` draws the components separately, in the
layout `r4ss::SSplotProfile()` uses.

```{r}
prof_M <- profile(
  fitted = model_1,
  param  = "M1",
  slots  = list(c(1, 1, 1)),
  values = list(seq(0.20, 0.45, by = 0.025))
)

plot_profile(prof_M, xlab = "M at age 1")
```

Read it by **where each curve bottoms out**, not by how deep it is: every series
is re-zeroed at its own minimum and a point marks it, so the spread of those
points along the x axis is the disagreement. The heavy black line is the total.

When one component dwarfs the rest the others flatten onto the axis and where
they prefer the parameter cannot be read. `relative = "scaled"` puts every curve
on 0 to 1 so the minima can be compared directly; it discards magnitude, so
raise `minfraction` alongside it.

`profile_components()` returns the same numbers, one row per grid point per
component, each labelled with the fleet or species it belongs to.

```{r}
comps <- profile_components(prof_M)
head(comps)
```

Under `random_rec = TRUE` the total is the Laplace-approximated marginal
likelihood while the components are the inner joint negative log-likelihood, so
they will not sum; compare the shapes, not the sums. See `?plot_profile` for
`relative`, `minfraction`, model overlays, and what `weighted = FALSE` does and
does not include.

## Comparing runs

Overlaying sensitivity runs is itself a useful diagnostic, large divergences in biomass or mortality deserve scrutiny.

```{r}
mod_list  <- list(model_1, model_1_retro$Rceattle_list$Year_2019)
mod_names <- 1:length(mod_list)

plot_biomass(Rceattle = mod_list, model_names = mod_names)
plot_recruitment(Rceattle = mod_list, model_names = mod_names, add_ci = TRUE)
plot_depletionSSB(Rceattle = mod_list, model_names = mod_names)
```


## Collecting it into report tables

Everything above returns its own shape. `report_tables()` gathers them into one
set of tidy tables, the fit, the likelihood decomposition, the time series with
uncertainty, the reference points, the fits to index and catch, and whichever
diagnostics you pass it:

```{r}
tabs <- report_tables(model_1,
                      retro = model_1_retro, jitter = jitters, osa = osa)
tabs
tabs$retrospective          # Mohn's rho by species and forecast year
tabs$osa                    # SDNR and tail statistics per fleet
```

It never refits, so a diagnostic you have not run is simply an absent section.
Check the `basis` column of `tabs$reference_points` before quoting a reference
point: quantities the fit never estimated are returned as `NA` with the reason,
rather than as the value left in the array. `standard_output()` relabels the
result for the `stockplotr` and `asar` reporting packages.

## Model average

To average estimates across model variants, see `?model_average`.

```{r}
mod_avg <- model_average(list(model_1, model_1), weights = c(1,1))
plot_biomass(mod_avg)
```
